---
title: "SEM diagnostic — sensitivity analysis"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: true
---

# Overview

## Role in the pipeline

This script is the **second step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — assembles the full SEM database

Its outputs are consumed directly by:

3.  `check_models_performance.qmd` — full rolling-window SEM execution
4.  `results_visualisation.qmd` — figure production and ecological interpretation

## What this script produces

| Output | File | Description |
|------------------------|------------------------|------------------------|
| Full benchmark table | `benchmark_results_database.csv` | Diagnostic statistics for every parameter combination x response variable |
| Selected model parameters | `model_parameters.txt` | Optimal window duration, completeness threshold, and transformation, saved as JSON |

## Statistical approach

The causal model has four structural equations organised in three tiers, with
`land_use` (0 = unmanaged, 1 = managed) as the exogenous binary driver.
At the diagnostic stage, each equation is fitted **independently as a classical
linear mixed model** (LME) rather than within the full piecewise SEM structure.
This isolates the residual behaviour of each response variable and avoids
confounding model-fitting issues across equations. Interaction terms are excluded
at this stage; the full SEM is additive throughout — no interaction terms are used in any script.
the optimal window and transformation have been determined.

**Tier 1 — land use shapes microclimate** (additive):

$$\text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$
$$\text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$

**Tier 2 — microclimate drives root:**

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu}$$

**Tier 3 — microclimate and root drive fauna:**

$$\text{Fauna} \leftarrow \beta_6 \cdot \text{mc}_1 + \beta_7 \cdot \text{mc}_2 + \beta_8 \cdot \text{Root} + \beta_9 \cdot \text{lu}$$

*where mc = microclimate index, lu = land_use.*

Random structures differ by equation type:

- **Microclimate equations:** `~ 1 | depth` — microclimate sensors have no
  orientation variable, so depth is the only available grouping factor.
- **Root and fauna equations:** `~ 1 | orientation/depth` — both orientation
  and depth are recorded for biological variables, and depth is nested within
  orientation. Falls back to `~ 1 | depth` on convergence failure.
  
All equations include an **AR(1) correlation structure** to account for temporal
autocorrelation within grouping levels.

## Sensitivity analysis rationale

Three structural choices must be validated before running the full SEM:

| Parameter | Question | Values tested |
|------------------------|------------------------|------------------------|
| **Window duration** | How many days are needed for a stable fit? | 5, 7, 10, 14, 21 days |
| **Min. data completeness** | How many missing images can be tolerated? | 10% to 50% |
| **Transformation** | Which transform best normalises the variables? | none, sqrt, Freeman-Tukey, log, Hellinger |

These parameters are crossed in a **full factorial grid** (5 × 4 × 5 = 100
combinations). The same transformation type is applied uniformly to all modelled
variables within each configuration, but **z-score standardisation is performed
independently within each rolling window** after the transformation. This ensures
that standardised coefficients reflect within-window variation rather than the
global distribution of the full time series — which would mix seasonal and
inter-annual variance in an uncontrolled way.

**Completeness criterion:** `min_rate` is the fraction of rows surviving
`drop_na()` relative to the total rows in the window. This definition is robust
to unequal sampling densities across depths and variables (e.g. root growth
measured at only one depth while microclimate is recorded at both).

| Diagnostic | Metric | Threshold |
|------------------------|------------------------|------------------------|
| Heteroscedasticity | Spearman rho between abs(residuals) and fitted values | \< 0.4 |
| Skewness | Absolute skewness of normalised residuals | \< 1.5 |
| Collinearity | Maximum VIF across predictors | \< 5 |
| Explanatory power | Marginal and conditional R-squared | — |

> **Note on VIF.** With no interaction terms in the diagnostic models, VIF is
> computed directly on the fitted model formula. `land_use` is excluded from VIF
> calculation as a binary exogenous driver; the check targets collinearity among
> continuous predictors only.

## Selection criterion

The optimal configuration is chosen by a **two-step procedure**:

1.  For each response variable **independently**, select the transformation
    maximising `prop_valid_models × avg_R2_marginal`.
2.  Across biological responses **jointly** (root and fauna), find the
    (window, min_rate) pair maximising `combined_validity × mean_R2`, where
    `combined_validity = min(prop_valid_root, prop_valid_fauna)`. Using the
    minimum ensures that neither response can mask a poor fit in the other.

------------------------------------------------------------------------

## 1. Setup

```{r setup}
#| message: false

invisible(Sys.setlocale("LC_TIME", "C"))

libs <- c(
  "tidyverse", "lubridate",
  "lme4", "purrr", "nlme", "moments", "car",
  "knitr", "kableExtra", "patchwork", "scales"
)
invisible(lapply(libs, require, character.only = TRUE))

# ── Diagnostic thresholds ──────────────────────────────────────────────────────
HETERO_THRESHOLD   <- 0.4
SKEWNESS_THRESHOLD <- 1.5
VIF_THRESHOLD      <- 5.0

# ── Data loading ───────────────────────────────────────────────────────────────
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep        = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    fauna    = all_taxa,
    root     = root_growth_cm2_count_24h
  )

fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")
```

------------------------------------------------------------------------

## 2. Functions

### 2.1 Rolling window splitter

```{r fn-rolling-windows}
#' Split a time series into overlapping rolling windows
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Duration of each window in days.
#' @param step_days   Integer. Step between consecutive window start dates.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )
  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 2.2 Per-observation diagnostic extractor

```{r fn-extract-diag}
#' Extract per-observation diagnostic quantities from a fitted lme model.
#'
#' Returns a tidy dataframe with one row per observation, carrying the
#' quantities needed for check_model-style figures (PPC, linearity,
#' homoscedasticity, influential observations, normality).
#'
#' Cook's distance is approximated from normalised residuals and leverage
#' because nlme does not expose cooks.distance() for lme objects.
#'
#' @param m         A fitted nlme::lme object.
#' @param w_name    Character. Window start date string (used as date_start).
#' @param resp_name Character. Response variable name (stored in `response`).
#' @return Tibble with columns: date_start, response, observed, fitted_pop,
#'         fitted_subj, residual, leverage, cooks_d.
extract_lme_diag <- function(m, w_name, resp_name) {
  res_norm <- residuals(m, type = "normalized")
  n_m      <- length(res_norm)
  n_fix    <- length(nlme::fixef(m))

  safe <- function(expr) {
    tryCatch(as.numeric(expr), error = function(e) rep(NA_real_, n_m))
  }

  lev    <- safe(hatvalues(m))
  cook_d <- (res_norm^2 * lev) / (n_fix * pmax((1 - lev)^2, 1e-6))

  tibble(
    date_start  = as.Date(w_name),
    response    = resp_name,
    observed    = safe(nlme::getResponse(m)),
    fitted_pop  = safe(fitted(m, level = 0)),
    fitted_subj = safe(fitted(m, level = length(m$groups))),
    residual    = res_norm,
    leverage    = lev,
    cooks_d     = cook_d
  )
}
```

### 2.3 Independent model fitting

Each structural equation is fitted as a standalone LME. Random structures differ
by equation type because microclimate sensors have no `orientation` variable:

- **Microclimate:** `~ 1 | depth` — the only available grouping factor.
- **Root / fauna:** `~ 1 | orientation/depth` (depth nested within orientation),
  with automatic fallback to `~ 1 | depth` on convergence failure.

No interaction terms are included — the diagnostic targets the additive structure only.

```{r fn-fit-models}
#' Fit the four SEM equations independently as classical LME models
#'
#' Random grouping structures:
#'   - microclimate_1 / microclimate_2: `~ 1 | depth`
#'     (microclimate sensors have no orientation variable)
#'   - root / fauna: `~ 1 | orientation/depth`, falling back to `~ 1 | depth`
#'     on convergence failure
#'
#' @param data Dataframe with columns: `land_use`, `root`, `fauna`,
#'             `microclimate_1`, `microclimate_2`, `orientation`, `depth`.
#'             Variables must already be transformed and z-scored within window.
#' @return Named list of four `lme` objects (NULL entries on convergence failure).
fit_models_independently <- function(data) {

  lme_ctrl <- lmeControl(msMaxIter = 200, opt = "optim")

  attempt <- function(formula, rand_form) {
    tryCatch(
      withCallingHandlers(
        lme(formula,
            data        = data,
            random      = rand_form,
            correlation = corAR1(form = rand_form),
            control     = lme_ctrl),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
  }

  # FIX 1: Microclimate sensors have no orientation variable.
  # Using ~ 1 | orientation/depth would either fail or duplicate values
  # artificially, inflating the apparent grouping structure.
  # Depth is the only valid grouping factor for these equations.
  mc1 <- attempt(microclimate_1 ~ land_use, ~ 1 | depth)
  mc2 <- attempt(microclimate_2 ~ land_use, ~ 1 | depth)

  # Root and fauna: both orientation and depth are recorded.
  # Depth is nested within orientation. Falls back to depth-only if nesting
  # causes convergence failure (e.g. too few levels per group in short windows).
  fit_bio <- function(formula) {
    m <- attempt(formula, ~ 1 | orientation / depth)
    if (is.null(m)) m <- attempt(formula, ~ 1 | depth)
    m
  }

  list(
    microclimate_1 = mc1,
    microclimate_2 = mc2,
    root           = fit_bio(root  ~ microclimate_1 + microclimate_2 + land_use),
    fauna          = fit_bio(fauna ~ microclimate_1 + microclimate_2 + root + land_use)
  )
}
```

### 2.4 Transformation functions

```{r fn-transforms}
#' Shift a vector to be strictly positive before a transform that requires x > 0.
#' If min(x) <= 0, shifts the whole vector by |min(x)| + 0.001 and emits a message.
shift_positive <- function(x, label = "") {
  x_min <- min(x, na.rm = TRUE)
  if (!is.na(x_min) && x_min <= 0) {
    shift <- abs(x_min) + 0.001
    message(sprintf("  [shift] '%s' shifted by +%.4f (min was %.4f)", label, shift, x_min))
    x <- x + shift
  }
  x
}

#' Freeman-Tukey: sqrt(x) + sqrt(x + 1). Stabilises variance in count data with zeros.
transform_ft <- function(x) sqrt(x) + sqrt(x + 1)

#' Log: log(x + 1) after shifting to positive if needed.
transform_log <- function(x, label = "") log1p(shift_positive(x, label))

#' Sqrt after shifting to positive if needed.
transform_sqrt <- function(x, label = "") sqrt(shift_positive(x, label))

#' Hellinger: sqrt(x / sum(x)). Window-relative — must be applied per window.
transform_hellinger <- function(x) sqrt(x / sum(x, na.rm = TRUE))

#' Compute marginal and conditional R² for an nlme::lme model.
#'
#' Uses the fitted-value variance approach: R²m = var(fixed-effect predictions)
#' / var(observed); R²c = var(full predictions including random) / var(observed).
#' This avoids parsing the VarCorr character matrix, which is fragile across
#' nlme versions. fitted(m, level = 0) returns population-level predictions
#' (fixed effects only); fitted(m, level = k) adds all k levels of random effects.
#'
#' @param m A fitted nlme::lme object.
#' @return Named numeric vector c(R2m = ..., R2c = ...).
r2_lme <- function(m) {
  tryCatch({
    y_obs <- as.numeric(nlme::getResponse(m))
    y_fix <- as.numeric(fitted(m, level = 0))                # fixed effects only
    y_all <- as.numeric(fitted(m, level = length(m$groups))) # fixed + all random

    ss_obs <- var(y_obs, na.rm = TRUE)
    if (is.na(ss_obs) || ss_obs == 0)
      return(c(R2m = NA_real_, R2c = NA_real_))

    r2m <- var(y_fix, na.rm = TRUE) / ss_obs
    r2c <- var(y_all, na.rm = TRUE) / ss_obs

    # Clamp to [0, 1]: numerical noise can produce tiny negative or >1 values
    c(R2m = min(max(r2m, 0), 1),
      R2c = min(max(r2c, 0), 1))
  }, error = function(e) c(R2m = NA_real_, R2c = NA_real_))
}
```

### 2.5 Main diagnostic pipeline

```{r fn-pipeline}
#' Run the rolling-window diagnostic pipeline for one taxon variable
#'
#' Each window is processed independently:
#'   1. Subset and drop NAs
#'   2. Apply transformation
#'   3. Z-score standardise within the window (excludes land_use)
#'   4. Fit four LME models independently
#'   5. Extract residual diagnostics and R² per model
#'
#' Completeness is defined as complete rows / total rows in the raw window.
#' This is robust to unequal sampling densities across depths and variables
#' (e.g. root growth measured at only one depth while microclimate is at both).
#'
#' @param taxon_var   Character. Column name of the fauna variable.
#' @param data_raw    Dataframe. Full SEM database (raw, untransformed).
#' @param m_indices   Character vector of length 2. Microclimate index names.
#' @param transfo_type Character. One of "none", "sqrt", "sqrt_plus_one",
#'                    "log", "hellinger".
#' @param window_days Integer. Rolling window duration (days).
#' @param step_days   Integer. Step between window start dates.
#' @param min_rate    Numeric [0, 1]. Minimum completeness ratio
#'                    (complete rows / total rows in window).
#' @param n_windows   Integer or NULL. Random subsample size for benchmarking.
#' @return Dataframe with one row per model x window, or NULL.
run_sem_pipeline <- function(taxon_var, data_raw, m_indices, transfo_type,
                             window_days = 14, step_days = 1,
                             min_rate = 0.3, n_windows = NULL) {

  # Rename fauna and microclimate columns to generic names.
  # depth is retained: it is used as a grouping variable in all LME random effects.
  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2]),
      depth          = as.factor(depth)
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  # Split into overlapping windows and retain only sufficiently complete ones.
  # Completeness = complete rows / total rows — does not assume a fixed expected
  # row count, so it remains valid across different depth × variable combinations.
  windows <- create_rolling_windows(sem_data, window_days, step_days)

  valid_windows <- keep(windows, function(w) {
    n_total    <- nrow(w)
    n_complete <- nrow(drop_na(w))
    n_total > 0 && (n_complete / n_total) >= min_rate
  })

  if (length(valid_windows) == 0) {
    message(sprintf(
      "  [skip] window=%d min_rate=%.2f transfo=%s — no windows reach %.0f%% completeness.",
      window_days, min_rate, transfo_type, min_rate * 100
    ))
    return(NULL)
  }

  if (!is.null(n_windows) && length(valid_windows) > n_windows) {
    valid_windows <- valid_windows[sample(seq_along(valid_windows), n_windows)]
  }

  vars_to_process <- c("fauna", "root", "microclimate_1", "microclimate_2")

  results_list <- map(names(valid_windows), function(w_name) {

    win_data <- valid_windows[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date)

    # ── A. Apply transformation within this window ───────────────────────────
    # Hellinger is inherently window-relative (sums to 1 within window).
    # Other transforms are element-wise but applied here for consistency.
    win_data <- win_data %>%
      mutate(across(all_of(vars_to_process), function(x) {
        col <- cur_column()
        switch(transfo_type,
          "sqrt"          = transform_sqrt(x, col),
          "sqrt_plus_one" = transform_ft(x),
          "log"           = transform_log(x, col),
          "hellinger"     = transform_hellinger(x),
          x
        )
      }))

    # ── B. Z-score standardise within this window ────────────────────────────
    # land_use is excluded: it is binary and its scale must remain 0/1
    # for coefficient interpretation.
    win_data <- win_data %>%
      mutate(across(all_of(vars_to_process), ~ as.numeric(scale(.))))

    # ── Guards ────────────────────────────────────────────────────────────────
    stdevs <- sapply(win_data[vars_to_process], sd, na.rm = TRUE)
    if (any(is.na(stdevs) | stdevs == 0)) return(NULL)
    if (length(unique(win_data$land_use)) < 2) return(NULL)
    if (any(!is.finite(as.matrix(win_data[vars_to_process])))) return(NULL)

    # ── C. Fit the four equations independently ───────────────────────────────
    models <- fit_models_independently(win_data)

    # ── D. Extract diagnostics per model ──────────────────────────────────────
    map_dfr(names(models), function(resp_name) {

      m <- models[[resp_name]]
      if (is.null(m)) return(NULL)

      res <- residuals(m, type = "normalized")
      fit <- fitted(m)

      phi_val <- tryCatch(
        coef(m$modelStruct$corStruct, unconstrained = FALSE),
        error = function(e) NA_real_
      )

      r2_vals <- r2_lme(m)

      vif_val <- tryCatch({
        pred_terms <- labels(terms(m))
        cont_terms <- setdiff(pred_terms, "land_use")
        if (length(cont_terms) > 1) {
          f_vif <- as.formula(
            paste(all.vars(formula(m))[1], "~",
                  paste(cont_terms, collapse = " + "))
          )
          max(car::vif(lm(f_vif, data = win_data)))
        } else NA_real_
      }, error = function(e) NA_real_)

      tibble(
        model_id        = paste(taxon_var, w_name, sep = "_"),
        response        = resp_name,
        n_obs           = length(res),
        vif_max         = vif_val,
        hetero_strength = cor(abs(res), fit, method = "spearman",
                              use = "complete.obs"),
        skewness        = moments::skewness(res, na.rm = TRUE),
        ar1_phi         = as.numeric(phi_val),
        r2_marginal     = as.numeric(r2_vals["R2m"]),
        r2_cond         = as.numeric(r2_vals["R2c"])
      )
    })
  }) %>%
    compact()

  if (length(results_list) == 0) return(NULL)
  return(map_dfr(results_list, ~ .x))
}
```


------------------------------------------------------------------------

## 3. Benchmarking

### 3.1 Configuration tester

```{r fn-test-config}
#' Test a single parameter configuration and summarise diagnostic outcomes
#'
#' Transformation type and window parameters are passed to run_sem_pipeline(),
#' which applies them within each window independently. No global scaling
#' is performed at this level.
#'
#' @param w_days       Integer. Window duration (days).
#' @param m_rate_val   Numeric. Minimum data completeness ratio.
#' @param transfo_type Character. One of: "none", "sqrt", "sqrt_plus_one",
#'                    "log", "hellinger".
#' @param data_raw     Full SEM dataframe (raw, untransformed).
#' @param taxon        Character. Fauna variable column name.
#' @param m_indices    Character vector of length 2. Microclimate index names.
#' @return Dataframe with per-response summary statistics, or NULL.
test_configuration <- function(w_days, m_rate_val, transfo_type,
                               data_raw, taxon, m_indices) {

  # run_sem_pipeline handles all transformation and per-window scaling
  res_diag <- run_sem_pipeline(
    taxon_var   = taxon,
    data_raw    = data_raw,
    m_indices   = m_indices,
    transfo_type = transfo_type,
    window_days = w_days,
    min_rate    = m_rate_val,
    n_windows   = 50
  )

  if (is.null(res_diag) || nrow(res_diag) == 0) return(NULL)

  res_diag %>%
    mutate(
      is_ok = abs(hetero_strength) < HETERO_THRESHOLD &
              abs(skewness)        < SKEWNESS_THRESHOLD &
              (is.na(vif_max) | vif_max < VIF_THRESHOLD)
    ) %>%
    group_by(response) %>%
    summarise(
      window             = w_days,
      min_rate           = m_rate_val,
      transformation     = transfo_type,
      hetero_strength    = mean(hetero_strength,       na.rm = TRUE),
      skewness           = mean(skewness,              na.rm = TRUE),
      vif_max            = mean(vif_max,               na.rm = TRUE),
      prop_valid_models  = mean(is_ok,                 na.rm = TRUE),
      avg_r2_marginal    = mean(r2_marginal,           na.rm = TRUE),
      avg_r2_conditional = mean(r2_cond,               na.rm = TRUE),
      delta_r2           = mean(r2_cond - r2_marginal, na.rm = TRUE),
      .groups = "drop"
    )
}
```

### 3.2 Sensitivity grid and execution

```{r benchmarking}
sensibility_grid <- expand.grid(
  w_days     = c(5, 7, 10, 14, 21),
  m_rate_val = c(0.1, 0.2, 0.3, 0.4),
  transfo    = c("none", "sqrt", "sqrt_plus_one", "log", "hellinger"),
  stringsAsFactors = FALSE
)

quiet_test <- purrr::safely(test_configuration)

# Run all configurations; safely() captures errors without stopping execution
safe_results <- sensibility_grid %>%
  purrr::pmap(~ quiet_test(
    ..1, ..2, ..3,
    data_sem_raw, fauna_vars[1], microclimate_index
  ))

n_failed <- sum(map_lgl(safe_results, ~ !is.null(.x$error)))
n_null   <- sum(map_lgl(safe_results, ~ is.null(.x$result)))
if (n_failed > 0) message(n_failed, " configuration(s) returned an error.")
if (n_null   > 0) message(n_null,   " configuration(s) returned no usable windows (NULL).")

benchmark_results <- map_dfr(safe_results, "result")
```

------------------------------------------------------------------------

## 4. Results

### 4.1 Global optimisation

```{r global-optimization}
EXPECTED_RESPONSES <- c("fauna", "root", "microclimate_1", "microclimate_2")

best_per_variable <- benchmark_results %>%
  group_by(window, min_rate, response) %>%
  arrange(desc(prop_valid_models), desc(avg_r2_marginal)) %>%
  slice(1) %>%
  ungroup()

global_optimization <- best_per_variable %>%
  group_by(window, min_rate) %>%
  filter(all(EXPECTED_RESPONSES %in% response)) %>%
  summarise(
    combined_validity      = min(prop_valid_models, na.rm = TRUE),
    mean_r2                = mean(avg_r2_marginal,  na.rm = TRUE),
    transfo_root           = transformation[response == "root"],
    transfo_fauna          = transformation[response == "fauna"],
    transfo_microclimate_1 = transformation[response == "microclimate_1"],
    transfo_microclimate_2 = transformation[response == "microclimate_2"],
    .groups = "drop"
  ) %>%
  mutate(
    mean_r2      = ifelse(is.nan(mean_r2), NA_real_, mean_r2),
    global_score = combined_validity * mean_r2
  ) %>%
  arrange(desc(combined_validity), desc(mean_r2))

global_optimization %>%
  head(5) %>%
  mutate(across(c(combined_validity, mean_r2, global_score), ~ round(., 3))) %>%
  select(
    "Window (days)"          = window,
    "Max missing data"       = min_rate,
    "Transfo root"           = transfo_root,
    "Transfo fauna"          = transfo_fauna,
    "Transfo microclimate 1" = transfo_microclimate_1,
    "Transfo microclimate 2" = transfo_microclimate_2,
    "Validity (joint)"       = combined_validity,
    "R2 Marginal (mean)"     = mean_r2,
    "Score"                  = global_score
  ) %>%
  kable(caption = "Top 5 configurations — ranked by validity then R² (lexicographic)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  row_spec(1, bold = TRUE, background = "#FEC288")

final_selection <- slice(global_optimization, 1)
```

### 4.2 Per-equation summary — selected configuration

```{r per-equation-summary}
sel_window   <- final_selection$window
sel_min_rate <- final_selection$min_rate

sel_transfo <- c(
  fauna          = final_selection$transfo_fauna,
  root           = final_selection$transfo_root,
  microclimate_1 = final_selection$transfo_microclimate_1,
  microclimate_2 = final_selection$transfo_microclimate_2
)

per_eq_stats <- imap_dfr(sel_transfo, function(transfo, resp) {
  benchmark_results %>%
    filter(response == resp, window == sel_window,
           min_rate == sel_min_rate, transformation == transfo)
}) %>%
  mutate(
    response_label = case_match(response,
      "fauna"          ~ "Faunal activity",
      "root"           ~ "Root growth",
      "microclimate_1" ~ microclimate_index[1],
      "microclimate_2" ~ microclimate_index[2],
      .default = response
    )
  ) %>%
  arrange(match(response, c("fauna", "root", "microclimate_1", "microclimate_2")))

validity_vals <- per_eq_stats$prop_valid_models
skewness_vals <- abs(per_eq_stats$skewness)
hetero_vals   <- per_eq_stats$hetero_strength

validity_col <- case_when(
  validity_vals >= 0.70 ~ "#2d6a4f",
  validity_vals >= 0.40 ~ "#c1603a",
  TRUE                  ~ "#b00020"
)

per_eq_stats %>%
  mutate(across(c(prop_valid_models, avg_r2_marginal, avg_r2_conditional,
                  delta_r2, hetero_strength, skewness), ~ round(., 3)),
         vif_max = round(vif_max, 2)) %>%
  select("Equation"    = response_label,
         "Transform"   = transformation,
         "Validity"    = prop_valid_models,
         "R² marg."    = avg_r2_marginal,
         "R² cond."    = avg_r2_conditional,
         "ΔR²"         = delta_r2,
         "Skewness"    = skewness,
         "Heterosced." = hetero_strength,
         "VIF max"     = vif_max) %>%
  kable(align = c("l", "c", "c", "c", "c", "c", "c", "c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, position = "center", font_size = 13) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  column_spec(3, color = validity_col, bold = TRUE) %>%
  column_spec(7, color = if_else(skewness_vals > SKEWNESS_THRESHOLD, "#c1603a", "black"),
              bold = skewness_vals > SKEWNESS_THRESHOLD) %>%
  column_spec(8, color = if_else(hetero_vals > HETERO_THRESHOLD, "#c1603a", "black"),
              bold = hetero_vals > HETERO_THRESHOLD)
```

```{r per-equation-barplot}
#| fig-cap: "Proportion of valid windows per equation under the selected configuration."
#| fig-width: 8
#| fig-height: 3.5

bar_dat <- per_eq_stats %>%
  mutate(
    failure_mode = case_when(
      prop_valid_models >= 0.70                 ~ "Acceptable",
      abs(skewness)   > SKEWNESS_THRESHOLD      ~ "Skewness",
      hetero_strength > HETERO_THRESHOLD        ~ "Heteroscedasticity",
      !is.na(vif_max) & vif_max > VIF_THRESHOLD ~ "Collinearity",
      TRUE                                      ~ "Mixed / other"
    ),
    response_label = factor(response_label, levels = rev(response_label))
  )

failure_colours <- c(
  "Acceptable"         = "#2d6a4f",
  "Skewness"           = "#6b9ac4",
  "Heteroscedasticity" = "#c1603a",
  "Collinearity"       = "#9b5de5",
  "Mixed / other"      = "#adb5bd"
)

ggplot(bar_dat, aes(x = prop_valid_models, y = response_label, fill = failure_mode)) +
  geom_col(width = 0.55, colour = "white", linewidth = 0.3) +
  geom_vline(xintercept = 0.70, linetype = "dashed", colour = "grey30", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", prop_valid_models * 100)),
            hjust = -0.15, size = 3.2, colour = "grey20") +
  scale_x_continuous(limits = c(0, 1.08),
                     labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
  scale_fill_manual(values = failure_colours, name = "Dominant issue") +
  labs(x = "Proportion of windows passing all diagnostics", y = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position    = "right",
        legend.key.size    = unit(0.4, "cm"),
        panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3))
```

### 4.3 Window selection — retain only temporally stable windows

For each rolling window under the selected configuration, we fit all four equations independently and compute the proportion that pass all diagnostic thresholds. Only windows where **at least `MIN_VALID_PROP` of equations are valid** are retained for the visual diagnostics and downstream SEM execution. This removes time periods where the model structure is unreliable regardless of which equation is examined.

```{r window-selection}
# Proportion of equations that must be valid for a window to be retained.
# With 4 equations, 0.75 = at least 3/4 must pass; 1.0 = all four must pass.
MIN_VALID_PROP <- 0.75

# FIX 2: was calling run_benchmark_pipeline (wrong name). This version uses
# run_sem_pipeline throughout — keeping the name consistent here.
message("Running full window scan under selected configuration...")
full_diag <- suppressMessages(
  run_sem_pipeline(
    taxon_var    = fauna_vars[1],
    data_raw     = data_sem_raw,
    m_indices    = microclimate_index,
    transfo_type = sel_transfo["fauna"],
    window_days  = sel_window,
    min_rate     = sel_min_rate,
    n_windows    = NULL
  )
)

# Per-window: proportion of equations passing all thresholds
window_validity <- full_diag %>%
  mutate(is_ok = abs(hetero_strength) < HETERO_THRESHOLD &
                 abs(skewness)        < SKEWNESS_THRESHOLD &
                 (is.na(vif_max) | vif_max < VIF_THRESHOLD)) %>%
  group_by(model_id) %>%
  summarise(
    prop_eq_valid = mean(is_ok, na.rm = TRUE),
    n_eq          = n(),
    .groups = "drop"
  ) %>%
  mutate(
    # model_id = "taxon_YYYY-MM-DD" — extract date portion
    date_start   = as.Date(str_extract(model_id, "\\d{4}-\\d{2}-\\d{2}")),
    window_valid = prop_eq_valid >= MIN_VALID_PROP
  )

selected_windows <- window_validity %>%
  filter(window_valid) %>%
  pull(date_start)

message(sprintf(
  "Window selection: %d / %d windows retained (>= %.0f%% of equations valid).",
  length(selected_windows),
  nrow(window_validity),
  MIN_VALID_PROP * 100
))

# Temporal distribution of valid windows
ggplot(window_validity, aes(x = date_start, y = prop_eq_valid, colour = window_valid)) +
  geom_point(size = 0.8, alpha = 0.6) +
  geom_hline(yintercept = MIN_VALID_PROP, linetype = "dashed",
             colour = "grey30", linewidth = 0.5) +
  scale_colour_manual(values = c("TRUE" = "#2d6a4f", "FALSE" = "#c1603a"),
                      labels = c("TRUE" = "Retained", "FALSE" = "Dropped"),
                      name = NULL) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  labs(x = NULL,
       y = "Proportion of valid equations per window",
       title = "Window-level diagnostic validity",
       subtitle = sprintf("Threshold = %.0f%% | %d windows retained out of %d",
                          MIN_VALID_PROP * 100,
                          length(selected_windows),
                          nrow(window_validity))) +
  theme_classic(base_size = 11) +
  theme(legend.position    = "top",
        panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3))
```

------------------------------------------------------------------------

## 5. Visual diagnostics — selected configuration

```{r fn-check-panels}
#' Build a 6-panel check_model-style diagnostic figure for one SEM equation.
#'
#' @param eq_name       Character. Internal response name.
#' @param eq_label      Character. Display label for plot titles.
#' @param resid_data    Dataframe of per-observation diagnostics.
#' @param vif_data      Dataframe of per-window VIF values.
#' @param n_ppc_windows Integer. Number of windows shown in PPC panel.
#' @param cook_thresh   Numeric vector. Cook's D contour levels.
#' @return A patchwork (3 rows × 2 columns), or NULL if no data.
make_check_panels <- function(eq_name, eq_label,
                               resid_data, vif_data,
                               n_ppc_windows = 80,
                               cook_thresh   = c(0.5, 1.0)) {
  col_obs  <- "#3d8c40"
  col_fit  <- "#6b9ac4"
  col_line <- "#c1603a"
  base_sz  <- 8

  eq_res <- filter(resid_data, response == eq_name, is.finite(residual))
  eq_vif <- filter(vif_data,   response == eq_name)

  n_wins <- n_distinct(eq_res$date_start)
  n_fix  <- switch(eq_name, microclimate_1 = 2L, microclimate_2 = 2L,
                   root = 4L, fauna = 5L, 2L)

  th <- theme_classic(base_size = base_sz) +
    theme(
      panel.border     = element_rect(color = "grey40", fill = NA, linewidth = 0.3),
      panel.grid.major = element_line(color = "grey94", linewidth = 0.15),
      axis.title       = element_text(size = base_sz - 1),
      plot.title       = element_text(face = "bold", size = base_sz, margin = margin(b = 1)),
      plot.subtitle    = element_text(color = "grey50", size = base_sz - 1.5,
                                      margin = margin(b = 3))
    )

  all_wins <- distinct(eq_res, date_start)$date_start
  if (length(all_wins) == 0) return(NULL)
  ppc_wins <- sample(all_wins, min(length(all_wins), n_ppc_windows))

  ppc_data <- filter(eq_res, date_start %in% ppc_wins, is.finite(fitted_subj))
  obs_data <- filter(eq_res, is.finite(observed))

  y_data_max <- max(
    max(density(obs_data$observed)$y),
    tapply(ppc_data$fitted_subj, ppc_data$date_start,
           \(x) max(density(x)$y)) |> unlist() |> max()
  )
  y_upper <- min(y_data_max * 1.10, 30)

  p1 <- ggplot() +
    geom_density(data = ppc_data, aes(x = fitted_subj, group = date_start),
                 color = col_fit, alpha = 0.20, linewidth = 0.25) +
    geom_density(data = obs_data, aes(x = observed),
                 color = col_obs, linewidth = 0.9) +
    annotate("text", x = -Inf, y = Inf,
             label = paste0(length(ppc_wins), " windows shown"),
             hjust = -0.08, vjust = 1.5, size = 2.2, color = "grey50", fontface = "italic") +
    coord_cartesian(ylim = c(0, y_upper)) +
    labs(title = "Posterior Predictive Check",
         subtitle = "Model-predicted lines should resemble observed data line",
         x = "Value", y = "Density") + th

  lin_dat <- eq_res %>% filter(is.finite(fitted_pop)) %>%
    { if (nrow(.) > 25000) slice_sample(., n = 25000) else . }

  p2 <- ggplot(lin_dat, aes(x = fitted_pop, y = residual)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.4) +
    geom_point(size = 0.2, alpha = 0.05, color = col_fit) +
    geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
                color = col_line, fill = col_line, alpha = 0.15, linewidth = 0.6) +
    labs(title = "Linearity", subtitle = "Reference line should be flat and horizontal",
         x = "Fitted values", y = "Residuals") + th

  p3 <- ggplot(mutate(lin_dat, sqr = sqrt(abs(residual))),
               aes(x = fitted_pop, y = sqr)) +
    geom_point(size = 0.2, alpha = 0.05, color = col_fit) +
    geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
                color = col_line, fill = col_line, alpha = 0.15, linewidth = 0.6) +
    labs(title = "Homogeneity of Variance", subtitle = "Reference line should be flat",
         x = "Fitted values", y = expression(sqrt("|Std. residuals|"))) + th

  inf_dat <- eq_res %>% filter(is.finite(leverage)) %>%
    { if (nrow(.) > 25000) slice_sample(., n = 25000) else . }

  has_leverage <- nrow(inf_dat) > 0 &&
    is.finite(quantile(inf_dat$leverage, 0.995, na.rm = TRUE))

  if (has_leverage) {
    max_h  <- quantile(inf_dat$leverage, 0.995, na.rm = TRUE)
    h_seq  <- seq(1e-4, max(max_h * 1.05, 1e-3), length.out = 400)
    r_clip <- quantile(abs(inf_dat$residual), 0.999, na.rm = TRUE) * 1.5

    cook_lines <- map_dfr(cook_thresh, function(d) {
      r_val <- sqrt(d * n_fix * (1 - h_seq)^2 / h_seq)
      bind_rows(tibble(h = h_seq, r =  r_val, level = as.character(d)),
                tibble(h = h_seq, r = -r_val, level = as.character(d)))
    }) %>% filter(is.finite(r), abs(r) < r_clip)

    cook_labels <- cook_lines %>% filter(r > 0) %>%
      group_by(level) %>% slice_max(h, n = 1)

    p4 <- ggplot(inf_dat, aes(x = leverage, y = residual)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.4) +
      geom_point(size = 0.2, alpha = 0.05, color = col_fit) +
      geom_line(data = cook_lines,
                aes(x = h, y = r, group = interaction(level, r > 0)),
                color = col_line, linewidth = 0.45, linetype = "dashed") +
      geom_text(data = cook_labels, aes(x = h, y = r, label = level),
                size = 2.4, color = col_line, hjust = -0.1, vjust = 0) +
      labs(title = "Influential Observations",
           subtitle = "Points should be inside the contour lines",
           x = "Leverage", y = "Std. Residuals") + th
  } else {
    cd_dat <- eq_res %>% filter(is.finite(cooks_d), is.finite(fitted_pop)) %>%
      { if (nrow(.) > 25000) slice_sample(., n = 25000) else . }

    p4 <- ggplot(cd_dat, aes(x = fitted_pop, y = cooks_d)) +
      geom_hline(yintercept = c(0.5, 1.0), linetype = "dashed",
                 color = col_line, linewidth = 0.35) +
      geom_point(size = 0.2, alpha = 0.06, color = col_fit) +
      annotate("text", x = Inf, y = c(0.52, 1.02), label = c("0.5", "1.0"),
               hjust = 1.1, size = 2.4, color = col_line) +
      labs(title = "Influential Observations",
           subtitle = "Leverage unavailable — showing Cook's D vs fitted",
           x = "Fitted values (population level)", y = "Cook's D") + th
  }

  if (nrow(eq_vif) == 0) {
    p5 <- ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = "Single predictor\n(VIF not applicable)",
               size = 3.2, color = "grey50", hjust = 0.5) +
      scale_x_continuous(limits = c(0, 1)) + scale_y_continuous(limits = c(0, 1)) +
      labs(title = "Collinearity") +
      th + theme(axis.text = element_blank(), axis.ticks = element_blank(),
                 axis.title = element_blank())
  } else {
    vif_summ <- eq_vif %>%
      group_by(predictor) %>%
      summarise(mean_vif = mean(vif, na.rm = TRUE), sd_vif = sd(vif, na.rm = TRUE),
                .groups = "drop") %>%
      mutate(predictor = fct_reorder(predictor, mean_vif), hi = mean_vif >= 10)

    p5 <- ggplot(vif_summ, aes(x = mean_vif, y = predictor)) +
      annotate("rect", xmin = c(1, 5, 10), xmax = c(5, 10, Inf),
               ymin = -Inf, ymax = Inf,
               fill = c("#d4edda", "#fff3cd", "#f8d7da"), alpha = 0.5) +
      geom_vline(xintercept = c(5, 10), linetype = "dashed",
                 color = "grey50", linewidth = 0.35) +
      geom_errorbarh(aes(xmin = pmax(mean_vif - sd_vif, 1), xmax = mean_vif + sd_vif),
                     height = 0.3, color = "grey40", linewidth = 0.45) +
      geom_point(aes(color = hi), size = 2.8) +
      scale_color_manual(values = c("FALSE" = "#3d5a3e", "TRUE" = col_line), guide = "none") +
      scale_x_log10(labels = scales::comma) +
      labs(title = "Collinearity",
           subtitle = "High VIF (>= 10) may inflate parameter uncertainty",
           x = "VIF (log-scaled, mean +/- SD across windows)", y = NULL) + th
  }

  qq_dat <- if (nrow(eq_res) > 40000) slice_sample(eq_res, n = 40000) else eq_res

  p6 <- ggplot(qq_dat, aes(sample = residual)) +
    geom_hline(yintercept = c(-2, 2), color = "grey70", linewidth = 0.3, linetype = "dashed") +
    stat_qq_line(color = col_line, linewidth = 0.55) +
    stat_qq(size = 0.15, alpha = 0.08, color = col_fit) +
    labs(title = "Normality of Residuals", subtitle = "Dots should fall along the line",
         x = "Standard Normal Quantiles", y = "Sample Quantile Deviations") + th

  (p1 | p2) / (p3 | p4) / (p5 | p6) +
    plot_annotation(
      title    = paste0("Model diagnostics — equation: ", eq_label),
      subtitle = sprintf("Pooled across %d windows  |  n = %s observations",
                         n_wins, scales::comma(nrow(eq_res))),
      theme = theme(
        plot.title    = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(color = "grey50", size = 8, margin = margin(b = 6))
      )
    )
}

```

```{r extract-selected-diagnostics}
#' Re-fit models under the selected configuration on the retained windows
#' and extract per-observation diagnostics for check_model figures.
#' Only windows in `selected_windows` (from the window-selection step) are used.
#'
#' Uses the same completeness criterion as run_sem_pipeline:
#' complete rows / total rows >= sel_min_rate.
extract_selected_diag <- function(resp_name, transfo, data_raw, n_sample = 100) {
  sem_data <- data_raw %>%
    mutate(fauna          = !!sym(fauna_vars[1]),
           microclimate_1 = !!sym(microclimate_index[1]),
           microclimate_2 = !!sym(microclimate_index[2])) %>%
    # FIX 3: depth must be retained — it is the grouping variable for microclimate
    # models and the inner grouping level for root/fauna models.
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  vars_to_scale <- c("fauna", "root", "microclimate_1", "microclimate_2")

  windows <- create_rolling_windows(sem_data, sel_window)

  # Same completeness criterion as run_sem_pipeline.
  valid_wins <- keep(windows, function(w) {
    n_total    <- nrow(w)
    n_complete <- nrow(drop_na(w))
    n_total > 0 && (n_complete / n_total) >= sel_min_rate
  })

  # Restrict to windows that passed the equation-level validity filter.
  valid_wins <- valid_wins[as.Date(names(valid_wins)) %in% selected_windows]

  if (length(valid_wins) == 0) return(NULL)
  sampled <- valid_wins[sample(seq_along(valid_wins), min(length(valid_wins), n_sample))]

  map_dfr(names(sampled), function(w_name) {
    win_data <- sampled[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date) %>%
      mutate(across(all_of(vars_to_scale), function(x) {
        switch(transfo,
          sqrt          = transform_sqrt(x, cur_column()),
          sqrt_plus_one = transform_ft(x),
          log           = transform_log(x, cur_column()),
          hellinger     = transform_hellinger(x),
          x
        )
      })) %>%
      mutate(across(all_of(vars_to_scale), ~ as.numeric(scale(.))))

    stdevs <- sapply(win_data[vars_to_scale], sd, na.rm = TRUE)
    if (any(is.na(stdevs) | stdevs == 0))                   return(NULL)
    if (length(unique(win_data$land_use)) < 2)               return(NULL)
    if (any(!is.finite(as.matrix(win_data[vars_to_scale])))) return(NULL)

    m <- fit_models_independently(win_data)[[resp_name]]
    if (is.null(m)) return(NULL)
    extract_lme_diag(m, w_name, resp_name)
  })
}

message("Extracting per-observation diagnostics for selected configuration...")
selected_resid <- imap_dfr(sel_transfo,
                            ~ extract_selected_diag(.y, .x, data_sem_raw))
message(sprintf("Done: %d rows across %d equations",
                nrow(selected_resid), n_distinct(selected_resid$response)))

# VIF from benchmark results (stored as max per window — used for collinearity panel)
selected_vif <- imap_dfr(sel_transfo, function(transfo, resp) {
  benchmark_results %>%
    filter(response == resp, window == sel_window,
           min_rate == sel_min_rate, transformation == transfo) %>%
    transmute(response = resp, predictor = resp, vif = vif_max)
}) %>% filter(!is.na(vif))
```

```{r fig-check-model, fig.width=18, fig.height=14, out.width="100%"}
#| fig-cap: >
#|   Model assumption diagnostics for each SEM equation under the selected
#|   configuration (window and transformation from model_parameters.txt).

response_labels <- c(
  microclimate_1 = microclimate_index[1],
  microclimate_2 = microclimate_index[2],
  root           = "Root growth",
  fauna          = "Faunal activity"
)

check_figs <- imap(response_labels, ~ make_check_panels(
  eq_name    = .y,
  eq_label   = .x,
  resid_data = selected_resid,
  vif_data   = selected_vif
))

walk(check_figs, print)

iwalk(check_figs, ~ ggsave(
  sprintf("../output/Fig_diagnostics_%s.tiff", .y), plot = .x,
  width = 17.5, height = 13.5, units = "cm", dpi = 300, compression = "lzw"
))
message("Diagnostic figures saved for: ", paste(names(check_figs), collapse = ", "))
```

------------------------------------------------------------------------

## 6. Export

```{r export}
write.csv(benchmark_results, "../output/benchmark_results_database.csv",
          row.names = FALSE, quote = TRUE)

# Selected window dates — read by script 3 to restrict SEM execution
writeLines(as.character(selected_windows), "../output/selected_windows.txt")

params_to_save <- list(
  selected_window          = final_selection$window,
  selected_min_rate        = final_selection$min_rate,
  transform_root           = final_selection$transfo_root,
  transform_fauna          = final_selection$transfo_fauna,
  transform_microclimate_1 = final_selection$transfo_microclimate_1,
  transform_microclimate_2 = final_selection$transfo_microclimate_2
)

jsonlite::write_json(params_to_save, "../output/model_parameters.txt", pretty = TRUE)

message(sprintf(
  "Best config: window = %d days | min_rate = %.1f | validity = %.3f | R2 = %.3f | score = %.3f\n  root: %s | fauna: %s | mc1: %s | mc2: %s",
  as.integer(final_selection$window), final_selection$min_rate,
  final_selection$combined_validity, final_selection$mean_r2, final_selection$global_score,
  final_selection$transfo_root, final_selection$transfo_fauna,
  final_selection$transfo_microclimate_1, final_selection$transfo_microclimate_2
))
```
