---
title: "SEM — model execution and path coefficient extraction"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: true
editor:
  markdown:
    wrap: 72
---

# Overview

## Role in the pipeline

This script is the **third step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — builds the full SEM database
2.  `sem_diagnostic.qmd` — selects the optimal window, completeness
    threshold, and transformations

## What this script produces

| Output | File | Description |
|------------------------|------------------------|------------------------|
| Observed path coefficients | `SEM_results_database.csv` | Standardised estimates, p-values, R², and system-specific effects for all taxon × window combinations |

## Statistical approach

The causal model has four structural equations organised in three tiers,
with `land_use` (0 = unmanaged / A, 1 = managed / C) as the exogenous
binary driver. The model is **additive** at this stage — no interaction
terms between `land_use` and the continuous predictors are included.
This means each path coefficient represents the average slope across
both management contexts rather than a system-specific estimate.

**Tier 1 — land use shapes microclimate** (additive):

$$\text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$
$$\text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$

**Tier 2 — microclimate drives root:**

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu}$$

**Tier 3 — microclimate and root drive fauna:**

$$\text{Fauna} \leftarrow \beta_6 \cdot \text{mc}_1 + \beta_7 \cdot \text{mc}_2 + \beta_8 \cdot \text{Root} + \beta_9 \cdot \text{lu}$$

*where mc = microclimate index, lu = land_use.*

The residual covariance `microclimate_1 %~~% microclimate_2` is modelled
explicitly because both indices share unmeasured physical drivers
(radiation, topography) beyond the binary land-use contrast. Without
this term, piecewiseSEM issues missing-path claims and inflates
d-separation test statistics.

Each equation uses a **nested random intercept** (`orientation / depth`)
and an **AR(1) correlation structure** to account for temporal
autocorrelation within scanner groups.

**Per-window standardisation.** Transformation and z-score scaling are
both applied independently within each rolling window, after subsetting
and dropping incomplete observations. This ensures that standardised
coefficients reflect within-window variance only, avoiding the confound
of seasonal and inter-annual variation that a global scaling would
introduce.

------------------------------------------------------------------------

## 1. Setup

```{r setup}
#| message: false

invisible(Sys.setlocale("LC_TIME", "C"))

libs <- c(
  "tidyverse",
  "lubridate",
  "piecewiseSEM",
  "performance",
  "lme4",
  "purrr",
  "nlme",
  "moments",
  "car",
  "jsonlite",
  "knitr",
  "kableExtra",
  "patchwork",    
  "scales" 
)
invisible(lapply(libs, require, character.only = TRUE))
```

------------------------------------------------------------------------

## 2. Parameter loading

```{r load-params}
if (file.exists("../output/model_parameters.txt")) {

  best_params            <- read_json("../output/model_parameters.txt")
  WIN_DAYS               <- as.numeric(best_params$selected_window)
  MIN_RATE               <- as.numeric(best_params$selected_min_rate)
  TRANSFO_ROOT           <- as.character(best_params$transform_root)
  TRANSFO_FAUNA          <- as.character(best_params$transform_fauna)
  TRANSFO_MICROCLIMATE_1 <- as.character(best_params$transform_microclimate_1)
  TRANSFO_MICROCLIMATE_2 <- as.character(best_params$transform_microclimate_2)

} else {
  WIN_DAYS               <- 14
  MIN_RATE               <- 0.3
  TRANSFO_ROOT           <- "sqrt_plus_one"
  TRANSFO_FAUNA          <- "sqrt_plus_one"
  TRANSFO_MICROCLIMATE_1 <- "none"
  TRANSFO_MICROCLIMATE_2 <- "none"
  warning("Configuration file not found. Using default fallback parameters.")
}

message(sprintf(
  paste0("Active config — Window: %d days | Min. rate: %.1f\n",
         "  Transfo root: %s | Transfo fauna: %s\n",
         "  Transfo microclimate_1: %s | Transfo microclimate_2: %s"),
  WIN_DAYS, MIN_RATE,
  TRANSFO_ROOT, TRANSFO_FAUNA,
  TRANSFO_MICROCLIMATE_1, TRANSFO_MICROCLIMATE_2
))
```

------------------------------------------------------------------------

## 3. Data loading and preprocessing

Transformations are applied globally here (they are element-wise and
order-independent). Z-score standardisation is deferred to within each
rolling window — see `run_sem_pipeline()`.

```{r data-preparation}

data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep        = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    # land_use is binary and intentionally excluded from z-scoring throughout.
    # Keeping it on the 0/1 scale means β_land_use is interpretable as the
    # average A → C contrast on the standardised response scale.
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    root     = root_growth_cm2_count_24h
  )

fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")

#' Apply a named variance-stabilising transformation to a numeric vector.
#'
#' For `sqrt` and `log`, a minimum shift is applied automatically when the
#' input contains non-positive values (e.g. negative temperatures, zero
#' humidity). The shift equals |min(x)| + 0.001, bringing the minimum to
#' +0.001 before transforming. This preserves the distribution shape while
#' avoiding NaN / -Inf output. A message is emitted when a shift is applied
#' so the adjustment is always visible in the render log.
#'
#' `hellinger` is window-relative and is handled inside run_sem_pipeline();
#' this function returns x unchanged for that method.
#'
#' @param x      Numeric vector.
#' @param method One of "none", "sqrt", "sqrt_plus_one", "log", "hellinger".
#' @param label  Optional label used in the shift message.
#' @return Transformed numeric vector of the same length.
apply_transfo <- function(x, method, label = "") {

  shift_if_needed <- function(x, method_name) {
    x_min <- min(x, na.rm = TRUE)
    if (x_min <= 0) {
      shift <- abs(x_min) + 0.001
      message(sprintf(
        "  [apply_transfo] shifting '%s' by +%.4f before %s (min was %.4f)",
        label, shift, method_name, x_min
      ))
      x <- x + shift
    }
    x
  }

  switch(method,
    "sqrt_plus_one" = sqrt(x) + sqrt(x + 1),
    "sqrt"          = sqrt(shift_if_needed(x, "sqrt")),
    "log"           = log1p(shift_if_needed(x, "log")),
    "hellinger"     = x,   # handled per-window inside run_sem_pipeline()
    x
  )
}

# Apply transformations globally (except hellinger, which is window-relative).
# Scaling is NOT performed here — it happens per window inside run_sem_pipeline().
# The label argument triggers a message if a shift is needed (negative values).
data_transformed <- data_sem_raw %>%
  mutate(root = apply_transfo(root, TRANSFO_ROOT, "root")) %>%
  mutate(across(
    all_of(intersect(fauna_vars, names(.))),
    ~ apply_transfo(., TRANSFO_FAUNA, cur_column())
  )) %>%
  mutate(across(
    all_of(microclimate_index[1]),
    ~ apply_transfo(., TRANSFO_MICROCLIMATE_1, microclimate_index[1])
  )) %>%
  mutate(across(
    all_of(microclimate_index[2]),
    ~ apply_transfo(., TRANSFO_MICROCLIMATE_2, microclimate_index[2])
  ))
```

------------------------------------------------------------------------

## 4. Analytical functions

### 4.1 Rolling window splitter

```{r rolling-window-generator}
#' Split a time series into overlapping rolling windows.
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Window duration in days.
#' @param step_days   Integer. Step between consecutive window starts.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )
  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 4.2 SEM structure

No interaction terms — each path coefficient is an average across both
management contexts. `land_use` enters each equation as a main effect
only, capturing the residual mean difference between A and C after
conditioning on the continuous predictors.

```{r SEM-structure}
#' Fit the additive piecewise SEM for a single data window.
#'
#' @param data Dataframe (already transformed and z-scored within window)
#'             with columns: land_use, root, fauna, microclimate_1,
#'             microclimate_2, orientation, depth.
#' @return A psem object, or NULL on convergence failure.
fit_psem <- function(data) {
  psem(

    lme(microclimate_1 ~ land_use,
        data        = data,
        random      = ~ 1 | depth,
        correlation = corAR1(form = ~ 1 | depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(microclimate_2 ~ land_use,
        data        = data,
        random      = ~ 1 | depth,
        correlation = corAR1(form = ~ 1 | depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(root ~ microclimate_1 + microclimate_2 + land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(fauna ~ microclimate_1 + microclimate_2 + root + land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    microclimate_1 %~~% microclimate_2
  )
}
```

### 4.3 Main pipeline

```{r fixed-pipeline}
#' Run the full rolling-window SEM pipeline for one taxon variable.
#'
#' Within each window the following steps are applied in order:
#'   1. Subset and drop incomplete rows
#'   2. Apply hellinger transformation if selected (window-relative)
#'   3. Z-score standardise all numeric variables except land_use
#'   4. Fit the piecewise SEM
#'   5. Extract path coefficients, R², and normalised residuals
#'
#' @param taxon_var   Character. Fauna variable column name.
#' @param data_raw    Dataframe. Transformed (non-hellinger) SEM database.
#' @param m_indices   Character vector of length 2. Microclimate column names.
#' @param window_days Integer. Window duration (days).
#' @param min_rate    Numeric [0,1]. Minimum data completeness ratio.
#' @return List with two dataframes: `coefficients` and `residuals`.
run_sem_pipeline <- function(taxon_var, data_raw, m_indices,
                             window_days, min_rate) {

  cat("\n--- Processing taxon:", taxon_var, "---\n")

  vars_to_scale <- c("fauna", "root", "microclimate_1", "microclimate_2")

  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2]),
      depth          = as.factor(depth)
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  windows    <- create_rolling_windows(sem_data, window_days, step_days = 1)
  valid_wins <- keep(windows,
                     ~ (nrow(drop_na(.x)) / (window_days * 16 * 2)) >= min_rate)

  cat(sprintf("  %d / %d windows passed the completeness filter.\n",
              length(valid_wins), length(windows)))

  if (length(valid_wins) == 0) return(NULL)

  results <- map(names(valid_wins), function(w_name) {

    win_data <- valid_wins[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date)

    # ── Per-window hellinger (window-relative, applied after subsetting) ─────
    # Sparsity guard: if more than 80% of fauna values are zero, hellinger is
    # numerically unstable — a few non-zero observations receive extreme
    # relative weights after sqrt(x/sum), producing residual outliers.
    # Such windows carry no reliable faunal signal and are skipped entirely.
    if (TRANSFO_FAUNA == "hellinger") {
      if (mean(win_data$fauna == 0, na.rm = TRUE) > 0.8) return(NULL)
      win_data <- win_data %>%
        mutate(fauna = sqrt(fauna / sum(fauna, na.rm = TRUE)))
    }

    # ── Per-window z-score standardisation (land_use excluded) ───────────────
    win_data <- win_data %>%
      mutate(across(all_of(vars_to_scale), ~ as.numeric(scale(.))))

    # ── Guards ────────────────────────────────────────────────────────────────
    # sd() returns NA (not 0) when all values in a column are NA — this can
    # happen in sparse taxa windows after drop_na(). NA == 0 yields NA,
    # and if(NA) crashes. The is.na() check catches both cases.
    stdevs <- sapply(win_data[vars_to_scale], sd, na.rm = TRUE)
    if (any(is.na(stdevs) | stdevs == 0))            return(NULL)
    if (length(unique(win_data$land_use)) < 2)        return(NULL)
    if (any(!is.finite(as.matrix(win_data[vars_to_scale])))) return(NULL)

    # ── Fit SEM ───────────────────────────────────────────────────────────────
    model <- tryCatch(
      withCallingHandlers(
        fit_psem(win_data),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
    if (is.null(model)) return(NULL)

    sm <- tryCatch(summary(model), error = function(e) NULL)
    if (is.null(sm)) return(NULL)

    # ── Path coefficients ─────────────────────────────────────────────────────
    coefs  <- as.data.frame(sm$coefficients)
    r2_raw <- as.data.frame(sm$R2)
    r2_df  <- r2_raw %>%
      mutate(
        r2_cond = .[[grep("R.squared|Conditional", names(.))[1]]],
        r2_marg = .[[grep("Marginal",              names(.))[1]]]
      ) %>%
      select(response = Response, r2_cond, r2_marg)

    coef_df <- tibble(
      taxon        = taxon_var,
      date_start   = as.Date(w_name),
      n_obs        = sm$n,
      response     = coefs$Response,
      predictor    = coefs$Predictor,
      estimate     = coefs$Estimate,
      std_estimate = coefs$Std.Estimate,
      p_value      = suppressWarnings(as.numeric(gsub("[^0-9.]", "", coefs$P.Value)))
    ) %>%
      left_join(r2_df, by = "response")

    # ── Per-model diagnostic quantities (for check_model-style figures) ─────────
    # Extracted from each constituent lme independently so cross-equation
    # convergence failures do not block residual extraction.
    #
    # Quantities extracted per observation:
    #   residual    — normalised residual (divided by within-group SD estimate)
    #   observed    — response values as seen by the model
    #   fitted_pop  — population-level prediction (fixed effects only, level = 0)
    #   fitted_subj — subject-level prediction (fixed + all random effects)
    #   leverage    — hat value (h_ii); used for influential-observation panel
    #   cooks_d     — Cook's D approximation: r² × h / (p × (1 − h)²)
    lme_mods <- keep(model, inherits, "lme")

    resid_df <- map_dfr(lme_mods, function(m) {
      resp_name <- all.vars(formula(m))[1]
      res_norm  <- residuals(m, type = "normalized")
      n_m       <- length(res_norm)
      fit_pop   <- tryCatch(as.numeric(fitted(m, level = 0)),
                            error = function(e) rep(NA_real_, n_m))
      fit_subj  <- tryCatch(as.numeric(fitted(m, level = length(m$groups))),
                            error = function(e) rep(NA_real_, n_m))
      obs_vals  <- tryCatch(as.numeric(nlme::getResponse(m)),
                            error = function(e) rep(NA_real_, n_m))
      lev       <- tryCatch(as.numeric(hatvalues(m)),
                            error = function(e) rep(NA_real_, n_m))
      n_fix     <- length(nlme::fixef(m))
      cook_d    <- (res_norm^2 * lev) / (n_fix * pmax((1 - lev)^2, 1e-6))

      tibble(
        taxon       = taxon_var,
        date_start  = as.Date(w_name),
        response    = resp_name,
        observed    = obs_vals,
        fitted_pop  = fit_pop,
        fitted_subj = fit_subj,
        residual    = res_norm,
        leverage    = lev,
        cooks_d     = cook_d
      )
    })
 # ── VIF per predictor per equation per window ─────────────────────────────
    # land_use excluded (binary exogenous driver); VIF only meaningful when
    # there are at least two continuous predictors.
    vif_df <- map_dfr(lme_mods, function(m) {
      resp_name  <- all.vars(formula(m))[1]
      pred_terms <- labels(terms(m))
      cont_terms <- setdiff(pred_terms, "land_use")
      if (length(cont_terms) <= 1) return(NULL)
      tryCatch({
        f_vif    <- as.formula(paste(resp_name, "~",
                                     paste(pred_terms, collapse = " + ")))
        vif_vals <- car::vif(lm(f_vif, data = win_data))
        tibble(
          taxon      = taxon_var,
          date_start = as.Date(w_name),
          response   = resp_name,
          predictor  = names(vif_vals),
          vif        = as.numeric(vif_vals)
        )
      }, error = function(e) NULL)
    })

    list(coefficients = coef_df, residuals = resid_df, vif = vif_df)
  })

  results <- compact(results)
  if (length(results) == 0) return(NULL)

  list(
    coefficients = map_dfr(results, "coefficients"),
    residuals    = map_dfr(results, "residuals"),
    vif          = map_dfr(results, "vif")
  )
}
```

------------------------------------------------------------------------

## 5. Model execution

```{r observed-models}
message(">>> Running observed models across all taxa...")

pipeline_output <- map(
  fauna_vars,
  ~ run_sem_pipeline(.x, data_transformed, microclimate_index,
                     WIN_DAYS, MIN_RATE)
) %>%
  compact()

sem_results_raw <- map_dfr(pipeline_output, "coefficients")
all_residuals   <- map_dfr(pipeline_output, "residuals")
all_vif         <- map_dfr(pipeline_output, "vif")

message(sprintf("Observed models complete: %d coefficient rows across %d taxa.",
                nrow(sem_results_raw), n_distinct(sem_results_raw$taxon)))
```

------------------------------------------------------------------------

## 6. Residual diagnostics

Six assumption checks are shown per equation, mirroring
`performance::check_model()`. Because the pipeline runs \~500 windows,
calling `check_model()` on each would be impractical. Instead, all
windows' diagnostic quantities are pooled, giving a view of assumption
behaviour across the full temporal range rather than for a single
window.

| Panel | What to look for |
|------------------------------------|------------------------------------|
| Posterior Predictive Check | Blue per-window curves should envelope the green observed density |
| Linearity | Smoother should track the dashed zero line |
| Homogeneity of Variance | Smoother should be flat; slope indicates heteroscedasticity |
| Influential Observations | Most points should lie inside the Cook's D = 0.5 contour |
| Collinearity | VIF (mean ± SD across windows); red = high (≥ 10) |
| Normality | Points should follow the diagonal |

```{r fn-check-panels}
#' Build a 6-panel check_model-style diagnostic figure for one SEM equation.
#'
#' All panels are built from the full set of rolling-window model fits pooled
#' together, so they capture assumption behaviour across the entire time series.
#'
#' @param eq_name       Character. Internal response name (matches all_residuals).
#' @param eq_label      Character. Display label for plot titles.
#' @param resid_data    Dataframe. all_residuals from run_sem_pipeline().
#' @param vif_data      Dataframe. all_vif from run_sem_pipeline().
#' @param n_ppc_windows Integer. Number of windows shown in PPC panel.
#' @param cook_thresh   Numeric vector. Cook's D contour levels.
#' @return A patchwork (3 rows x 2 columns).
make_check_panels <- function(eq_name, eq_label,
                               resid_data, vif_data,
                               n_ppc_windows = 80,
                               cook_thresh   = c(0.5, 1.0)) {

  col_obs  <- "#3d8c40"
  col_fit  <- "#6b9ac4"
  col_line <- "#c1603a"
  base_sz  <- 8

  eq_res <- resid_data %>% filter(response == eq_name, is.finite(residual))
  eq_vif <- vif_data   %>% filter(response == eq_name)

  n_wins_total <- n_distinct(eq_res$date_start)
  n_taxa_total <- n_distinct(eq_res$taxon)

  n_fix <- switch(eq_name,
    microclimate_1 = 2L, microclimate_2 = 2L,
    root = 4L, fauna = 5L, 2L)

  th <- theme_classic(base_size = base_sz) +
    theme(
      panel.border     = element_rect(color = "grey40", fill = NA, linewidth = 0.3),
      panel.grid.major = element_line(color = "grey94", linewidth = 0.15),
      axis.title       = element_text(size = base_sz - 1),
      plot.title       = element_text(face = "bold", size = base_sz,
                                      margin = margin(b = 1)),
      plot.subtitle    = element_text(color = "grey50", size = base_sz - 1.5,
                                      margin = margin(b = 3))
    )

  # ── 1. Posterior Predictive Check ────────────────────────────────────────
  # geom_density with group = date_start draws one curve per window (thin blue).
  # The thick green curve is the overall observed density across all windows.
  all_wins <- eq_res %>% distinct(date_start) %>% pull(date_start)
  if (length(all_wins) == 0) return(NULL)  # guard: no data for this equation
  ppc_wins <- sample(all_wins, min(length(all_wins), n_ppc_windows))

  p1 <- ggplot() +
    geom_density(
      data  = eq_res %>% filter(date_start %in% ppc_wins, is.finite(fitted_subj)),
      aes(x = fitted_subj, group = date_start),
      color = col_fit, alpha = 0.20, linewidth = 0.25
    ) +
    geom_density(
      data  = eq_res %>% filter(is.finite(observed)),
      aes(x = observed),
      color = col_obs, linewidth = 0.9
    ) +
    annotate("text", x = -Inf, y = Inf,
             label    = paste0(length(ppc_wins), " windows shown"),
             hjust = -0.08, vjust = 1.5, size = 2.2,
             color = "grey50", fontface = "italic") +
    labs(title    = "Posterior Predictive Check",
         subtitle = "Model-predicted lines should resemble observed data line",
         x = "Value", y = "Density") + th

  # ── 2. Linearity ─────────────────────────────────────────────────────────
  lin_dat <- eq_res %>% filter(is.finite(fitted_pop))
  if (nrow(lin_dat) > 25000) lin_dat <- slice_sample(lin_dat, n = 25000)

  p2 <- ggplot(lin_dat, aes(x = fitted_pop, y = residual)) +
    geom_hline(yintercept = 0, linetype = "dashed",
               color = "grey60", linewidth = 0.4) +
    geom_point(size = 0.2, alpha = 0.05, color = col_fit) +
    geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
                color = col_line, fill = col_line,
                alpha = 0.15, linewidth = 0.6) +
    labs(title    = "Linearity",
         subtitle = "Reference line should be flat and horizontal",
         x = "Fitted values", y = "Residuals") + th

  # ── 3. Homogeneity of Variance ───────────────────────────────────────────
  p3 <- ggplot(lin_dat %>% mutate(sqr = sqrt(abs(residual))),
               aes(x = fitted_pop, y = sqr)) +
    geom_point(size = 0.2, alpha = 0.05, color = col_fit) +
    geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
                color = col_line, fill = col_line,
                alpha = 0.15, linewidth = 0.6) +
    labs(title    = "Homogeneity of Variance",
         subtitle = "Reference line should be flat and horizontal",
         x = "Fitted values",
         y = expression(sqrt("|Std. residuals|"))) + th

  # ── 4. Influential Observations ──────────────────────────────────────────
  # Cook's D contour: r = ±sqrt(D * p * (1-h)^2 / h)
  # hatvalues() can fail or return all-NA for single-predictor equations
  # (e.g. microclimate ~ land_use only). In that case the panel falls back
  # to a plain residual vs fitted_pop scatter without Cook's D contours.
  inf_dat <- eq_res %>% filter(is.finite(leverage))
  if (nrow(inf_dat) > 25000) inf_dat <- slice_sample(inf_dat, n = 25000)

  has_leverage <- nrow(inf_dat) > 0 &&
                  is.finite(quantile(inf_dat$leverage, 0.995, na.rm = TRUE))

  if (has_leverage) {
    max_h  <- quantile(inf_dat$leverage, 0.995, na.rm = TRUE)
    h_seq  <- seq(1e-4, max(max_h * 1.05, 1e-3), length.out = 400)
    r_clip <- quantile(abs(inf_dat$residual), 0.999, na.rm = TRUE) * 1.5

    cook_lines <- map_dfr(cook_thresh, function(d) {
      r_val <- sqrt(d * n_fix * (1 - h_seq)^2 / h_seq)
      bind_rows(
        tibble(h = h_seq, r =  r_val, level = as.character(d)),
        tibble(h = h_seq, r = -r_val, level = as.character(d))
      )
    }) %>% filter(is.finite(r), abs(r) < r_clip)

    cook_labels <- cook_lines %>%
      filter(r > 0) %>% group_by(level) %>% slice_max(h, n = 1)

    p4 <- ggplot(inf_dat, aes(x = leverage, y = residual)) +
      geom_hline(yintercept = 0, linetype = "dashed",
                 color = "grey60", linewidth = 0.4) +
      geom_point(size = 0.2, alpha = 0.05, color = col_fit) +
      geom_line(data = cook_lines,
                aes(x = h, y = r, group = interaction(level, r > 0)),
                color = col_line, linewidth = 0.45, linetype = "dashed") +
      geom_text(data = cook_labels,
                aes(x = h, y = r, label = level),
                size = 2.4, color = col_line, hjust = -0.1, vjust = 0) +
      labs(title    = "Influential Observations",
           subtitle = "Points should be inside the contour lines",
           x = "Leverage", y = "Std. Residuals") + th

  } else {
    # Fallback: leverage unavailable (single binary predictor, nested lme).
    # Show Cook's D directly on the y-axis instead.
    cd_dat <- eq_res %>% filter(is.finite(cooks_d), is.finite(fitted_pop))
    if (nrow(cd_dat) > 25000) cd_dat <- slice_sample(cd_dat, n = 25000)

    p4 <- ggplot(cd_dat, aes(x = fitted_pop, y = cooks_d)) +
      geom_hline(yintercept = c(0.5, 1.0), linetype = "dashed",
                 color = col_line, linewidth = 0.35) +
      geom_point(size = 0.2, alpha = 0.06, color = col_fit) +
      annotate("text", x = Inf, y = 0.52, label = "0.5",
               hjust = 1.1, size = 2.4, color = col_line) +
      annotate("text", x = Inf, y = 1.02, label = "1.0",
               hjust = 1.1, size = 2.4, color = col_line) +
      labs(title    = "Influential Observations",
           subtitle = "Leverage unavailable — showing Cook's D vs fitted",
           x = "Fitted values (population level)", y = "Cook's D") + th
  }

  # ── 5. Collinearity ──────────────────────────────────────────────────────
  if (nrow(eq_vif) == 0) {
    p5 <- ggplot() +
      annotate("text", x = 0.5, y = 0.5,
               label = "Single predictor\n(VIF not applicable)",
               size = 3.2, color = "grey50", hjust = 0.5, vjust = 0.5) +
      scale_x_continuous(limits = c(0, 1)) +
      scale_y_continuous(limits = c(0, 1)) +
      labs(title = "Collinearity") +
      th + theme(axis.text  = element_blank(),
                 axis.ticks = element_blank(),
                 axis.title = element_blank())
  } else {
    vif_summ <- eq_vif %>%
      group_by(predictor) %>%
      summarise(mean_vif = mean(vif, na.rm = TRUE),
                sd_vif   = sd(vif,   na.rm = TRUE),
                .groups  = "drop") %>%
      mutate(predictor = fct_reorder(predictor, mean_vif),
             hi        = mean_vif >= 10)

    p5 <- ggplot(vif_summ, aes(x = mean_vif, y = predictor)) +
      annotate("rect", xmin = 1,  xmax = 5,   ymin = -Inf, ymax = Inf,
               fill = "#d4edda", alpha = 0.5) +
      annotate("rect", xmin = 5,  xmax = 10,  ymin = -Inf, ymax = Inf,
               fill = "#fff3cd", alpha = 0.5) +
      annotate("rect", xmin = 10, xmax = Inf, ymin = -Inf, ymax = Inf,
               fill = "#f8d7da", alpha = 0.5) +
      geom_vline(xintercept = c(5, 10), linetype = "dashed",
                 color = "grey50", linewidth = 0.35) +
      geom_errorbarh(
        aes(xmin = pmax(mean_vif - sd_vif, 1), xmax = mean_vif + sd_vif),
        height = 0.3, color = "grey40", linewidth = 0.45
      ) +
      geom_point(aes(color = hi), size = 2.8) +
      scale_color_manual(values = c("FALSE" = "#3d5a3e", "TRUE" = col_line),
                         guide  = "none") +
      scale_x_log10(labels = scales::comma) +
      labs(title    = "Collinearity",
           subtitle = "High VIF (>= 10) may inflate parameter uncertainty",
           x = "VIF (log-scaled, mean +/- SD across windows)", y = NULL) + th
  }

  # ── 6. Normality ─────────────────────────────────────────────────────────
  qq_dat <- eq_res
  if (nrow(qq_dat) > 40000) qq_dat <- slice_sample(qq_dat, n = 40000)

  p6 <- ggplot(qq_dat, aes(sample = residual)) +
    geom_hline(yintercept = c(-2, 2), color = "grey70",
               linewidth = 0.3, linetype = "dashed") +
    stat_qq_line(color = col_line, linewidth = 0.55) +
    stat_qq(size = 0.15, alpha = 0.08, color = col_fit) +
    labs(title    = "Normality of Residuals",
         subtitle = "Dots should fall along the line",
         x        = "Standard Normal Distribution Quantiles",
         y        = "Sample Quantile Deviations") + th

  # ── Assemble ────────────────────────────────────────────────────────────
  (p1 | p2) / (p3 | p4) / (p5 | p6) +
    plot_annotation(
      title    = paste0("Model diagnostics — equation: ", eq_label),
      subtitle = sprintf(
        "Pooled across %d windows x %d taxa  |  n = %s observations",
        n_wins_total, n_taxa_total,
        scales::comma(nrow(eq_res))
      ),
      theme = theme(
        plot.title    = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(color = "grey50", size = 8,
                                     margin = margin(b = 6))
      )
    )
}
```

```{r fig-check-model, fig.width = 18, fig.height = 14, out.width = "100%"}
#| fig-cap: >
#|   Model assumption diagnostics for each of the four SEM equations, built
#|   from all rolling-window model fits pooled together. Figures are also
#|   saved individually as TIFF files for submission.

response_order <- c("microclimate_1", "microclimate_2", "root", "fauna")
response_labels <- c(
  microclimate_1 = microclimate_index[1],
  microclimate_2 = microclimate_index[2],
  root           = "Root growth",
  fauna          = "Faunal activity"
)

check_figs <- imap(response_labels, function(eq_label, eq_name) {
  make_check_panels(
    eq_name    = eq_name,
    eq_label   = eq_label,
    resid_data = all_residuals,
    vif_data   = all_vif
  )
})

# Display all four figures inline
walk(check_figs, print)

# Save each equation as a separate TIFF (GCB double-column 175 mm)
iwalk(check_figs, function(fig, eq_name) {
  ggsave(
    filename    = sprintf("../output/Fig_diagnostics_%s.tiff", eq_name),
    plot        = fig,
    width       = 17.5, height = 13.5, units = "cm",
    dpi         = 300,  compression = "lzw"
  )
})
message("Diagnostic figures saved for: ", paste(names(check_figs), collapse = ", "))
```

```{r}
#| label: tbl-residual-summary
#| tbl-cap: >
#|   Summary statistics for normalised residuals pooled across all
#|   rolling-window models and taxa. Skewness and kurtosis are computed on
#|   the full pooled distribution. Prop. |z| > 2: proportion of residuals
#|   with absolute value exceeding 2 (expected ≈ 5% under normality).
#|   Values coloured in orange exceed the skewness threshold (|skew| > 1.5).

# Pre-compute stats for clean column_spec colouring (avoids double evaluation)
resid_stats <- all_residuals %>%
  filter(response %in% response_order) %>%
  mutate(response = factor(response,
                           levels = response_order,
                           labels = response_labels[response_order])) %>%
  group_by(response) %>%
  summarise(
    n_residuals = n(),
    mean        = round(mean(residual,                    na.rm = TRUE), 3),
    sd          = round(sd(residual,                      na.rm = TRUE), 3),
    skewness    = round(moments::skewness(residual,       na.rm = TRUE), 3),
    kurtosis    = round(moments::kurtosis(residual,       na.rm = TRUE), 3),
    prop_out2sd = round(mean(abs(residual) > 2,           na.rm = TRUE), 3),
    .groups = "drop"
  )

resid_stats %>%
  kable(
    col.names = c("Response", "N residuals", "Mean", "SD",
                  "Skewness", "Kurtosis", "Prop. |z| > 2"),
    caption   = "Residual distribution summary by equation (pooled across all windows and taxa)"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE, position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  column_spec(5,
    color = if_else(abs(resid_stats$skewness) > 1.5, "#c1603a", "black"),
    bold  = if_else(abs(resid_stats$skewness) > 1.5, TRUE,       FALSE)
  ) %>%
  column_spec(6,
    color = if_else(resid_stats$kurtosis > 10, "#c1603a", "black"),
    bold  = if_else(resid_stats$kurtosis > 10, TRUE,      FALSE)
  )
```

------------------------------------------------------------------------

## 7. Consolidation and export

```{r consolidation}

# ── Filter out correlated error terms (~~ rows) ───────────────────────────────
# piecewiseSEM includes %~~% entries in the coefficient table; they are not
# directed paths and are excluded from the results database.
sem_results <- sem_results_raw %>%
  filter(!str_detect(predictor, "~~")) %>%
  mutate(path_id = paste(predictor, "\u2192", response))

# ── Placeholder columns for downstream compatibility ─────────────────────────
# The additive model has no interaction terms, so std_estimate_C equals
# std_estimate and has_interaction is FALSE throughout. These columns are
# retained to maintain a consistent schema with the visualisation script,
# which expects them for the A / C contrast display.
sem_results <- sem_results %>%
  mutate(
    std_estimate_C      = std_estimate,
    interaction_p_value = NA_real_,
    has_interaction     = FALSE
  )

# ── Season labels ─────────────────────────────────────────────────────────────
season_lookup <- tibble(
  month  = 1:12,
  season = c("Winter", "Winter",
             "Spring", "Spring", "Spring",
             "Summer", "Summer", "Summer",
             "Autumn", "Autumn", "Autumn",
             "Winter")
)

sem_results <- sem_results %>%
  mutate(month = month(date_start)) %>%
  left_join(season_lookup, by = "month")

# ── Export ────────────────────────────────────────────────────────────────────
write.csv(sem_results,    "../output/SEM_results_database.csv",    row.names = FALSE)
write.csv(all_residuals,  "../output/SEM_residuals_database.csv",  row.names = FALSE)

message(sprintf(
  "Export complete: %d coefficient rows | %d taxa | %d paths",
  nrow(sem_results),
  n_distinct(sem_results$taxon),
  n_distinct(sem_results$path_id)
))
```

------------------------------------------------------------------------

## 8. Quick sanity check

```{r sanity-check-table}
sem_results %>%
  filter(response %in% c("root", "fauna")) %>%
  group_by(taxon, path_id) %>%
  summarise(
    n_windows    = n(),
    mean_beta    = round(mean(std_estimate, na.rm = TRUE), 3),
    sd_beta      = round(sd(std_estimate,  na.rm = TRUE), 3),
    prop_sig     = round(mean(p_value < 0.05, na.rm = TRUE), 2),
    mean_r2_marg = round(mean(r2_marg, na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(taxon, path_id) %>%
  knitr::kable(
    caption = "Path summary: mean standardised coefficient across all windows",
    col.names = c("Taxon", "Path", "N windows",
                  "Mean β", "SD β", "Prop. sig.", "Mean R² marg.")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")
```
