---
title: "SEM — model execution and path coefficient extraction"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: true
editor:
  markdown:
    wrap: 72
---

# Overview

## Role in the pipeline

This script is the **third step** of the SEM pipeline, following:

1.  `data_preparation.qmd` — builds the full SEM database
2.  `sem_diagnostic.qmd` — selects the optimal window, completeness
    threshold, and transformations

## What this script produces

| Output | File | Description |
|----|----|----|
| Observed path coefficients | `SEM_results_database.csv` | Standardised estimates, p-values, R², and land-use-conditional effects for all taxon × window combinations |

## Statistical approach

The causal model has four structural equations organised in three tiers,
with `land_use` (0 = unmanaged / A, 1 = managed / C) as the exogenous
binary driver. **Tiers 2 and 3 include interaction terms** between
`land_use` and all continuous predictors, so the strength of each causal
path is estimated separately for each management context.

**Tier 1 — land use shapes microclimate** (additive; `land_use` is the
sole driver at this level):

$$\text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$
$$\text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$

**Tier 2 — microclimate × land use drives root** (interactions test
whether microclimate sensitivity of root differs by management):

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu} + \beta_6 \cdot (\text{mc}_1 \times \text{lu}) + \beta_7 \cdot (\text{mc}_2 \times \text{lu})$$

**Tier 3 — (microclimate + root) × land use drives fauna** (the critical
interaction $\beta_{14}$ tests whether root drives fauna differently in
managed vs unmanaged soils):

$$\text{Fauna} \leftarrow \beta_8 \cdot \text{mc}_1 + \beta_9 \cdot \text{mc}_2 + \beta_{10} \cdot \text{Root} + \beta_{11} \cdot \text{lu} + \beta_{12} \cdot (\text{mc}_1 \times \text{lu}) + \beta_{13} \cdot (\text{mc}_2 \times \text{lu}) + \beta_{14} \cdot (\text{Root} \times \text{lu})$$

*where mc = microclimate index, lu = land_use.*

### Reading the interaction coefficients

Because `land_use` is binary (0 = A, 1 = C) and is **not** z-scored, the
coefficients have a direct, asymmetric interpretation:

| Coefficient | Meaning |
|----|----|
| `root` (β₁₀) | Root → fauna slope **in the unmanaged system** (baseline, land_use = 0) |
| `root:land_use` (β₁₄) | **Difference** in slope between managed and unmanaged |
| β₁₀ + β₁₄ | Root → fauna slope **in the managed system** |
| `land_use` (β₁₁) | Direct management effect on fauna, independent of microclimate and root |

A significant positive β₁₄ means root is a **stronger** driver of fauna
in the managed system; a negative β₁₄ means it is **weaker**. The same
logic applies to all microclimate interaction terms.

The residual covariance `microclimate_1 %~~% microclimate_2` is modelled
explicitly because both indices share unmeasured physical drivers
(radiation, topography) beyond the binary land-use contrast. Without
this term, piecewiseSEM issues missing-path claims and inflates
d-separation test statistics.

Each equation uses a **nested random intercept** (`orientation / depth`)
and an **AR(1) correlation structure** to account for temporal
autocorrelation within scanner groups. The optimizer is set to `optim`
with 200 maximum iterations to improve convergence stability.

**Transformations** are applied differentially before z-score
standardisation:

-   `root` — transformation selected during sensitivity analysis
    (`TRANSFO_ROOT`)
-   `fauna` variables — transformation from sensitivity analysis
    (`TRANSFO_FAUNA`)
-   `microclimate_1` / `microclimate_2` — — transformation from
    sensitivity analysis (`TRANSFO_MICROCLIMATE_1` /
    `TRANSFO_MICROCLIMATE_2`)
-   `land_use` — excluded from z-scoring to preserve the binary 0/1
    contrast and ensure interaction coefficients remain interpretable as
    slope differences

------------------------------------------------------------------------

## 1. Setup

```{r setup}
#| message: false

# ── Locale ────────────────────────────────────────────────────────────────────
invisible(Sys.setlocale("LC_TIME", "C"))

# ── Libraries ─────────────────────────────────────────────────────────────────
libs <- c(
  "tidyverse",    # Data wrangling and visualisation
  "lubridate",    # Date/time manipulation
  "piecewiseSEM", # Piecewise structural equation modelling
  "performance",  # Model performance metrics
  "lme4",         # Linear mixed-effects models
  "purrr",        # Functional iteration
  "nlme",         # LME with correlation structures (corAR1)
  "moments",      # Skewness and kurtosis
  "car",          # VIF
  "jsonlite",     # Reading the JSON parameter file
  "knitr"         # Table rendering
)
invisible(lapply(libs, require, character.only = TRUE))
```

------------------------------------------------------------------------

## 2. Parameter loading

Parameters were determined in `sem_diagnostic.qmd` and saved to
`model_parameters.txt`. If the file is missing, safe fallback defaults
are applied with a warning.

```{r load-params}
if (file.exists("../output/model_parameters.txt")) {

  best_params            <- read_json("../output/model_parameters.txt")
  WIN_DAYS               <- as.numeric(best_params$selected_window)
  MIN_RATE               <- as.numeric(best_params$selected_min_rate)
  TRANSFO_ROOT           <- as.character(best_params$transform_root)
  TRANSFO_FAUNA          <- as.character(best_params$transform_fauna)
  TRANSFO_MICROCLIMATE_1 <- as.character(best_params$transform_microclimate_1)
  TRANSFO_MICROCLIMATE_2 <- as.character(best_params$transform_microclimate_2)

} else {
  WIN_DAYS               <- 14
  MIN_RATE               <- 0.3
  TRANSFO_ROOT           <- "sqrt_plus_one"
  TRANSFO_FAUNA          <- "sqrt_plus_one"
  TRANSFO_MICROCLIMATE_1 <- "none"
  TRANSFO_MICROCLIMATE_2 <- "none"
  warning("Configuration file not found. Using default fallback parameters.")
}

message(sprintf(
  paste0("Active config — Window: %d days | Min. rate: %.1f\n",
         "  Transfo root: %s | Transfo fauna: %s\n",
         "  Transfo microclimate_1: %s | Transfo microclimate_2: %s"),
  WIN_DAYS, MIN_RATE,
  TRANSFO_ROOT, TRANSFO_FAUNA,
  TRANSFO_MICROCLIMATE_1, TRANSFO_MICROCLIMATE_2
))
```

------------------------------------------------------------------------

## 3. Data loading and preprocessing

```{r data-preparation}

# ── Load raw database ─────────────────────────────────────────────────────────
data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep        = ",",
                         na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    # Binary land-use contrast: 0 = unmanaged (A), 1 = managed (C)
    # Kept as numeric and NOT z-scored so that:
    #   (a) main effect coefficients are interpretable as A → C contrasts
    #   (b) interaction coefficients represent the *difference* in slope
    #       between management types on the original β scale
    land_use = as.numeric(position == "C"),
    date     = ymd_hms(date, tz = "UTC"),
    root     = root_growth_cm2_count_24h   # 24-hour root growth flux (cm²)
  )

# ── Load metadata ─────────────────────────────────────────────────────────────
fauna_vars         <- c("all_taxa", readLines("../output/fauna_vars.txt"))
microclimate_index <- readLines("../output/microclimate_index.txt")

# ── Transformation helper ─────────────────────────────────────────────────────
#' Apply a named variance-stabilising transformation to a numeric vector
#'
#' @param x       Numeric vector (counts, densities, or continuous indices).
#' @param method  Character. One of: "none","sqrt", "sqrt_plus_one", "log", "hellinger".
#' @return Transformed numeric vector of the same length.
apply_transfo <- function(x, method) {
  switch(method,
    "sqrt_plus_one" = sqrt(x) + sqrt(x + 1),          # Freeman-Tukey (counts)
    "sqrt"          = sqrt(x), 
    "log"           = log1p(x),                         # log(x + 1)
    "hellinger"     = sqrt(x / sum(x, na.rm = TRUE)),  # window-relative scale
    x                                                   # "none": identity
  )
}

# ── Apply differential transformations ────────────────────────────────────────
data_transformed <- data_sem_raw %>%

  # Root: biological flux — all four transforms are valid candidates
  mutate(root = apply_transfo(root, TRANSFO_ROOT)) %>%

  # Fauna variables: same rationale as root (count data)
  mutate(across(
    all_of(intersect(fauna_vars, names(.))),
    ~ apply_transfo(., TRANSFO_FAUNA)
  )) %>%

  # Microclimate indices: restricted to log or none
  # Count-specific transforms are not meaningful for continuous environmental
  # variables, so they are silently downgraded to "none" if selected
  mutate(across(
    all_of(microclimate_index[1]),
    ~ apply_transfo(.,  TRANSFO_MICROCLIMATE_1)
  )) %>%
  mutate(across(
    all_of(microclimate_index[2]),
    ~ apply_transfo(., TRANSFO_MICROCLIMATE_2)
  )) %>%

  # Z-score standardisation: all numeric variables EXCEPT land_use
  # Excluding land_use ensures interaction coefficients represent actual
  # slope differences between management types, not scaled contrasts
  mutate(across(
    where(is.numeric) & !all_of("land_use"),
    ~ as.numeric(scale(.))
  ))
```

------------------------------------------------------------------------

## 4. Analytical functions

### 4.1 Rolling window splitter

```{r rolling-window-generator}
#' Split a time series into overlapping rolling windows
#'
#' @param df          Dataframe with a `date` column (POSIXct).
#' @param window_days Integer. Window duration in days.
#' @param step_days   Integer. Step between consecutive window start dates.
#' @return Named list of dataframes, one per window start date.
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )

  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  return(windows)
}
```

### 4.2 SEM structure

Tiers 2 and 3 use `* land_use` syntax, which R expands into the main
effects plus all pairwise interactions with `land_use`. This allows the
slope of every continuous predictor to differ between management
contexts. Tier 1 remains additive because `land_use` is the sole
predictor — an interaction with itself is undefined.

The residual covariance `microclimate_1 %~~% microclimate_2` accounts
for shared physical drivers (radiation, topography) not captured by the
binary land-use contrast.

```{r SEM-structure}
#' Fit the piecewise SEM with interaction terms for a single data window
#'
#' @param data Dataframe with columns: `land_use`, `root`, `fauna`,
#'             `microclimate_1`, `microclimate_2`, `orientation`, `depth`.
#' @return A `psem` object.
fit_psem <- function(data) {
  psem(

    # ── Tier 1: land use shapes microclimate (additive) ───────────────────────
    lme(microclimate_1 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    lme(microclimate_2 ~ land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # ── Tier 2: (microclimate) × land use drives root ─────────────────────────
    # Interactions test whether microclimate sensitivity of root differs by
    # management type
    lme(root ~ (microclimate_1 + microclimate_2) * land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # ── Tier 3: (microclimate + root) × land use drives fauna ────────────────
    # Key interaction: root:land_use coefficient = difference in root → fauna
    # slope between managed (C) and unmanaged (A) systems
    lme(fauna ~ (microclimate_1 + microclimate_2 + root) * land_use,
        data        = data,
        random      = ~ 1 | orientation / depth,
        correlation = corAR1(form = ~ 1 | orientation / depth),
        control     = lmeControl(msMaxIter = 200, opt = "optim")),

    # Residual covariance between microclimate indices (shared exogenous drivers)
    microclimate_1 %~~% microclimate_2
  )
}
```

### 4.3 Main pipeline

```{r fixed-pipeline}
#' Run the full rolling-window SEM pipeline for one taxon variable
#'
#' Both management positions (A = 0, C = 1) are modelled jointly via
#' `land_use` and its interactions with continuous predictors.
#' Windows are checked for data sufficiency, variance, and land_use level
#' representativeness before fitting.
#'
#' @param taxon_var   Character. Column name of the fauna variable to model.
#' @param data_raw    Dataframe. Transformed and z-score standardised SEM database.
#' @param m_indices   Character vector of length 2. Microclimate index column names.
#' @param window_days Integer. Window duration (days).
#' @param min_rate    Numeric [0, 1]. Minimum data completeness ratio.
#'                    Denominator = window_days × 16 slots/day × 2 positions.
#' @return Dataframe with one row per path coefficient (including interactions),
#'         or NULL if no valid windows.
run_sem_pipeline <- function(taxon_var, data_raw, m_indices,
                             window_days, min_rate) {

  cat("\n--- Processing taxon:", taxon_var, "---\n")

  # Rename dynamic columns to fixed internal SEM names
  sem_data <- data_raw %>%
    mutate(
      fauna          = !!sym(taxon_var),
      microclimate_1 = !!sym(m_indices[1]),
      microclimate_2 = !!sym(m_indices[2]),
      depth          = as.factor(depth)
    ) %>%
    select(date, land_use, orientation, depth,
           fauna, root, microclimate_1, microclimate_2)

  windows <- create_rolling_windows(sem_data, window_days, step_days = 1)
  if (length(windows) == 0) return(NULL)

  # Filter windows by data completeness
  # Max possible rows: window_days × 16 reading slots × 2 land-use positions
  valid_wins <- keep(windows,
                     ~ (nrow(drop_na(.x)) / (window_days * 16 * 2)) >= min_rate)

  cat(sprintf("  %d / %d windows passed the completeness filter.\n",
              length(valid_wins), length(windows)))

  if (length(valid_wins) == 0) return(NULL)

  results_list <- map(names(valid_wins), function(w_name) {

    win_data <- valid_wins[[w_name]] %>%
      drop_na() %>%
      # corAR1 requires observations sorted in time order within groups
      arrange(land_use, orientation, depth, date)

    # Guard 1: skip windows where any modelled variable has zero variance
    # (causes singular Cholesky decomposition in lme and summary.psem)
    vars_to_check <- c("fauna", "root", "microclimate_1", "microclimate_2")
    if (any(sapply(win_data[vars_to_check], sd, na.rm = TRUE) == 0)) return(NULL)

    # Guard 2: skip windows where land_use has only one level
    # (interactions cannot be estimated without both positions present)
    if (length(unique(win_data$land_use)) < 2) return(NULL)

    # Guard 3: skip windows with non-finite values (from transforms / scaling)
    if (any(!is.finite(as.matrix(win_data[vars_to_check])))) return(NULL)

    # Fit SEM — mute optimisation warnings; catch hard convergence errors
    model <- tryCatch(
      withCallingHandlers(
        fit_psem(win_data),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
    if (is.null(model)) return(NULL)

    # summary.psem() can itself fail on singular fits — wrap separately
    sm <- tryCatch(summary(model), error = function(e) NULL)
    if (is.null(sm)) return(NULL)

    # ── Path coefficients ────────────────────────────────────────────────────
    coefs <- as.data.frame(sm$coefficients)

    # ── R² (marginal = fixed effects; conditional = fixed + random) ──────────
    # Column names vary across piecewiseSEM versions — detect dynamically
    r2_raw <- as.data.frame(sm$R2)
    r2_df  <- r2_raw %>%
      mutate(
        r2_cond = .[[grep("R.squared|Conditional", names(.))[1]]],
        r2_marg = .[[grep("Marginal",              names(.))[1]]]
      ) %>%
      select(response = Response, r2_cond, r2_marg)

    tibble(
      taxon        = taxon_var,
      date_start   = as.Date(w_name),
      n_obs        = sm$n,
      response     = coefs$Response,
      predictor    = coefs$Predictor,
      estimate     = coefs$Estimate,
      std_estimate = coefs$Std.Estimate,
      # piecewiseSEM returns "<0.001" as a string — strip the "<" for numeric use
      p_value      = as.numeric(str_replace(coefs$P.Value, "<", ""))
    ) %>%
      left_join(r2_df, by = "response")
  }) %>%
    compact()

  if (length(results_list) == 0) return(NULL)
  return(map_dfr(results_list, ~ .x))
}
```

------------------------------------------------------------------------

## 5. Model execution

```{r observed-models}
message(">>> Running observed models across all taxa...")

sem_results_raw <- map_dfr(
  fauna_vars,
  ~ run_sem_pipeline(.x, data_transformed, microclimate_index, WIN_DAYS, MIN_RATE)
)

message(sprintf("Observed models complete: %d rows across %d taxa.",
                nrow(sem_results_raw), n_distinct(sem_results_raw$taxon)))
```

------------------------------------------------------------------------

## 6. Consolidation and export

```{r consolidation}

# ── Classify paths ────────────────────────────────────────────────────────────
# piecewiseSEM labels interaction predictors as "predictor:land_use".
# We tag each row with its path type for downstream filtering:
#   - "main"        : standard directed path (e.g. root → fauna)
#   - "interaction" : land-use-modulated path (e.g. root:land_use → fauna)
# Correlated error terms (~~) are dropped entirely.

sem_results <- sem_results_raw %>%
  filter(!str_detect(predictor, "~~")) %>%
  mutate(
    path_type = if_else(str_detect(predictor, ":land_use"),
                        "interaction", "main"),
    path_id   = paste(predictor, "→", response)
  )

# ── Derive land-use-conditional slopes ────────────────────────────────────────
# For each main path X → Y that has a corresponding interaction X:land_use → Y:
#   Effect in unmanaged (A, land_use = 0) = main effect coefficient (std_estimate)
#   Effect in managed   (C, land_use = 1) = main + interaction coefficient
#
# This is computed here so that both conditional slopes are available as
# first-class columns in the results database, ready for direct A vs C
# comparison in the visualisation script.

interaction_coefs <- sem_results %>%
  filter(path_type == "interaction") %>%
  mutate(base_predictor = str_remove(predictor, ":land_use")) %>%
  select(taxon, date_start, response,
         base_predictor,
         interaction_std_estimate = std_estimate,
         interaction_p_value      = p_value)

sem_results <- sem_results %>%
  filter(path_type == "main") %>%
  left_join(interaction_coefs,
            by = c("taxon", "date_start", "response",
                   "predictor" = "base_predictor")) %>%
  mutate(
    # std_estimate     = slope in unmanaged system (A, land_use = 0)
    # std_estimate_C   = slope in managed system   (C, land_use = 1)
    # NA when no interaction exists for this path (e.g. Tier 1 paths)
    std_estimate_C  = std_estimate + replace_na(interaction_std_estimate, 0),
    has_interaction = !is.na(interaction_std_estimate)
  )

# ── Season labels ─────────────────────────────────────────────────────────────
season_lookup <- tibble(
  month  = 1:12,
  season = c("Winter", "Winter",
             "Spring", "Spring", "Spring",
             "Summer", "Summer", "Summer",
             "Autumn", "Autumn", "Autumn",
             "Winter")
)

sem_results <- sem_results %>%
  mutate(month = month(date_start)) %>%
  left_join(season_lookup, by = "month")

# ── Export ────────────────────────────────────────────────────────────────────
write.csv(sem_results, "../output/SEM_results_database.csv", row.names = FALSE)

message(sprintf(
  "Export complete: %d rows | %d taxa | %d paths (%d with land-use interaction)",
  nrow(sem_results),
  n_distinct(sem_results$taxon),
  n_distinct(sem_results$path_id),
  sum(sem_results$has_interaction, na.rm = TRUE)
))
```

------------------------------------------------------------------------

## 7. Quick sanity check

```{r sanity-check-table}
# For each path, report the mean slope in each management context (A and C),
# the mean difference (Δ = C − A = mean interaction coefficient), and the
# proportion of windows where the interaction is significant.
# This is the primary check that the interaction structure produced
# meaningful, interpretable output.

sem_results %>%
  filter(response %in% c("root", "fauna")) %>%
  group_by(taxon, path_id, has_interaction) %>%
  summarise(
    n_windows    = n(),
    # Slope in unmanaged context (land_use = 0, baseline)
    mean_beta_A  = round(mean(std_estimate,   na.rm = TRUE), 3),
    # Slope in managed context (land_use = 1)
    mean_beta_C  = round(mean(std_estimate_C, na.rm = TRUE), 3),
    # Mean difference: positive = stronger effect in managed system
    mean_delta   = round(mean(std_estimate_C - std_estimate, na.rm = TRUE), 3),
    prop_sig_A   = round(mean(p_value < 0.05,               na.rm = TRUE), 2),
    prop_sig_int = round(mean(interaction_p_value < 0.05,   na.rm = TRUE), 2),
    mean_r2_marg = round(mean(r2_marg,                      na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  arrange(taxon, path_id) %>%
  knitr::kable(
    caption = "Path summary: slopes in unmanaged (A) vs managed (C) contexts",
    col.names = c("Taxon", "Path", "Has interaction",
                  "N windows",
                  "Mean β (A)", "Mean β (C)", "Mean Δ (C−A)",
                  "Prop. sig. main", "Prop. sig. interaction",
                  "Mean R² marg.")
  )
```
