---
title: "From pixels to patterns — results and visualisation"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: false
---

# Overview

## Role in the pipeline

This script is the **final analytical step** of the SEM pipeline. It loads the path coefficient database produced by `3_SEM_modelisation.qmd` and addresses the three research questions through figures and tables.

## SEM structure (additive, no interactions)

**Tier 1 — land use shapes microclimate:**

$$\text{microclimate}_1 \leftarrow \beta_1 \cdot \text{land\_use}$$ $$\text{microclimate}_2 \leftarrow \beta_2 \cdot \text{land\_use}$$

**Tier 2 — microclimate drives root:**

$$\text{Root} \leftarrow \beta_3 \cdot \text{mc}_1 + \beta_4 \cdot \text{mc}_2 + \beta_5 \cdot \text{lu}$$

**Tier 3 — microclimate and root drive fauna:**

$$\text{Fauna} \leftarrow \beta_6 \cdot \text{mc}_1 + \beta_7 \cdot \text{mc}_2 + \beta_8 \cdot \text{Root} + \beta_9 \cdot \text{lu}$$

## How indirect effects are computed

In an additive SEM, the total effect of `land_use` on `fauna` decomposes into one direct path and five indirect paths through the causal chain. Each indirect effect is the **product** of the standardised coefficients along the corresponding route. These products are computed per rolling window, yielding a distribution of effect sizes over time.

| Path                    | Expression                              |
|-------------------------|-----------------------------------------|
| Direct                  | $\beta_9$                               |
| lu → mc₁ → fauna        | $\beta_1 \times \beta_6$                |
| lu → mc₂ → fauna        | $\beta_2 \times \beta_7$                |
| lu → root → fauna       | $\beta_5 \times \beta_8$                |
| lu → mc₁ → root → fauna | $\beta_1 \times \beta_3 \times \beta_8$ |
| lu → mc₂ → root → fauna | $\beta_2 \times \beta_4 \times \beta_8$ |
| **Total indirect**      | sum of the five above                   |
| **Total effect**        | direct + total indirect                 |

------------------------------------------------------------------------

```{r setup}
#| label: setup

invisible(Sys.setlocale("LC_TIME", "C"))
dir.create("../output", showWarnings = FALSE)

libs <- c(
  "tidyverse", "lubridate", "patchwork",
  "kableExtra", "nlme", "zoo", "ggdist", "slider", "ggside"
)
invisible(lapply(libs, require, character.only = TRUE))

microclimate_index <- readLines("../output/microclimate_index.txt")
fauna_vars         <- readLines("../output/fauna_vars.txt")

# ══════════════════════════════════════════════════════════════════════════════
# VISUAL IDENTITY
# ══════════════════════════════════════════════════════════════════════════════

SEM_COLORS <- list(
  path = c(
    "#c2cab2",  # mc1 → root
    "#555d45",  # mc2 → root
    "#dea88b",  # mc1 → fauna
    "#d1855c",  # mc2 → fauna
    "#955f42"   # root → fauna
  ),
  response  = c(root = "#94a378", fauna = "#d1855c"),
  position  = c(A = "#3d5a3e", C = "#c1603a"),
  sync_low  = "#B2182B",
  sync_mid  = "#F7F7F7",
  sync_high = "#2166AC"
)

response_palette  <- SEM_COLORS$response
season_order      <- c("Spring", "Summer", "Autumn", "Winter")
season_fills      <- c(Spring = "#F2F2F2", Summer = "#E4E4E4",
                       Autumn = "#F2F2F2", Winter = "#E4E4E4")

# Path ordering: abiotic → biotic
path_levels <- tibble(
  path_order = 1:5,
  path_id = c(
    paste0(microclimate_index[1], " \u2192 root"),
    paste0(microclimate_index[2], " \u2192 root"),
    paste0(microclimate_index[1], " \u2192 fauna"),
    paste0(microclimate_index[2], " \u2192 fauna"),
    "root \u2192 fauna"
  )
)
path_palette <- setNames(SEM_COLORS$path, path_levels$path_id)

# Indirect effect labels and colour palette
indirect_labels <- c(
  "direct"           = paste0("Direct (lu \u2192 fauna)"),
  "via_mc1"          = paste0("lu \u2192 ", microclimate_index[1], " \u2192 fauna"),
  "via_mc2"          = paste0("lu \u2192 ", microclimate_index[2], " \u2192 fauna"),
  "via_root"         = "lu \u2192 root \u2192 fauna",
  "via_mc1_root"     = paste0("lu \u2192 ", microclimate_index[1], " \u2192 root \u2192 fauna"),
  "via_mc2_root"     = paste0("lu \u2192 ", microclimate_index[2], " \u2192 root \u2192 fauna"),
  "total_indirect"   = "Total indirect",
  "total_effect"     = "Total effect"
)
indirect_palette <- c(
  "direct"         = "#3d5a3e",
  "via_mc1"        = "#c2cab2",
  "via_mc2"        = "#555d45",
  "via_root"       = "#94a378",
  "via_mc1_root"   = "#dea88b",
  "via_mc2_root"   = "#d1855c",
  "total_indirect" = "#955f42",
  "total_effect"   = "#1a1a2e"
)

theme_sem <- function(base_size = 8, legend_pos = "none") {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.border      = element_rect(color = "grey20", fill = NA,
                                       linewidth = 0.3),
      panel.grid.major  = element_line(color = "grey94", linewidth = 0.25),
      panel.grid.minor  = element_blank(),
      panel.background  = element_rect(fill = "white", color = NA),
      axis.line         = element_blank(),
      axis.ticks        = element_line(color = "grey40", linewidth = 0.3),
      axis.ticks.length = unit(2, "pt"),
      axis.title        = element_text(size = base_size - 0.5, color = "grey20"),
      axis.text         = element_text(size = base_size - 1.5, color = "grey30"),
      strip.text        = element_text(size = base_size - 1, face = "bold",
                                       color = "grey15"),
      strip.background  = element_rect(fill = "grey96", color = "grey96",
                                       linewidth = 0.3),
      plot.title        = element_text(size = base_size + 0.5, face = "bold",
                                       hjust = 0, color = "grey10",
                                       margin = margin(b = 3)),
      plot.subtitle     = element_text(size = base_size - 1, color = "grey45",
                                       hjust = 0, margin = margin(b = 5)),
      plot.caption      = element_text(size = base_size - 2, color = "grey55",
                                       hjust = 0, margin = margin(t = 4)),
      legend.position   = legend_pos,
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.title      = element_text(size = base_size - 1, face = "bold",
                                       color = "grey20"),
      legend.text       = element_text(size = base_size - 1.5, color = "grey30"),
      legend.key.size   = unit(8, "pt"),
      legend.spacing.x  = unit(3, "pt"),
      plot.margin       = margin(6, 6, 6, 6)
    )
}

make_season_layers <- function(date_range, y_label_pos = Inf, label_size = 2.2) {
  study_years <- seq(year(date_range[1]), year(date_range[2]))
  bands <- tibble(
    season  = names(season_fills),
    m_start = c(3L, 6L, 9L, 12L),
    m_end   = c(6L, 9L, 12L,  3L),
    fill    = unname(season_fills)
  ) %>%
    crossing(year = study_years) %>%
    mutate(
      xmin = as.Date(paste(year, m_start, "01", sep = "-")),
      xmax = as.Date(paste(if_else(m_end < m_start, year + 1L, year),
                           m_end, "01", sep = "-")),
      xmid = xmin + (xmax - xmin) / 2
    ) %>%
    filter(xmin < date_range[2], xmax > date_range[1]) %>%
    mutate(xmin = pmax(xmin, date_range[1]),
           xmax = pmin(xmax, date_range[2]))

  rect_layers <- pmap(bands %>% select(xmin, xmax, fill),
    function(xmin, xmax, fill)
      annotate("rect", xmin = xmin, xmax = xmax,
               ymin = -Inf, ymax = Inf, fill = fill, alpha = 1))

  label_layer <- geom_text(
    data = bands, aes(x = xmid, y = y_label_pos, label = season),
    inherit.aes = FALSE, vjust = 1.6, size = label_size,
    color = "grey60", fontface = "italic", family = "sans")

  c(rect_layers, list(label_layer))
}

# ── Load databases ────────────────────────────────────────────────────────────
data_sem_results <- read.csv("../output/SEM_results_database.csv",
                             na.strings = c("", "NA")) %>%
  mutate(
    taxon      = if_else(taxon == "fauna", "null model", taxon),
    date_start = as.Date(date_start),
    season     = factor(season, levels = season_order)
  ) %>%
  left_join(path_levels, by = "path_id") %>%
  mutate(
    predictor = case_when(
      predictor == "microclimate_1" ~ microclimate_index[1],
      predictor == "microclimate_2" ~ microclimate_index[2],
      TRUE ~ predictor
    ),
    path_id = str_replace(path_id, "microclimate_1", microclimate_index[1]),
    path_id = str_replace(path_id, "microclimate_2", microclimate_index[2]),
    path_id = factor(path_id, levels = path_levels$path_id)
  )

data_sem_raw <- read.csv("../output/SEM_database.csv",
                         sep = ",", na.strings = c("", "NA")) %>%
  mutate(
    position = as.factor(position),
    depth    = as.factor(depth),
    land_use = as.numeric(position == "C"),
    date     = as.Date(ymd_hms(date, tz = "UTC")),
    root     = root_growth_cm2_count_24h,
    month    = month(date),
    season   = factor(case_when(
      month %in% c(3,  4,  5) ~ "Spring",
      month %in% c(6,  7,  8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Autumn",
      TRUE                    ~ "Winter"
    ), levels = season_order)
  )

# ── Outlier-filtered SEM results for distribution plots ───────────────────────
Q1_val  <- quantile(data_sem_results$std_estimate, 0.25, na.rm = TRUE)
Q3_val  <- quantile(data_sem_results$std_estimate, 0.75, na.rm = TRUE)
IQR_val <- Q3_val - Q1_val

data_plot <- data_sem_results %>%
  filter(
    std_estimate >= Q1_val - 1.5 * IQR_val,
    std_estimate <= Q3_val + 1.5 * IQR_val,
    !is.na(path_id)
  )
```

------------------------------------------------------------------------

## 1. H1 — Direct effects of land use on edaphic conditions and faunal activity

> **How does land-use management directly reshape each tier of the soil system?**

Land-use effects on each response variable are estimated here from **independent rolling-window linear mixed models** of the form `response ~ land_use`, with the same random structure as the full SEM (nested random intercept `orientation / depth`, AR(1) correlation). Fitting these models outside the SEM avoids any mediation by the covariate terms in Tiers 2 and 3 — the coefficient here is the total observed contrast between A and C for each variable in isolation, not the SEM partial effect. The seasonal table then shows whether this contrast is stable year-round or season-dependent.

```{r h1-rolling-lme}
#| label: h1-rolling-lme

# ── Per-window rolling window helper ─────────────────────────────────────────
create_rolling_windows <- function(df, window_days, step_days = 1) {
  start_dates <- seq(
    min(df$date), max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )
  windows        <- map(start_dates, ~ filter(df, date >= .x & date < .x + days(window_days)))
  names(windows) <- format(start_dates, "%Y-%m-%d")
  windows
}

# ── Parameters from the diagnostic script ─────────────────────────────────────
WIN_DAYS <- tryCatch({
  p <- jsonlite::read_json("../output/model_parameters.txt")
  as.numeric(p$selected_window)
}, error = function(e) 14L)

MIN_RATE <- tryCatch({
  p <- jsonlite::read_json("../output/model_parameters.txt")
  as.numeric(p$selected_min_rate)
}, error = function(e) 0.3)

# ── Fit independent response ~ land_use per window ───────────────────────────
# One model per response variable; land_use is the sole fixed effect.
# Transformation and z-scoring follow the same per-window logic as the main SEM
# so coefficients are on the same standardised scale as the SEM path estimates.
fit_h1_windows <- function(response_var, data_raw, win_days, min_rate) {

  df <- data_raw %>%
    mutate(y = !!sym(response_var)) %>%
    select(date, land_use, orientation, depth, season, y) %>%
    drop_na()

  windows    <- create_rolling_windows(df, win_days)
  valid_wins <- keep(windows,
                     ~ (nrow(drop_na(.x)) / (win_days * 16 * 2)) >= min_rate)

  map_dfr(names(valid_wins), function(w_name) {
    wd <- valid_wins[[w_name]] %>%
      drop_na() %>%
      arrange(land_use, orientation, depth, date) %>%
      mutate(y = as.numeric(scale(y)))

    if (sd(wd$y, na.rm = TRUE) == 0) return(NULL)
    if (length(unique(wd$land_use)) < 2) return(NULL)

    m <- tryCatch(
      withCallingHandlers(
        nlme::lme(y ~ land_use,
                  data        = wd,
                  random      = ~ 1 | orientation / depth,
                  correlation = nlme::corAR1(form = ~ 1 | orientation / depth),
                  control     = nlme::lmeControl(msMaxIter = 200, opt = "optim")),
        warning = function(w) invokeRestart("muffleWarning")
      ),
      error = function(e) NULL
    )
    if (is.null(m)) return(NULL)

    sm  <- summary(m)$tTable
    r2  <- tryCatch(MuMIn::r.squaredGLMM(m), error = function(e) c(R2m = NA, R2c = NA))
    seas <- wd %>% count(season, sort = TRUE) %>% slice(1) %>% pull(season)

    tibble(
      response   = response_var,
      date_start = as.Date(w_name),
      season     = seas,
      beta       = sm["land_use", "Value"],
      se         = sm["land_use", "Std.Error"],
      p_value    = sm["land_use", "p-value"],
      r2_marg    = as.numeric(r2["R2m"]),
      r2_cond    = as.numeric(r2["R2c"])
    )
  })
}

# Response variables: both microclimate indices, root, fauna (all_taxa)
h1_responses <- c(microclimate_index[1], microclimate_index[2],
                  "root_growth_cm2_count_24h", "all_taxa")

h1_results <- map_dfr(h1_responses, ~ fit_h1_windows(.x, data_sem_raw,
                                                      WIN_DAYS, MIN_RATE))

# Clean display labels
response_labels_h1 <- setNames(
  c(paste0("\u03b2\u2081 (", microclimate_index[1], ")"),
    paste0("\u03b2\u2082 (", microclimate_index[2], ")"),
    "\u03b2\u2085 (root growth)",
    "\u03b2\u2089 (faunal activity)"),
  h1_responses
)

h1_results <- h1_results %>%
  mutate(
    label = response_labels_h1[response],
    group = case_when(
      response %in% microclimate_index ~ "Microclimate",
      response == "root_growth_cm2_count_24h" ~ "Root growth",
      TRUE ~ "Faunal activity"
    ),
    group  = factor(group, levels = c("Microclimate", "Root growth",
                                      "Faunal activity")),
    season = factor(season, levels = season_order)
  )

# Global summary
h1_global <- h1_results %>%
  group_by(label, group) %>%
  summarise(
    mean_b  = mean(beta, na.rm = TRUE),
    sd_b    = sd(beta,   na.rm = TRUE),
    pct_sig = mean(p_value < 0.05, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    sig_label = paste0(round(pct_sig, 0), "% sig."),
    direction = if_else(mean_b > 0, "#c1603a", "#3d5a3e"),
    linetype  = case_when(
      group == "Microclimate"    ~ "dotted",
      group == "Root growth"     ~ "dashed",
      group == "Faunal activity" ~ "solid"
    )
  ) %>%
  arrange(group, mean_b) %>%
  mutate(label = fct_inorder(label))

label_colors_h1 <- deframe(select(h1_global, label, direction))

# Seasonal summary
h1_season <- h1_results %>%
  group_by(label, group, season) %>%
  summarise(
    mean_b  = mean(beta, na.rm = TRUE),
    sd_b    = sd(beta,   na.rm = TRUE),
    pct_sig = mean(p_value < 0.05, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    b_label   = paste0(round(mean_b, 2), " \u00b1 ", round(sd_b, 2)),
    sig_label = paste0(round(pct_sig, 0), "%")
  )
```

```{r fig-h1-landuse}
#| label: fig-h1-landuse
#| fig-cap: >
#|   Direct effect of land-use management on each soil variable, estimated
#|   from independent rolling-window LME (response ~ land_use). Each line
#|   connects the unmanaged baseline (A = 0) to the mean standardised
#|   coefficient in the managed system (C); error bars = ± 1 SD across
#|   windows. Dotted = microclimate; dashed = root growth; solid = faunal
#|   activity. Labels = proportion of windows where p < 0.05.
#|   Terracotta = higher in C; olive = lower in C.
#| fig-width: 7
#| fig-height: 5

p_h1 <- ggplot() +

  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.25,
             linetype = "dashed") +

  geom_segment(
    data = h1_global,
    aes(x = 0, xend = 1, y = 0, yend = mean_b,
        color = label, linetype = linetype),
    linewidth = 0.55, lineend = "round", show.legend = FALSE
  ) +

  geom_linerange(
    data = h1_global,
    aes(x = 1, ymin = mean_b - sd_b, ymax = mean_b + sd_b, color = label),
    linewidth = 0.4, show.legend = FALSE
  ) +

  geom_point(data = h1_global,
             aes(x = 0, y = 0,      color = label), size = 1.2, shape = 16) +
  geom_point(data = h1_global,
             aes(x = 1, y = mean_b, color = label), size = 1.2, shape = 16) +

  ggrepel::geom_text_repel(
    data = h1_global,
    aes(x = 1, y = mean_b, color = label,
        label    = paste0(label, "  ", sig_label),
        fontface = if_else(group == "Faunal activity", "plain", "bold")),
    nudge_x = 0.08, hjust = 0, size = 2.1,
    segment.size = 0.2, segment.color = "grey70",
    direction = "y", box.padding = 0.12, show.legend = FALSE
  ) +

  scale_color_manual(values = label_colors_h1, guide = "none") +
  scale_linetype_identity() +
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c("Unmanaged (A)", "Managed (C)"),
    expand = expansion(mult = c(0.12, 0.35))
  ) +
  labs(x = NULL,
       y = "Land-use effect (standardised \u03b2)") +
  theme_sem() +
  theme(panel.grid.major.x = element_blank(),
        axis.ticks.x       = element_blank())

p_h1

tiff("../output/Fig1_landuse_direct_effects.tiff",
     width = 12, height = 12, units = "cm", res = 300)
print(p_h1)
invisible(dev.off())
```

```{r tbl-h1}
#| label: tbl-h1
#| tbl-cap: >
#|   Seasonal breakdown of direct land-use effects per response variable.
#|   Mean β ± SD across rolling windows within each season; % sig = proportion
#|   of windows with p < 0.05. Terracotta = higher under management; olive = lower.

table_levels_h1 <- h1_global %>%
  arrange(group, mean_b) %>%
  pull(label) %>%
  as.character()

h1_wide <- h1_season %>%
  select(label, season, b_label, sig_label) %>%
  pivot_wider(names_from  = season,
              values_from = c(b_label, sig_label),
              names_glue  = "{season}_{.value}") %>%
  mutate(label = factor(label, levels = table_levels_h1)) %>%
  arrange(label)

col_order_h1 <- as.vector(rbind(paste0(season_order, "_b_label"),
                                paste0(season_order, "_sig_label")))
display_h1 <- h1_wide %>% select(Label = label, all_of(col_order_h1))

col_names_h1  <- c("Variable", rep(c("Mean \u03b2 \u00b1 SD", "% sig."), 4))
b_pos_h1      <- which(str_detect(names(display_h1), "_b_label$"))
sig_pos_h1    <- which(str_detect(names(display_h1), "_sig_label$"))

h1_means <- h1_season %>%
  select(label, season, mean_b) %>%
  pivot_wider(names_from = season, values_from = mean_b,
              names_glue = "{season}_mean_b") %>%
  mutate(label = factor(label, levels = table_levels_h1)) %>%
  arrange(label)

n_mc_h1   <- sum(h1_global$group == "Microclimate")
n_root_h1 <- sum(h1_global$group == "Root growth")
n_fn_h1   <- sum(h1_global$group == "Faunal activity")

tbl_h1 <- display_h1 %>%
  kable(col.names = col_names_h1) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),
                full_width = TRUE, font_size = 11) %>%
  add_header_above(setNames(c(1L, rep(2L, 4)), c(" ", season_order)),
                   bold = TRUE, background = "#ebebeb") %>%
  row_spec(0, bold = TRUE, background = "#d8d8d8") %>%
  pack_rows("Microclimate (\u03b2\u2081, \u03b2\u2082)", 1, n_mc_h1,
            label_row_css = "background-color:#e8e8e8; font-weight:bold;") %>%
  pack_rows("Root growth (\u03b2\u2085)", n_mc_h1 + 1, n_mc_h1 + n_root_h1,
            label_row_css = "background-color:#eeeeee; font-weight:bold;") %>%
  pack_rows("Faunal activity (\u03b2\u2089)", n_mc_h1 + n_root_h1 + 1,
            n_mc_h1 + n_root_h1 + n_fn_h1,
            label_row_css = "background-color:#f4f4f4; font-weight:bold;") %>%
  reduce(seq_along(b_pos_h1), function(t, i) {
    vals <- h1_means[[paste0(season_order[i], "_mean_b")]]
    column_spec(t, b_pos_h1[i], bold = TRUE,
                color = if_else(vals > 0, "#c1603a",
                        if_else(vals < 0, "#3d5a3e", "grey55")))
  }, .init = .) %>%
  reduce(sig_pos_h1, function(t, col_i) {
    column_spec(t, col_i, italic = TRUE, color = "grey45")
  }, .init = .)

tbl_h1
```

------------------------------------------------------------------------

## 2. H2 — Indirect effects of land use through environmental pathways

> **Through which causal pathways does land use influence faunal activity, and how does the relative contribution of each pathway vary across taxa and seasons?**

The additive SEM produces one standardised coefficient per directed link per rolling window. Indirect effects are computed per window as **products of coefficients** along each causal route (see the table in the Overview). This gives a distribution of indirect effects over time — directly comparable in magnitude to the direct effect — showing how much of the total land-use signal reaches fauna through each pathway.

```{r h2-indirect-effects}
#| label: h2-indirect-effects

# ── Pivot SEM coefficients to wide format: one column per (predictor, response) ─
# Column naming convention: predictor__response
coefs_wide <- data_sem_results %>%
  filter(!str_detect(predictor, "~~"),
         !is.na(std_estimate)) %>%
  mutate(coef_id = paste0(predictor, "__", response)) %>%
  select(taxon, date_start, season, coef_id, std_estimate) %>%
  pivot_wider(names_from = coef_id, values_from = std_estimate,
              values_fn = mean)

# Helper: safe column retrieval (returns NA if column absent)
get_col <- function(df, col) {
  if (col %in% names(df)) df[[col]] else rep(NA_real_, nrow(df))
}

# ── Compute indirect effects per window × taxon ───────────────────────────────
# β notation mirrors the SEM equation overview:
#   β1 = lu → mc1,  β2 = lu → mc2
#   β3 = mc1 → root, β4 = mc2 → root, β5 = lu → root
#   β6 = mc1 → fauna, β7 = mc2 → fauna, β8 = root → fauna, β9 = lu → fauna

mc1 <- microclimate_index[1]
mc2 <- microclimate_index[2]

indirect_effects <- coefs_wide %>%
  mutate(
    b1 = get_col(., paste0("land_use__", mc1)),
    b2 = get_col(., paste0("land_use__", mc2)),
    b3 = get_col(., paste0(mc1, "__root")),
    b4 = get_col(., paste0(mc2, "__root")),
    b5 = get_col(., "land_use__root"),
    b6 = get_col(., paste0(mc1, "__fauna")),
    b7 = get_col(., paste0(mc2, "__fauna")),
    b8 = get_col(., "root__fauna"),
    b9 = get_col(., "land_use__fauna"),

    direct       = b9,
    via_mc1      = b1 * b6,
    via_mc2      = b2 * b7,
    via_root     = b5 * b8,
    via_mc1_root = b1 * b3 * b8,
    via_mc2_root = b2 * b4 * b8,

    total_indirect = via_mc1 + via_mc2 + via_root + via_mc1_root + via_mc2_root,
    total_effect   = direct + total_indirect
  ) %>%
  select(taxon, date_start, season,
         direct, via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
         total_indirect, total_effect)

# Long format for plotting
indirect_long <- indirect_effects %>%
  pivot_longer(cols = c(direct, via_mc1, via_mc2, via_root,
                        via_mc1_root, via_mc2_root,
                        total_indirect, total_effect),
               names_to  = "effect_type",
               values_to = "effect") %>%
  mutate(
    effect_type = factor(effect_type, levels = names(indirect_labels)),
    label       = indirect_labels[effect_type],
    label       = factor(label, levels = unname(indirect_labels)),
    season      = factor(season, levels = season_order)
  ) %>%
  filter(!is.na(effect))

# Wilcoxon test per effect type × taxon: is the distribution shifted from zero?
sig_stars <- function(p) case_when(p < 0.001 ~ "***", p < 0.01 ~ "**",
                                   p < 0.05 ~ "*", TRUE ~ "ns")

wilcox_indirect <- indirect_long %>%
  group_by(taxon, effect_type, label) %>%
  summarise(
    p_wilcox = tryCatch(
      wilcox.test(effect, mu = 0, exact = FALSE)$p.value,
      error = function(e) NA_real_
    ),
    median_effect = median(effect, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(stars = sig_stars(p_wilcox))
```

```{r fig-h2-indirect-community}
#| label: fig-h2-indirect-community
#| fig-cap: >
#|   Distributions of direct, indirect, and total effects of land use on faunal
#|   activity (community-level, all_taxa run) across rolling windows. Each
#|   effect type is computed per window as a product of standardised SEM
#|   coefficients along the corresponding causal route (see Overview table).
#|   Half-eye = kernel density; box = IQR; dots = individual window estimates.
#|   Stars: one-sample Wilcoxon test of the effect distribution against zero
#|   (*** p < 0.001, ** p < 0.01, * p < 0.05, ns). Total indirect and total
#|   effect are shown in darker tones for visual emphasis.
#| fig-width: 9
#| fig-height: 5

# ── Outlier filter applied to indirect effects ────────────────────────────────
q1_ie  <- quantile(indirect_long$effect, 0.25, na.rm = TRUE)
q3_ie  <- quantile(indirect_long$effect, 0.75, na.rm = TRUE)
iqr_ie <- q3_ie - q1_ie

indirect_plot <- indirect_long %>%
  filter(effect >= q1_ie - 1.5 * iqr_ie,
         effect <= q3_ie + 1.5 * iqr_ie)

p_indirect_all <- indirect_plot %>%
  filter(taxon == "all_taxa") %>%
  ggplot(aes(y = label, x = effect,
             fill = effect_type, color = effect_type)) +

  geom_vline(xintercept = 0, color = "grey30", linewidth = 0.35,
             linetype = "dashed") +

  stat_halfeye(adjust = 1.2, width = 0.55, justification = -0.08,
               .width = 0, point_colour = NA, alpha = 0.75) +

  geom_boxplot(width = 0.13, outliers = FALSE, linewidth = 0.3,
               color = "grey20", fill = "white", alpha = 0.6) +

  stat_dots(side = "bottom", justification = 1.08,
            binwidth = unit(c(0.002, 0.012), "npc"),
            color = NA, alpha = 0.5, layout = "hex", dotsize = 0.8) +

  geom_text(
    data = wilcox_indirect %>% filter(taxon == "all_taxa"),
    aes(x = Inf, y = label, label = stars),
    inherit.aes = FALSE,
    hjust = 1.1, vjust = 0.4, size = 3, fontface = "bold",
    color = if_else(
      (wilcox_indirect %>% filter(taxon == "all_taxa"))$stars == "ns",
      "grey55", "grey15")
  ) +

  scale_fill_manual(values  = indirect_palette) +
  scale_color_manual(values = indirect_palette) +
  guides(fill = "none", color = "none") +

  labs(
    x = "Effect on faunal activity (standardised \u03b2 product)",
    y = NULL
  ) +
  theme_sem() +
  theme(axis.text.y = element_text(size = 7))

p_indirect_all

tiff("../output/Fig2a_indirect_effects_community.tiff",
     width = 16, height = 12, units = "cm", res = 300)
print(p_indirect_all)
invisible(dev.off())
```

```{r fig-h2-indirect-taxa}
#| label: fig-h2-indirect-taxa
#| fig-cap: >
#|   Median direct and indirect effects of land use on faunal activity per
#|   taxonomic group and effect type. Each panel shows the median effect
#|   (point) and IQR (error bar) across rolling windows. Effect types are
#|   ordered as in Fig. 2a. Consistent patterns across taxa indicate a
#|   community-level signal; divergent patterns indicate taxon-specific
#|   pathway use. Only taxa other than all_taxa are shown.
#| fig-width: 9
#| fig-height: 6

indirect_taxa_summary <- indirect_long %>%
  filter(taxon != "all_taxa",
         effect_type %in% c("direct", "via_mc1", "via_mc2",
                            "via_root", "total_indirect")) %>%
  group_by(taxon, effect_type, label) %>%
  summarise(
    median_e = median(effect, na.rm = TRUE),
    q25_e    = quantile(effect, 0.25, na.rm = TRUE),
    q75_e    = quantile(effect, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

p_indirect_taxa <- indirect_taxa_summary %>%
  ggplot(aes(x = taxon, y = median_e,
             color = effect_type, fill = effect_type)) +

  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.3,
             linetype = "dashed") +

  geom_linerange(aes(ymin = q25_e, ymax = q75_e),
                 linewidth = 0.5, position = position_dodge(width = 0.6)) +

  geom_point(size = 1.8, shape = 21, color = "white", stroke = 0.3,
             position = position_dodge(width = 0.6)) +

  scale_color_manual(values = indirect_palette,
                     labels = indirect_labels, name = "Effect type") +
  scale_fill_manual(values  = indirect_palette,
                    labels  = indirect_labels, name = "Effect type") +

  facet_wrap(~ effect_type, ncol = 3,
             labeller = labeller(effect_type = indirect_labels)) +

  labs(
    x = "Taxon",
    y = "Effect on faunal activity (median \u03b2 product, IQR)"
  ) +
  guides(color = "none", fill = "none") +
  theme_sem() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 7))

p_indirect_taxa

tiff("../output/Fig2b_indirect_effects_taxa.tiff",
     width = 16, height = 12, units = "cm", res = 300)
print(p_indirect_taxa)
invisible(dev.off())
```

```{r tbl-h2-indirect-summary}
#| label: tbl-h2-indirect-summary
#| tbl-cap: >
#|   Summary of direct and indirect effects of land use on faunal activity
#|   per effect type and season (community-level, all_taxa). Mean ± SD across
#|   rolling windows; % sig = proportion of windows where the Wilcoxon test
#|   against zero gave p < 0.05. Positive = management increases the effect;
#|   negative = management decreases it.

indirect_season_summary <- indirect_long %>%
  filter(taxon == "all_taxa") %>%
  group_by(label, effect_type, season) %>%
  summarise(
    mean_e  = mean(effect, na.rm = TRUE),
    sd_e    = sd(effect,   na.rm = TRUE),
    pct_sig = mean(
      abs(effect) > quantile(abs(effect), 0.95, na.rm = TRUE),
      na.rm = TRUE
    ) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    b_label   = paste0(round(mean_e, 3), " \u00b1 ", round(sd_e, 3)),
    sig_label = paste0(round(pct_sig, 0), "%")
  )

ind_wide <- indirect_season_summary %>%
  select(label, season, b_label, sig_label) %>%
  pivot_wider(names_from  = season,
              values_from = c(b_label, sig_label),
              names_glue  = "{season}_{.value}") %>%
  mutate(label = factor(label, levels = unname(indirect_labels))) %>%
  arrange(label)

col_order_ind <- as.vector(rbind(paste0(season_order, "_b_label"),
                                 paste0(season_order, "_sig_label")))
display_ind <- ind_wide %>% select(Label = label, all_of(col_order_ind))

b_pos_ind   <- which(str_detect(names(display_ind), "_b_label$"))
sig_pos_ind <- which(str_detect(names(display_ind), "_sig_label$"))

ind_means <- indirect_season_summary %>%
  select(label, season, mean_e) %>%
  pivot_wider(names_from = season, values_from = mean_e,
              names_glue = "{season}_mean_e") %>%
  mutate(label = factor(label, levels = unname(indirect_labels))) %>%
  arrange(label)

display_ind %>%
  kable(col.names = c("Effect type", rep(c("Mean \u03b2 \u00b1 SD", "% large"), 4))) %>%
  kable_styling(bootstrap_options = c("hover", "condensed"),
                full_width = TRUE, font_size = 11) %>%
  add_header_above(setNames(c(1L, rep(2L, 4)), c(" ", season_order)),
                   bold = TRUE, background = "#ebebeb") %>%
  row_spec(0, bold = TRUE, background = "#d8d8d8") %>%
  row_spec(which(ind_wide$label %in% c(indirect_labels["total_indirect"],
                                        indirect_labels["total_effect"])),
           bold = TRUE) %>%
  reduce(seq_along(b_pos_ind), function(t, i) {
    vals <- ind_means[[paste0(season_order[i], "_mean_e")]]
    column_spec(t, b_pos_ind[i], bold = TRUE,
                color = if_else(vals > 0, "#c1603a",
                        if_else(vals < 0, "#3d5a3e", "grey55")))
  }, .init = .) %>%
  reduce(sig_pos_ind, function(t, col_i) {
    column_spec(t, col_i, italic = TRUE, color = "grey45")
  }, .init = .)
```

------------------------------------------------------------------------

## 3. H3 — Temporal stability of causal rules

> **Are the causal links governing faunal activity stable over time?**

Each rolling window yields one estimate of every path coefficient. The dispersion of this distribution (CV) quantifies how variable the causal rule is over the study period — a high CV indicates a link whose strength fluctuates, while a low CV indicates a stable relationship. Cross-path correlations over time reveal whether the abiotic and biotic links are activated synchronously or independently.

```{r fig-temporal-paths}
#| label: fig-temporal-paths
#| fig-cap: >
#|   Temporal dynamics of standardised SEM path coefficients across the study
#|   period (all_taxa run). Semi-transparent points = individual window estimates;
#|   lines = LOESS trends (span = 0.25); ribbons = 95% pointwise confidence
#|   intervals. Grey band = |β| < 0.2 (approximate null zone). Wide ribbons
#|   indicate high temporal variability of that causal rule.
#| fig-width: 9
#| fig-height: 5

temporal_data <- data_plot %>%
  filter(taxon == "null model", !is.na(path_id))

date_range_sem <- range(temporal_data$date_start, na.rm = TRUE)
x_breaks_sem   <- seq(floor_date(date_range_sem[1], "month"),
                      ceiling_date(date_range_sem[2], "month"),
                      by = "3 months")

p_temporal <- ggplot() +

  make_season_layers(date_range_sem, label_size = 2.0) +

  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -0.2, ymax = 0.2,
           fill = "grey50", alpha = 0.08) +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.3) +

  geom_point(data = temporal_data,
             aes(date_start, std_estimate, color = path_id),
             alpha = 0.18, size = 0.15, shape = 16) +

  geom_smooth(data = temporal_data,
              aes(date_start, std_estimate, color = path_id, fill = path_id),
              method = "loess", span = 0.25,
              se = TRUE, linewidth = 0.7, alpha = 0.12) +

  scale_color_manual(values = path_palette, name = NULL) +
  scale_fill_manual(values  = path_palette, name = NULL) +
  scale_x_date(breaks = x_breaks_sem, date_labels = "%b\n%y",
               expand = expansion(mult = c(0.01, 0.01))) +
  scale_y_continuous(limits = c(-1.1, 1.1), breaks = seq(-1, 1, 0.5)) +

  facet_wrap(~ path_id, ncol = 2,
             labeller = labeller(path_id = label_wrap_gen(22))) +

  labs(x = NULL,
       y = expression("Standardised path coefficient"~(beta))) +
  theme_sem(legend_pos = "bottom") +
  theme(
    panel.spacing.x  = unit(10, "pt"),
    panel.spacing.y  = unit(8,  "pt"),
    axis.text.x      = element_text(size = 7.5, lineheight = 0.85, hjust = 0.5),
    legend.direction = "horizontal"
  )

p_temporal

tiff("../output/Fig3_temporal_dynamics.tiff",
     width = 15, height = 13, units = "cm", res = 300)
print(p_temporal)
invisible(dev.off())
```

```{r tbl-temporal-stability}
#| label: tbl-temporal-stability
#| tbl-cap: >
#|   Temporal instability of SEM path coefficients per season (all_taxa run).
#|   Mean β ± SD and CV (coefficient of variation = SD / |mean| × 100)
#|   quantify typical strength and temporal variability of each causal rule.
#|   A high CV indicates a link whose strength fluctuates substantially across
#|   the study period; a low CV indicates a stable relationship.

stability_stats <- data_sem_results %>%
  filter(taxon == "null model", !is.na(path_id)) %>%
  group_by(path_id, season) %>%
  summarise(
    mean_b = mean(std_estimate, na.rm = TRUE),
    sd_b   = sd(std_estimate,   na.rm = TRUE),
    cv     = (sd_b / abs(mean_b)) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    label_b  = paste0(round(mean_b, 2), " \u00b1 ", round(sd_b, 2)),
    label_cv = as.character(round(cv, 1))
  )

global_stability <- data_sem_results %>%
  filter(taxon == "null model", !is.na(path_id)) %>%
  group_by(path_id) %>%
  summarise(
    mean_b = mean(std_estimate, na.rm = TRUE),
    sd_b   = sd(std_estimate,   na.rm = TRUE),
    cv     = (sd_b / abs(mean_b)) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    season   = "Whole period",
    label_b  = paste0(round(mean_b, 2), " \u00b1 ", round(sd_b, 2)),
    label_cv = as.character(round(cv, 1))
  )

stab_long <- bind_rows(global_stability, stability_stats) %>%
  mutate(season = factor(season, levels = c("Whole period", season_order))) %>%
  select(path_id, season, label_b, label_cv) %>%
  pivot_longer(cols = c(label_b, label_cv), names_to = "metric") %>%
  pivot_wider(names_from = path_id, values_from = value) %>%
  arrange(season) %>%
  select(-season) %>%
  mutate(Metric = if_else(metric == "label_b",
                          "Mean \u03b2 \u00b1 SD",
                          "CV (%)")) %>%
  select(Metric, everything(), -metric)

cv_rows <- seq(2, nrow(stab_long), by = 2)

stab_tbl <- stab_long %>%
  kable(booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                font_size = 11, full_width = TRUE) %>%
  pack_rows("Whole period", 1, 2,
            label_row_css = "background-color:#e8e8e8; font-weight:bold;") %>%
  pack_rows("Spring",  3,  4, label_row_css = "background-color:#f4f4f4;") %>%
  pack_rows("Summer",  5,  6, label_row_css = "background-color:#f4f4f4;") %>%
  pack_rows("Autumn",  7,  8, label_row_css = "background-color:#f4f4f4;") %>%
  pack_rows("Winter",  9, 10, label_row_css = "background-color:#f4f4f4;") %>%
  column_spec(1, bold = TRUE, width = "14em")

# Colour CV rows: high CV (> 100%) in terracotta, low (< 50%) in olive
cv_vals_raw <- bind_rows(global_stability, stability_stats) %>%
  mutate(season = factor(season, levels = c("Whole period", season_order))) %>%
  arrange(season) %>%
  pull(cv)

reduce(2:ncol(stab_long), function(t, col_i) {
  cell_colors <- map_chr(1:nrow(stab_long), function(row_j) {
    if (row_j %in% cv_rows) {
      val <- as.numeric(stab_long[[col_i]][row_j])
      if (is.na(val))  return("black")
      if (val > 100)   return("#c1603a")
      if (val < 50)    return("#3d5a3e")
      return("grey30")
    } else return("black")
  })
  column_spec(t, col_i, color = cell_colors,
              bold = if_else(1:nrow(stab_long) %in% cv_rows, TRUE, FALSE))
}, .init = stab_tbl)
```
