---
title: "From pixels to patterns — results and visualisation"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: false
---

# Overview

## Role in the pipeline

Final step of the SEM pipeline. Loads pre-computed databases from
`3_SEM_modelisation.qmd` and produces all tables and figures.
**No modelling is performed here.**

| Input file | Produced by | Used for |
|---|---|---|
| `H1_adjusted_values.csv` | 3_SEM_modelisation | H1 boxplot + temporal figure |
| `H1_model_summary.csv` | 3_SEM_modelisation | H1 effect-size annotation |
| `DAG_coefficients.csv` | 3_SEM_modelisation | H2 DAG annotation table |
| `DAG_model_fit.csv` | 3_SEM_modelisation | H2 model-fit footnote |
| `SEM_results_database.csv` | 3_SEM_modelisation | H2 temporal time-series |
| `indirect_effects.csv` | 3_SEM_modelisation | H2 temporal time-series |

---

```{r setup}
invisible(Sys.setlocale("LC_TIME", "C"))
dir.create("../output", showWarnings = FALSE)

libs <- c(
  "tidyverse", "lubridate", "patchwork", "jsonlite",
  "kableExtra", "colorspace"
)
invisible(lapply(libs, require, character.only = TRUE))

# ── Index files ───────────────────────────────────────────────────────────────
microclimate_index <- readLines("../output/microclimate_index.txt")
mc1_pred <- microclimate_index[1]
mc2_pred <- microclimate_index[2]

# ── Colour palette ────────────────────────────────────────────────────────────
pos_colours <- c("Treed (A)" = "#3d5a3e", "Cultivated (C)" = "#c1603a")

path_meta <- tribble(
  ~path_id,         ~label,                                                ~colour,   ~ltype,
  "direct",         "Direct (lu\u2192fauna)",                              "#DCB879", "solid",
  "via_mc1",        paste0("lu\u2192", mc1_pred, "\u2192fauna"),           "#c2cab2", "dashed",
  "via_mc2",        paste0("lu\u2192", mc2_pred, "\u2192fauna"),           "#555d45", "dashed",
  "via_root",       "lu\u2192root\u2192fauna",                             "#955f42", "dashed",
  "via_mc1_root",   paste0("lu\u2192", mc1_pred, "\u2192root\u2192fauna"), "#dea88b", "dotted",
  "via_mc2_root",   paste0("lu\u2192", mc2_pred, "\u2192root\u2192fauna"), "#d1855c", "dotted",
  "total_indirect", "Total indirect",                                       "#999999", "solid",
  "total_effect",   "Total effect",                                         "#444444", "solid"
)

path_colours <- setNames(path_meta$colour, path_meta$path_id)
path_ltypes  <- setNames(path_meta$ltype,  path_meta$path_id)
path_labels  <- setNames(path_meta$label,  path_meta$path_id)

# ── Season helpers ────────────────────────────────────────────────────────────
season_order <- c("Spring", "Summer", "Autumn", "Winter")
season_fills <- c(Spring = "#F2F2F2", Summer = "#E4E4E4",
                  Autumn = "#F2F2F2", Winter = "#E4E4E4")

make_season_layers <- function(date_range, y_label_pos = Inf, label_size = 2.2) {
  bands <- tibble(
    season  = names(season_fills),
    m_start = c(3L, 6L, 9L, 12L),
    m_end   = c(6L, 9L, 12L,  3L),
    fill    = unname(season_fills)
  ) %>%
    crossing(year = seq(year(date_range[1]), year(date_range[2]))) %>%
    mutate(
      xmin = as.Date(paste(year, m_start, "01", sep = "-")),
      xmax = as.Date(paste(if_else(m_end < m_start, year + 1L, year),
                           m_end, "01", sep = "-")),
      xmid = xmin + (xmax - xmin) / 2
    ) %>%
    filter(xmin < date_range[2], xmax > date_range[1]) %>%
    mutate(xmin = pmax(xmin, date_range[1]),
           xmax = pmin(xmax, date_range[2]))

  rect_layers <- pmap(
    bands %>% select(xmin, xmax, fill),
    \(xmin, xmax, fill) annotate("rect", xmin = xmin, xmax = xmax,
                                 ymin = -Inf, ymax = Inf, fill = fill, alpha = 1)
  )
  label_layer <- geom_text(
    data = bands, aes(x = xmid, y = y_label_pos, label = season),
    inherit.aes = FALSE, vjust = 1.6, size = label_size,
    color = "grey60", fontface = "italic", family = "sans"
  )
  c(rect_layers, list(label_layer))
}

# ── Base theme ────────────────────────────────────────────────────────────────
theme_sem <- function(base_size = 8, legend_pos = "none") {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.border      = element_rect(color = "grey20", fill = NA, linewidth = 0.3),
      panel.grid.major  = element_line(color = "grey94", linewidth = 0.25),
      panel.grid.minor  = element_blank(),
      panel.background  = element_rect(fill = "white", color = NA),
      axis.line         = element_blank(),
      axis.ticks        = element_line(color = "grey40", linewidth = 0.3),
      axis.ticks.length = unit(2, "pt"),
      axis.title        = element_text(size = base_size - 0.5, color = "grey20"),
      axis.text         = element_text(size = base_size - 1.5, color = "grey30"),
      strip.text        = element_text(size = base_size - 1, face = "bold",
                                       color = "grey15"),
      strip.background  = element_rect(fill = "grey96", color = "grey96",
                                       linewidth = 0.3),
      plot.title        = element_text(size = base_size + 0.5, face = "bold",
                                       hjust = 0, color = "grey10",
                                       margin = margin(b = 3)),
      plot.subtitle     = element_text(size = base_size - 1, color = "grey45",
                                       hjust = 0, margin = margin(b = 5)),
      plot.caption      = element_text(size = base_size - 2, color = "grey55",
                                       hjust = 0, margin = margin(t = 4)),
      legend.position   = legend_pos,
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.title      = element_text(size = base_size - 1, face = "bold",
                                       color = "grey20"),
      legend.text       = element_text(size = base_size - 1.5, color = "grey30"),
      legend.key.size   = unit(8, "pt"),
      legend.spacing.x  = unit(3, "pt"),
      plot.margin       = margin(6, 6, 6, 6)
    )
}

# ── Load all databases ────────────────────────────────────────────────────────
h1_adj     <- read.csv("../output/H1_adjusted_values.csv",
                       na.strings = c("", "NA")) %>%
  mutate(date           = as.Date(date),
         position_label = factor(position_label,
                                 levels = c("Treed (A)", "Cultivated (C)")),
         season         = factor(season, levels = season_order))

h1_summary <- read.csv("../output/H1_model_summary.csv", na.strings = c("", "NA"))

dag_table  <- read.csv("../output/DAG_coefficients.csv",  na.strings = c("", "NA"))
dag_fit    <- read.csv("../output/DAG_model_fit.csv",     na.strings = c("", "NA"))

sem_results <- read.csv("../output/SEM_results_database.csv",
                        na.strings = c("", "NA")) %>%
  mutate(date_start = as.Date(date_start),
         season     = factor(season, levels = season_order))

indirect_effects <- read.csv("../output/indirect_effects.csv",
                              na.strings = c("", "NA")) %>%
  mutate(date_start = as.Date(date_start))

transfo_fauna <- h1_summary$transformation[1]
```

---

## 1. H1 — Direct effects of land use on faunal activity

> **Land-use type directly structures faunal activity, with treed habitats
> sustaining lower but more temporally stable activity than cultivated zones.**

Values are adjusted for orientation and depth random effects
(y_adj = Xβ̂ + ε̂ = y − Zb̂). This removes grouping-level variance while
preserving the land_use signal on the original transformed scale.

```{r h1-prep}
# ── Daily aggregates for temporal panel ──────────────────────────────────────
h1_daily <- h1_adj %>%
  group_by(date, position_label, season) %>%
  summarise(mean_adj = mean(adj_value, na.rm = TRUE),
            se_adj   = sd(adj_value,   na.rm = TRUE) / sqrt(n()),
            .groups  = "drop")

date_range_h1 <- range(h1_daily$date)

# Effect-size annotation strings from pre-computed summary
hedges_g <- round(h1_summary$hedges_g, 2)
p_lu     <- h1_summary$p_value
p_label  <- if (p_lu < 0.001) "p < 0.001" else paste0("p = ", round(p_lu, 3))
```

```{r fig-h1}
#| fig-cap: "H1 — Faunal activity by land-use type (random-effect adjusted values)"
#| fig-width: 14
#| fig-height: 5.5

# ── Panel A: pooled distribution ─────────────────────────────────────────────
p_box <- ggplot(h1_adj,
                aes(x = position_label, y = adj_value,
                    fill = position_label)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_boxplot(width = 0.45, outlier.size = 0.4, outlier.alpha = 0.25,
               outlier.color = "grey50", linewidth = 0.35, fatten = 1.2) +
  annotate("text",
    x = 1.5, y = max(h1_adj$adj_value, na.rm = TRUE) * 0.97,
    label    = paste0("g = ", hedges_g, "\n", p_label),
    size     = 2.8, color = "grey25", fontface = "italic",
    hjust    = 0.5, lineheight = 1.2) +
  scale_fill_manual(values = pos_colours) +
  labs(x = NULL,
       y        = paste0("Faunal activity\n(",
                         transfo_fauna, "-transformed, RE-adjusted)"),
       title    = "A — Distribution by land-use type",
       subtitle = "Pooled across all dates and depths") +
  theme_sem() +
  theme(axis.text.x = element_text(size = 7.5, face = "bold"))

# ── Panel B: temporal fluctuations ───────────────────────────────────────────
p_temp <- ggplot(h1_daily,
                 aes(x = date, color = position_label,
                     fill = position_label)) +
  make_season_layers(date_range_h1,
                     y_label_pos = max(h1_daily$mean_adj, na.rm = TRUE)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_ribbon(aes(ymin = mean_adj - se_adj, ymax = mean_adj + se_adj),
              alpha = 0.12, color = NA) +
  geom_point(aes(y = mean_adj), size = 0.6, alpha = 0.25, shape = 16) +
  geom_smooth(aes(y = mean_adj), method = "loess", formula = y ~ x,
              span = 0.18, linewidth = 0.9, se = FALSE) +
  scale_color_manual(values = pos_colours, name = "Land-use type") +
  scale_fill_manual( values = pos_colours, name = "Land-use type") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  labs(x = NULL,
       y        = "Daily mean adjusted activity\n(\u00b1 SE across sensors)",
       title    = "B — Temporal fluctuations",
       subtitle = "Daily mean \u00b1 SE; LOESS smoother (span = 0.18)") +
  theme_sem(legend_pos = "top") +
  theme(axis.text.x      = element_text(angle = 30, hjust = 1, size = 6.5),
        legend.direction = "horizontal")

# ── Combine ───────────────────────────────────────────────────────────────────
p_h1 <- p_box + p_temp +
  plot_layout(widths = c(1, 2.8)) +
  plot_annotation(
    title    = "H1 — Effect of land-use type on faunal activity",
    subtitle = paste0(
      "RE-adjusted values (y \u2212 Z\u03b2\u0302) | ",
      "Model: all_taxa ~ land_use + (1 | orientation/depth) + AR(1) | ",
      "n = ", nrow(h1_adj), " obs."
    ),
    theme = theme(
      plot.title    = element_text(size = 9, face = "bold", color = "grey10"),
      plot.subtitle = element_text(size = 7, color = "grey45",
                                   margin = margin(b = 4))
    )
  )

p_h1

tiff("../output/Fig1_H1_adjusted.tiff",
     width = 21, height = 10, units = "cm", res = 300)
print(p_h1)
invisible(dev.off())
```

---

## 2. H2 — Indirect effects of land use through contrasting causal pathways

> **Land-use effects are mediated through contrasting causal pathways —
> agricultural management drives faunal activity primarily through altered
> microclimate, whereas treed habitats sustain faunal activity more
> consistently through root resource supply.**

### 2a — Full-dataset psem: DAG annotation table

Coefficients from the single piecewise SEM fitted on the complete dataset
(all_taxa). Use these values to annotate the DAG manually.

```{r tbl-h2-dag}
#| tbl-cap: >
#|   Full-dataset piecewise SEM path coefficients for DAG annotation
#|   (all_taxa, complete time series, all depths). Variables z-scored after
#|   transformation. A = 0, C = 1 for land_use.

fit_line <- with(dag_fit, paste0(
  "Model fit \u2014 Fisher\u2019s C = ",
  if (is.na(fishers_C)) "NA" else round(fishers_C, 2),
  " (df = ", if (is.na(df)) "NA" else df, ", p ",
  if (is.na(p_value))       "= NA (saturated model)"
  else if (p_value < 0.001) "< 0.001"
  else                       paste0("= ", round(p_value, 3)),
  ") \u2502 AIC = ",
  if (is.na(AIC)) "NA" else round(AIC, 1)
))

dag_table %>%
  filter(!str_detect(response, "~~")) %>%
  mutate(
    across(c(estimate, std_estimate, se, t_value, p_value,
             r2_marginal, r2_conditional),
           ~ suppressWarnings(as.numeric(.))),
    across(c(estimate, std_estimate, se), ~ sprintf("%.3f", .)),
    t_fmt   = if_else(is.na(t_value), "\u2014", sprintf("%.2f", t_value)),
    p_fmt   = if_else(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value)),
    r2_marg = sprintf("%.3f", r2_marginal),
    r2_cond = sprintf("%.3f", r2_conditional)
  ) %>%
  select(
    "Equation"   = equation,
    "Predictor"  = predictor,
    "\u03b2"     = estimate,
    "Std \u03b2" = std_estimate,
    "SE"         = se,
    "t"          = t_fmt,
    "p"          = p_fmt,
    "sig"        = sig,
    "R\u00b2m"   = r2_marg,
    "R\u00b2c"   = r2_cond
  ) %>%
  kbl(format  = "html", escape = FALSE,
      align   = c("l", "l", "r", "r", "r", "r", "r", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8") %>%
  collapse_rows(columns = c(1, 9, 10), valign = "middle") %>%
  footnote(
    general       = paste0(fit_line,
                           "<br>*** p < 0.001 &nbsp; ** p < 0.01 &nbsp;",
                           " * p < 0.05 &nbsp; . p < 0.10"),
    general_title = "", escape = FALSE, footnote_as_chunk = TRUE
  )
```

### 2b — Temporal variation of pathway estimates

```{r h2-temporal-prep}
# ── Combine direct path (from SEM results) with computed indirect effects ─────
# Both restricted to all_taxa (community level).

direct_ts <- sem_results %>%
  filter(taxon == "all_taxa",
         predictor == "land_use",
         response  == "fauna") %>%
  transmute(path_id = "direct", date_start, season, std_estimate)

indirect_ts <- indirect_effects %>%
  filter(taxon == "all_taxa") %>%
  pivot_longer(c(via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
                 total_indirect, total_effect),
               names_to = "path_id", values_to = "std_estimate") %>%
  filter(!is.na(std_estimate)) %>%
  mutate(
    month  = month(date_start),
    season = factor(case_when(
      month %in% 3:5  ~ "Spring",
      month %in% 6:8  ~ "Summer",
      month %in% 9:11 ~ "Autumn",
      TRUE            ~ "Winter"
    ), levels = season_order)
  ) %>%
  select(path_id, date_start, season, std_estimate)

h2_ts <- bind_rows(direct_ts, indirect_ts) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label),
    season  = factor(season,  levels = season_order)
  ) %>%
  # Winsorise at 2.5–97.5th percentile per path before smoothing
  group_by(path_id) %>%
  mutate(
    lo           = quantile(std_estimate, 0.025, na.rm = TRUE),
    hi           = quantile(std_estimate, 0.975, na.rm = TRUE),
    std_estimate = pmax(lo, pmin(hi, std_estimate))
  ) %>%
  ungroup()

# LOESS smoother per pathway
smooth_path <- function(df, span = 0.20) {
  df <- arrange(df, date_start)
  if (nrow(df) < 5) return(mutate(df, smoothed = std_estimate))
  fit <- tryCatch(
    loess(std_estimate ~ as.numeric(date_start), data = df, span = span),
    error = \(e) NULL
  )
  df$smoothed <- if (!is.null(fit)) predict(fit) else df$std_estimate
  df
}

h2_smoothed <- h2_ts %>%
  group_by(path_id) %>%
  group_modify(~ smooth_path(.x)) %>%
  ungroup()

date_range_h2 <- range(h2_smoothed$date_start)

# Marginal summaries for left-panel dot-range plot
h2_marginal <- h2_ts %>%
  group_by(path_id, label, colour, ltype) %>%
  summarise(
    median_b = median(std_estimate, na.rm = TRUE),
    q025     = quantile(std_estimate, 0.025, na.rm = TRUE),
    q975     = quantile(std_estimate, 0.975, na.rm = TRUE),
    q25      = quantile(std_estimate, 0.25,  na.rm = TRUE),
    q75      = quantile(std_estimate, 0.75,  na.rm = TRUE),
    .groups  = "drop"
  )
```

```{r fig-h2}
#| fig-cap: "H2 — Causal pathway estimates from land use to faunal activity"
#| fig-width: 17
#| fig-height: 7

# ── Panel A: marginal summary (dot + IQR + 95% range) ────────────────────────
p_marginal <- ggplot(h2_marginal,
                     aes(x = median_b, y = label, colour = path_id)) +
  geom_vline(xintercept = 0, linetype = "dashed",
             color = "grey55", linewidth = 0.35) +
  geom_linerange(aes(xmin = q025, xmax = q975),
                 linewidth = 0.55, alpha = 0.40) +
  geom_linerange(aes(xmin = q25, xmax = q75),
                 linewidth = 2.0, alpha = 0.70) +
  geom_point(size = 2.6, shape = 21, fill = "white", stroke = 1.2) +
  scale_colour_manual(values = path_colours, guide = "none") +
  scale_y_discrete(limits = rev(levels(h2_ts$label))) +
  labs(x        = "Standardised \u03b2",
       y        = NULL,
       title    = "A — Marginal distribution",
       subtitle = "Median \u2022 IQR (thick) \u2022 95% range (thin)") +
  theme_sem() +
  theme(axis.text.y = element_text(size = 6.5, hjust = 1),
        plot.margin = margin(5, 2, 5, 5))

# ── Panel B: LOESS-smoothed temporal series ───────────────────────────────────
p_time <- ggplot(h2_smoothed,
                 aes(x     = date_start, y = smoothed,
                     colour = path_id,  linetype = path_id)) +
  make_season_layers(date_range_h2,
                     y_label_pos = max(h2_smoothed$smoothed, na.rm = TRUE)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey55", linewidth = 0.35) +
  geom_line(linewidth = 0.65, alpha = 0.85) +
  scale_colour_manual( values = path_colours, labels = path_labels,
                       name = "Pathway") +
  scale_linetype_manual(values = path_ltypes,  labels = path_labels,
                        name = "Pathway") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  labs(x        = NULL,
       y        = "Standardised \u03b2 (LOESS-smoothed)",
       title    = "B — Temporal fluctuations",
       subtitle = "LOESS smoother (span = 0.20) \u2022 community (all_taxa) only") +
  theme_sem(legend_pos = "right") +
  theme(axis.text.x      = element_text(angle = 30, hjust = 1, size = 6.5),
        legend.text      = element_text(size = 6.5),
        legend.title     = element_text(size = 7, face = "bold"),
        legend.key.width = unit(18, "pt"),
        plot.margin      = margin(5, 5, 5, 2))

# ── Combine ───────────────────────────────────────────────────────────────────
p_h2 <- p_marginal + p_time +
  plot_layout(widths = c(1, 2.6)) +
  plot_annotation(
    title    = "H2 — Causal pathways from land use to faunal activity",
    subtitle = paste0(
      "Community (all_taxa) \u2502 ",
      nrow(filter(h2_ts, path_id == "direct")),
      " rolling windows \u2502 A = 0, C = 1 \u2192 negative = stronger in treed system"
    ),
    theme = theme(
      plot.title    = element_text(size = 9, face = "bold", color = "grey10"),
      plot.subtitle = element_text(size = 7, color = "grey45",
                                   margin = margin(b = 4))
    )
  )

p_h2

tiff("../output/Fig2_H2_temporal_pathways.tiff",
     width = 25, height = 12, units = "cm", res = 300)
print(p_h2)
invisible(dev.off())
```
