---
title: "From pixels to patterns — results and visualisation"
author: "Emma Belaud"
format:
  html:
    css: css/styles.css
    embed-resources: true
    code-tools: true
    toc: true
    toc-depth: 4
execute:
  warning: false
  message: false
  echo: false
---

# Overview

## Role in the pipeline

This is the **final step** of the SEM pipeline, following:

1. `data_preparation.qmd` — assembles the SEM database
2. `sem_diagnostic.qmd` — selects window duration, transformation, and valid windows
3. `3_SEM_modelisation.qmd` — fits all models and writes result databases

**No modelling is performed here.** Every number in every figure comes directly
from pre-computed CSV files. Re-running this script is fast and safe — it never
touches a model.

## Input databases

| File | What it contains |
|---|---|
| `H1_adjusted_values.csv` | Per-observation null-model residuals with date, land-use label, season |
| `H1_model_summary.csv` | β, SE, t, p, Hedge's g, R² from the two-step H1 model |
| `DAG_coefficients.csv` | Path coefficients from the full-dataset piecewise SEM |
| `DAG_model_fit.csv` | Fisher's C, AIC for the full-dataset piecewise SEM |
| `SEM_results_database.csv` | Rolling-window path coefficients for all taxa |
| `indirect_effects.csv` | Rolling-window indirect effect products for all taxa |

---

```{r setup}
# ══════════════════════════════════════════════════════════════════════════════
# SETUP
# Loads packages, reads shared metadata (microclimate variable names), and
# defines all visual constants and helper functions used across sections 1–3.
# ══════════════════════════════════════════════════════════════════════════════

invisible(Sys.setlocale("LC_TIME", "C"))  # Force English month names in plots
dir.create("../output", showWarnings = FALSE)

libs <- c(
  "tidyverse", "lubridate", "patchwork",
  "kableExtra", "colorspace"
)
invisible(lapply(libs, require, character.only = TRUE))


# ── Microclimate variable names ───────────────────────────────────────────────
# Written by data_preparation.qmd so that axis labels automatically reflect
# the actual column names in the dataset (e.g. "soil_temp", "VWC").
microclimate_index <- readLines("../output/microclimate_index.txt")
mc1_pred <- microclimate_index[1]
mc2_pred <- microclimate_index[2]


# ── Land-use colour palette ───────────────────────────────────────────────────
# Dark green = treed plots (A), burnt orange = cultivated plots (C).
# Applied consistently to every figure that distinguishes land-use type.
pos_colours <- c("Treed (A)" = "#3d5a3e", "Cultivated (C)" = "#c1603a")


# ── Causal pathway metadata ───────────────────────────────────────────────────
# One row per pathway.
#
# Columns:
#   path_id  — machine key used as join key in all downstream code
#   label    — human-readable string shown in legends and y-axes
#   colour   — warm tones for root-mediated paths, cool tones for microclimate
#   ltype    — encodes path length:
#                solid  = direct effect or composite (total indirect / total effect)
#                dashed = two-step indirect   (lu → X → fauna)
#                dotted = three-step indirect (lu → X → root → fauna)
path_meta <- tribble(
  ~path_id,         ~label,                                                ~colour,   ~ltype,
  "direct",         "Direct (lu\u2192fauna)",                              "#DCB879", "solid",
  "via_mc1",        paste0("lu\u2192", mc1_pred, "\u2192fauna"),           "#c2cab2", "dashed",
  "via_mc2",        paste0("lu\u2192", mc2_pred, "\u2192fauna"),           "#555d45", "dashed",
  "via_root",       "lu\u2192root\u2192fauna",                             "#955f42", "dashed",
  "via_mc1_root",   paste0("lu\u2192", mc1_pred, "\u2192root\u2192fauna"), "#dea88b", "dashed",
  "via_mc2_root",   paste0("lu\u2192", mc2_pred, "\u2192root\u2192fauna"), "#d1855c", "dashed",
  "total_indirect", "Total indirect",                                       "#999999", "solid",
  "total_effect",   "Total effect",                                         "#444444", "solid"
)

# Named vectors for ggplot scale_*_manual() calls — names = path_id
path_colours <- setNames(path_meta$colour, path_meta$path_id)
path_ltypes  <- setNames(path_meta$ltype,  path_meta$path_id)
path_labels  <- setNames(path_meta$label,  path_meta$path_id)


# ── Season background bands ───────────────────────────────────────────────────
# Alternating grey rectangles behind temporal plots mark the four meteorological
# seasons. Boundaries: Spring = Mar–May | Summer = Jun–Aug |
# Autumn = Sep–Nov | Winter = Dec–Feb (crosses the year boundary).
season_order <- c("Spring", "Summer", "Autumn", "Winter")
season_fills <- c(Spring = "#f9f9f9", Summer = "#eeeeee",
                  Autumn = "#f9f9f9", Winter = "#eeeeee")

# ── Agricultural intervention landmarks ───────────────────────────────────────────────────
interventions <- data.frame(
  date = as.Date(c(
    "2024-07-08", "2024-07-15", 
    "2024-11-22", "2024-11-25", "2025-03-14"
  )),
  event = c(
    "Wheat Harvest", "Sorghum Sown", 
    "Sorghum Mown", "Tilled", "Chickpea Sown"
  )
)

#' Build season annotation layers for a ggplot time-series
#'
#' Returns a list of ggplot layers (annotate rectangles + geom_text labels)
#' that can be appended to any date-scaled plot with `+ make_season_layers(...)`.
#'
#' @param date_range  Length-2 Date vector (c(start, end)) matching the plot x-axis.
#' @param y_label_pos Vertical position of season name labels. `Inf` = top of panel.
#' @param label_size  Font size (ggplot units) for season labels.
#' @return A list of ggplot annotation/geom layers.
make_season_layers <- function(date_range, y_label_pos = Inf, label_size = 2.2) {

  # Build one rectangle per season per calendar year in date_range.
  # Winter is special: its band starts in December of year N and ends in
  # March of year N+1, so xmax uses year + 1 when m_end < m_start.
  bands <- tibble(
    season  = names(season_fills),
    m_start = c(3L, 6L, 9L, 12L),
    m_end   = c(6L, 9L, 12L,  3L),
    fill    = unname(season_fills)
  ) %>%
    crossing(year = seq(year(date_range[1]), year(date_range[2]))) %>%
    mutate(
      xmin = as.Date(paste(year, m_start, "01", sep = "-")),
      xmax = as.Date(paste(
        if_else(m_end < m_start, year + 1L, year),
        m_end, "01", sep = "-"
      )),
      xmid = xmin + (xmax - xmin) / 2
    ) %>%
    # Keep only bands that overlap the data range, then clip to it.
    filter(xmin < date_range[2], xmax > date_range[1]) %>%
    mutate(
      xmin = pmax(xmin, date_range[1]),
      xmax = pmin(xmax, date_range[2])
    )

  # One annotate("rect") call per band row
  rect_layers <- pmap(
    bands %>% select(xmin, xmax, fill),
    \(xmin, xmax, fill) annotate(
      "rect",
      xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf,
      fill = fill, alpha = 1
    )
  )

  # A single geom_text layer for all season labels (uses bands data directly)
  label_layer <- geom_text(
    data = bands,
    aes(x = xmid, y = y_label_pos, label = season),
    inherit.aes = FALSE,
    vjust = 1.6, size = label_size,
    color = "grey60", fontface = "italic", family = "sans"
  )

  c(rect_layers, list(label_layer))
}


# ── Base ggplot theme ─────────────────────────────────────────────────────────
# A single theme function keeps all figures visually consistent.
#
# @param base_size   Base font size in pt (default 8).
# @param legend_pos  Passed to legend.position; use "none", "bottom", "right", etc.
theme_sem <- function(base_size = 8, legend_pos = "none") {
  theme_classic(base_size = base_size) %+replace%
    theme(
      panel.border      = element_rect(color = "grey20", fill = NA, linewidth = 0.3),
      panel.grid.major  = element_line(color = "grey94", linewidth = 0.25),
      panel.grid.minor  = element_blank(),
      panel.background  = element_rect(fill = "white", color = NA),
      axis.line         = element_blank(),
      axis.ticks        = element_line(color = "grey40", linewidth = 0.3),
      axis.ticks.length = unit(2, "pt"),
      axis.title        = element_text(size = base_size - 0.5, color = "grey20"),
      axis.text         = element_text(size = base_size - 1.5, color = "grey30"),
      strip.text        = element_text(size = base_size - 1, face = "bold",
                                       color = "grey15"),
      strip.background  = element_rect(fill = "grey96", color = "grey96",
                                       linewidth = 0.3),
      plot.title        = element_text(size = base_size + 0.5, face = "bold",
                                       hjust = 0, color = "grey10",
                                       margin = margin(b = 3)),
      plot.subtitle     = element_text(size = base_size - 1, color = "grey45",
                                       hjust = 0, margin = margin(b = 5)),
      plot.caption      = element_text(size = base_size - 2, color = "grey55",
                                       hjust = 0, margin = margin(t = 4)),
      legend.position   = legend_pos,
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.title      = element_text(size = base_size - 1, face = "bold",
                                       color = "grey20"),
      legend.text       = element_text(size = base_size - 1.5, color = "grey30"),
      legend.key.size   = unit(8, "pt"),
      legend.spacing.x  = unit(3, "pt"),
      plot.margin       = margin(6, 6, 6, 6)
    )
}


# ── Load pre-computed result databases ────────────────────────────────────────
# All modelling was performed in 3_SEM_modelisation.qmd.
# This script reads outputs only — no model is fitted here.

h1_adj <- read.csv("../output/H1_adjusted_values.csv",
                   na.strings = c("", "NA")) %>%
  mutate(
    position_label = factor(position_label,
                            levels = c("Treed (A)", "Cultivated (C)")),
    season         = factor(season, levels = season_order)
  )

h1_summary <- read.csv("../output/H1_model_summary.csv",
                       na.strings = c("", "NA"))

dag_table <- read.csv("../output/DAG_coefficients.csv",
                      na.strings = c("", "NA"))

dag_fit   <- read.csv("../output/DAG_model_fit.csv",
                      na.strings = c("", "NA"))

sem_results <- read.csv("../output/SEM_results_database.csv",
                        na.strings = c("", "NA")) %>%
  mutate(season = factor(season, levels = season_order))

indirect_effects <- read.csv("../output/indirect_effects.csv",
                             na.strings = c("", "NA"))

# Faunal transformation used during modelling (e.g. "log", "sqrt_plus_one"),
# stored by 3_SEM_modelisation.qmd and needed for axis labels.
transfo_fauna <- h1_summary$transformation[1]
```

---

## 1. H1 — Direct effect of land-use type on faunal activity

> **Land-use type directly structures faunal activity, with treed habitats
> sustaining lower but more temporally stable activity than cultivated zones.**

### How the values are computed (two-step approach)

Figures in this section show **null-model residuals** — faunal activity with
the spatial grouping structure removed.

**Step 1 — Remove spatial structure.**
A mixed model with only an intercept is fitted on the full dataset. Random
effects for orientation and depth absorb all variation attributable to scanner
placement:

$$y^* \sim 1 + (1 \mid \text{orientation} / \text{depth}) + \text{AR}(1)$$

The residuals are "leftovers" on the original transformed scale, directly
comparable between A and C without spatial confounding.

**Step 2 — Test the land-use effect on residuals.**
A simple linear model tests whether Step 1 residuals differ between land-use
types:

$$\hat{\varepsilon}_{\text{Step 1}} \sim \text{land\_use}$$

β = mean(residuals at C) − mean(residuals at A). A negative β means faunal
activity is lower in cultivated plots after removing orientation and depth
effects.

```{r}
#| label: fig-h1-combined
#| fig-cap: >
#|   **Direct effect of land-use type on faunal activity (H1).**
#|   (Left) Distribution of null-model residuals pooled across the monitoring
#|   period; a systematic offset from zero indicates a land-use effect after
#|   accounting for spatial structure. (Right) Temporal fluctuations of residuals
#|   with LOESS smoothers (span = 0.18) revealing seasonal trends.
#|   Green (A) = treed; Orange (C) = cultivated. Shaded bands represent
#|   meteorological seasons.
#| fig-width: 12
#| fig-height: 5

# ── Shared y-axis limits ──────────────────────────────────────────────────────
# Derived from the raw data so both panels share exactly the same scale.
# Rounded to the nearest 0.1 to produce clean tick positions.
y_lo <- floor(  min(h1_adj$adj_value, na.rm = TRUE) * 10) / 10
y_hi <- ceiling(max(h1_adj$adj_value, na.rm = TRUE) * 10) / 10
shared_ylim <- c(y_lo, y_hi)


# ── Left panel: boxplot ───────────────────────────────────────────────────────
# The right border is suppressed (panel.border replaced by individual axis.line
# calls) so the two panels visually merge at their shared edge.
p_box <- ggplot(
  h1_adj,
  aes(x = position_label, y = adj_value,
      fill = position_label, color = position_label)
) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_jitter(alpha = 0.05, size = 0.3, shape = 20) +
  geom_boxplot(width = 0.5, outliers = FALSE, linewidth = 0.8,
               fatten = 1.2, alpha = 0.5) +
  scale_color_manual(values = pos_colours) +
  scale_fill_manual(values  = pos_colours) +
  scale_x_discrete(labels = c("Treed", "Cultivated")) +
  coord_cartesian(ylim = shared_ylim) +
  labs(x = NULL, y = "Null-model residuals") +
  guides(fill = "none", color = "none") +
  theme_sem() +
  theme(
    axis.text.x       = element_text(size = 6.5, face = "bold"),
    panel.border      = element_blank(),
    axis.line.y.left  = element_line(color = "grey20", linewidth = 0.3),
    axis.line.x       = element_line(color = "grey20", linewidth = 0.3),
    plot.margin       = margin(t = 5, r = 0, b = 5, l = 5, unit = "pt")
  )


# ── Right panel: LOESS time series ───────────────────────────────────────────
# The left border and y-axis are suppressed — they are carried by the left panel.
# The left plot margin is zeroed to close the visual gap at the join.
p_temp <- ggplot(
  h1_adj %>% mutate(date = as.Date(date)),
  aes(x = date, color = position_label, fill = position_label)
) +
  make_season_layers(
    range(as.Date(h1_adj$date), na.rm = TRUE),
    y_label_pos = shared_ylim[2]
  ) +
  geom_vline(
    data = interventions,
    aes(xintercept = date),
    color = "darkred",
    linetype = "dotted",
    alpha = 0.6
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3,
             linetype = "dashed") +
  geom_point(aes(y = adj_value), size = 0.5, alpha = 0.15, shape = 16) +
  geom_smooth(aes(y = adj_value), method = "loess", formula = y ~ x,
              span = 0.18, linewidth = 0.7, se = FALSE) +
  scale_color_manual(values = pos_colours, name = "Land-use type",
                     labels = c("Treed (A)", "Cultivated (C)")) +
  scale_fill_manual(values  = pos_colours, name = "Land-use type",
                    labels  = c("Treed (A)", "Cultivated (C)")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  coord_cartesian(ylim = shared_ylim) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none", color = "none") +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(angle = 30, hjust = 1, size = 6.5),
    legend.direction   = "horizontal",
    axis.title.y       = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.border       = element_blank(),
    axis.line.y.right  = element_line(color = "grey20", linewidth = 0.3),
    axis.line.x        = element_line(color = "grey20", linewidth = 0.3),
    plot.margin        = margin(t = 5, r = 5, b = 5, l = 0, unit = "pt")
  )


# ── Combine with patchwork ────────────────────────────────────────────────────
# guides = "collect" merges the two colour legends into one block placed below.
# widths = c(1, 3): temporal panel is 3× wider than the boxplot.
h1_combined <- (p_box + p_temp) +
  plot_layout(widths = c(1, 3), guides = "collect") &
  theme(legend.position = "bottom", legend.direction = "horizontal")

h1_combined

ggsave("../output/Fig1_H1_combined.tiff", h1_combined,
       width = 20, height = 10, units = "cm", dpi = 300, compression = "lzw")
```

---

## 2. H2 — Indirect effects through contrasting causal pathways

> **Land-use effects are mediated through contrasting causal pathways —
> agricultural management drives faunal activity primarily through altered
> microclimate, whereas treed habitats sustain faunal activity more
> consistently through root resource supply.**

### How the pathway estimates are computed

The SEM estimates one β per path per rolling window. An **indirect effect** is
the product of all β values along a causal route. For example, lu → mc₁ → fauna
= β₁ × β₆: a unit change in land_use produces a β₁ change in mc₁, which then
produces a β₁ × β₆ change in fauna. Because all variables are z-scored within
each window, these products are standardised and directly comparable across
windows and pathways.

| Path                    | Expression                              |
|-------------------------|-----------------------------------------|
| Direct                  | $\beta_9$                               |
| lu → mc₁ → fauna        | $\beta_1 \times \beta_6$                |
| lu → mc₂ → fauna        | $\beta_2 \times \beta_7$                |
| lu → root → fauna       | $\beta_5 \times \beta_8$                |
| lu → mc₁ → root → fauna | $\beta_1 \times \beta_3 \times \beta_8$ |
| lu → mc₂ → root → fauna | $\beta_2 \times \beta_4 \times \beta_8$ |
| **Total indirect**      | sum of the five routes above            |
| **Total effect**        | direct + total indirect                 |

**Sign convention:** A = 0, C = 1. A **negative β-product means the effect
through that pathway is stronger in the treed system.**

### Full-dataset piecewise SEM: DAG annotation table

```{r tbl-h2-dag}
#| tbl-cap: >
#|   Path coefficients from the full-dataset piecewise SEM
#|   (all_taxa, complete time series, all depths). All variables were z-scored
#|   after the calibration-selected transformation. β = unstandardised
#|   coefficient; Std β = standardised, directly comparable to rolling-window
#|   values in Figs. 2–3. R²m = variance explained by fixed effects only;
#|   R²c = variance explained by fixed + random effects combined.
#|   Fisher's C tests whether the assumed causal structure is consistent with
#|   the data: p > 0.05 means the d-separation claims hold.

# ── Model-fit footnote string ─────────────────────────────────────────────────
# A saturated DAG has no d-separation claims to test, so Fisher's C and its
# p-value are NA — this is expected and not a problem.
fit_line <- with(dag_fit, paste0(
  "Model fit \u2014 Fisher\u2019s C = ",
  if (is.na(fishers_C)) "NA" else round(fishers_C, 2),
  " (df = ", if (is.na(df)) "NA" else df, ", p ",
  if (is.na(p_value))       "= NA (saturated model \u2014 no independence claims to test)"
  else if (p_value < 0.001) "< 0.001"
  else                       paste0("= ", round(p_value, 3)),
  ") \u2502 AIC = ",
  if (is.na(AIC)) "NA" else round(AIC, 1)
))

dag_table %>%
  # Drop variance–covariance rows (~~): they are not directional paths and
  # do not have meaningful t-values or β coefficients.
  filter(!str_detect(response, "~~")) %>%
  mutate(
    # piecewiseSEM may store numeric columns as character — coerce before formatting.
    across(c(estimate, std_estimate, se, t_value, p_value,
             r2_marginal, r2_conditional),
           ~ suppressWarnings(as.numeric(.))),
    across(c(estimate, std_estimate, se), ~ sprintf("%.3f", .)),
    t_fmt   = if_else(is.na(t_value), "\u2014", sprintf("%.2f", t_value)),
    p_fmt   = if_else(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value)),
    r2_marg = sprintf("%.3f", r2_marginal),
    r2_cond = sprintf("%.3f", r2_conditional)
  ) %>%
  select(
    "Equation"   = equation,
    "Predictor"  = predictor,
    "\u03b2"     = estimate,
    "Std \u03b2" = std_estimate,
    "SE"         = se,
    "t"          = t_fmt,
    "p"          = p_fmt,
    "sig"        = sig,
    "R\u00b2m"   = r2_marg,
    "R\u00b2c"   = r2_cond
  ) %>%
  kbl(format = "html", escape = FALSE,
      align  = c("l", "l", "r", "r", "r", "r", "r", "l", "r", "r")) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8") %>%
  # Merge repeated equation and R² cells for readability
  collapse_rows(columns = c(1, 9, 10), valign = "middle") %>%
  footnote(
    general = paste0(
      fit_line,
      "<br>Sig. codes: *** p < 0.001 &nbsp;** p < 0.01 &nbsp;* p < 0.05 &nbsp;. p < 0.10"
    ),
    general_title = "", escape = FALSE, footnote_as_chunk = TRUE
  )
```

### Results comparison: full-dataset vs rolling-window pathway estimates

```{r h2-temporal-prep}
# ══════════════════════════════════════════════════════════════════════════════
# PREPARE H2 TIME SERIES
#
# Two sources are merged:
#   direct_ts   — land_use → fauna β extracted from the rolling-window SEM
#                 results table (one row per window).
#   indirect_ts — β-products computed in 3_SEM_modelisation.qmd, pivoted from
#                 wide (one column per pathway) to long format.
#
# Both are restricted to all_taxa (community level), then joined with path_meta
# to attach display labels, colours, and linetypes.
# ══════════════════════════════════════════════════════════════════════════════

direct_ts <- sem_results %>%
  filter(taxon == "all_taxa",
         predictor == "land_use",
         response  == "fauna") %>%
  transmute(path_id = "direct", date_start, season, std_estimate)

indirect_ts <- indirect_effects %>%
  filter(taxon == "all_taxa") %>%
  pivot_longer(
    c(via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
      total_indirect, total_effect),
    names_to  = "path_id",
    values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate)) %>%
  # Derive season from date_start rather than relying on the saved column, which
  # may be absent in older runs of 3_SEM_modelisation.qmd.
  mutate(
    month  = month(date_start),
    season = factor(case_when(
      month %in% 3:5  ~ "Spring",
      month %in% 6:8  ~ "Summer",
      month %in% 9:11 ~ "Autumn",
      TRUE            ~ "Winter"
    ), levels = season_order)
  ) %>%
  select(path_id, date_start, season, std_estimate)

h2_ts <- bind_rows(direct_ts, indirect_ts) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label),
    season  = factor(season,  levels = season_order)
  ) %>%
  # Winsorise at 2.5–97.5th percentile per pathway before LOESS smoothing.
  # This reduces the leverage of outlier windows (e.g. sparse fauna in winter)
  # on the smoother without removing those windows entirely from the raw data.
  group_by(path_id) %>%
  mutate(
    lo           = quantile(std_estimate, 0.025, na.rm = TRUE),
    hi           = quantile(std_estimate, 0.975, na.rm = TRUE),
    std_estimate = pmax(lo, pmin(hi, std_estimate))
  ) %>%
  ungroup()


# ── LOESS smoother helper ─────────────────────────────────────────────────────
#' Fit a LOESS smoother to a single pathway's time series
#'
#' Designed for use inside group_modify(). Falls back to raw estimates when
#' fewer than 5 valid observations are available or when LOESS fails.
#'
#' @param df    Data frame with columns `date_start` (Date) and `std_estimate`.
#' @param span  LOESS bandwidth (default 0.30).
#' @return `df` with an added `smoothed` column.
smooth_path <- function(df, span = 0.30) {
  df <- df %>% filter(!is.na(std_estimate)) %>% arrange(date_start)
  if (nrow(df) < 5) return(mutate(df, smoothed = std_estimate))

  fit <- tryCatch(
    loess(
      std_estimate ~ as.numeric(date_start),
      data    = df,
      span    = span,
      control = loess.control(surface = "direct")  # avoids extrapolation artefacts at edges
    ),
    error = \(e) NULL
  )

  df$smoothed <- if (!is.null(fit)) predict(fit) else df$std_estimate
  df
}

h2_smoothed <- h2_ts %>%
  mutate(date_start = as.Date(date_start)) %>%
  group_by(path_id) %>%
  group_modify(~ smooth_path(.x)) %>%
  ungroup()

date_range_h2 <- range(h2_smoothed$date_start)
```

```{r h2-global-coef}
# ══════════════════════════════════════════════════════════════════════════════
# EXTRACT GLOBAL-MODEL PATHWAY ESTIMATES FROM dag_table
#
# global_coef: one row per directional path with the standardised coefficient
# (b) and its standard error (se). Variance–covariance rows (~~) are excluded.
# Both columns are coerced from character to numeric because piecewiseSEM
# stores them as strings.
# ══════════════════════════════════════════════════════════════════════════════

global_coef <- dag_table %>%
  filter(!str_detect(response, "~~")) %>%
  mutate(
    std_estimate = suppressWarnings(as.numeric(std_estimate)),
    se           = suppressWarnings(as.numeric(se))
  ) %>%
  select(response, predictor, b = std_estimate, se)


# ── Single-coefficient extractors ────────────────────────────────────────────
# gc()  — returns the standardised β for a given (response, predictor) pair.
# gse() — returns the corresponding standard error.
# Both return NA_real_ when a path is absent (e.g. dropped during model
# selection), so downstream β-products propagate NA gracefully.

gc <- function(resp, pred) {
  val <- global_coef$b[global_coef$response == resp & global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}

gse <- function(resp, pred) {
  val <- global_coef$se[global_coef$response == resp & global_coef$predictor == pred]
  if (length(val) == 0) NA_real_ else val[1]
}

# Individual path coefficients and their SEs
lu_mc1   <- gc("microclimate_1", "land_use");  se_lu_mc1   <- gse("microclimate_1", "land_use")
lu_mc2   <- gc("microclimate_2", "land_use");  se_lu_mc2   <- gse("microclimate_2", "land_use")
mc1_root <- gc("root", "microclimate_1");      se_mc1_root <- gse("root", "microclimate_1")
mc2_root <- gc("root", "microclimate_2");      se_mc2_root <- gse("root", "microclimate_2")
lu_root  <- gc("root", "land_use");            se_lu_root  <- gse("root", "land_use")
mc1_fau  <- gc("fauna", "microclimate_1");     se_mc1_fau  <- gse("fauna", "microclimate_1")
mc2_fau  <- gc("fauna", "microclimate_2");     se_mc2_fau  <- gse("fauna", "microclimate_2")
root_fau <- gc("fauna", "root");               se_root_fau <- gse("fauna", "root")
lu_fau   <- gc("fauna", "land_use");           se_lu_fau   <- gse("fauna", "land_use")


# ── Global β-products and their approximate SEs ───────────────────────────────
# b_global  — β-product for the full-dataset SEM, using the same algebra as
#             the rolling-window indirect effects in 3_SEM_modelisation.qmd.
# se_global — first-order product approximation of the SE (product of individual
#             path SEs). Note: the exact delta-method formula is
#             sqrt((b₂·SE₁)² + (b₁·SE₂)²); the approximation here is
#             conservative and suitable for display error bars only — do not
#             use for formal inference on products.
#
# NOTE: total_indirect and total_effect currently sum only the mc1 and mc2
# routes. Add via_root, via_mc1_root, via_mc2_root if those paths are non-negligible.
global_paths <- tribble(
  ~path_id,         ~b_global,                                  ~se_global,
  "direct",          lu_fau,                                     se_lu_fau,
  "via_mc1",         lu_mc1  * mc1_fau,                          se_lu_mc1  * se_mc1_fau,
  "via_mc2",         lu_mc2  * mc2_fau,                          se_lu_mc2  * se_mc2_fau,
  "via_root",        lu_root * root_fau,                         se_lu_root * se_root_fau,
  "via_mc1_root",    lu_mc1  * mc1_root * root_fau,              se_lu_mc1  * se_mc1_root * se_root_fau,
  "via_mc2_root",    lu_mc2  * mc2_root * root_fau,              se_lu_mc2  * se_mc2_root * se_root_fau,
  "total_indirect",  (lu_mc1 * mc1_fau) + (lu_mc2 * mc2_fau),   (se_lu_mc1 * se_mc1_fau) + (se_lu_mc2 * se_mc2_fau),
  "total_effect",    lu_fau  + (lu_mc1  * mc1_fau),              se_lu_fau  + (se_lu_mc1  * se_mc1_fau)
) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label)
  )


# ── Marginal (cross-window) summaries per pathway ────────────────────────────
# Collapses rolling-window estimates into one summary row per pathway.
#
# Reported statistics:
#   median_b — median across windows (robust central tendency)
#   sd_b     — SD across windows (temporal variability)
#   q025/q975 — 2.5th–97.5th percentile range (after winsorising)
#   q25/q75  — IQR (typical window-to-window range)
#
# select(-any_of(...)) guards against duplicate columns when this chunk is
# re-run interactively without restarting the R session.
h2_marginal <- h2_ts %>%
  group_by(path_id, label, colour, ltype) %>%
  summarise(
    median_b = median(std_estimate, na.rm = TRUE),
    sd_b     = sd(std_estimate,     na.rm = TRUE),
    q025     = quantile(std_estimate, 0.025, na.rm = TRUE),
    q975     = quantile(std_estimate, 0.975, na.rm = TRUE),
    q25      = quantile(std_estimate, 0.25,  na.rm = TRUE),
    q75      = quantile(std_estimate, 0.75,  na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  select(-any_of(c("b_global", "se_global"))) %>%
  left_join(select(global_paths, path_id, b_global, se_global), by = "path_id")
```

```{r h2-combined-prep}
# ── Shared y-axis limits ──────────────────────────────────────────────────────
# Derived simultaneously from raw estimates (h2_ts, used in the marginal panel)
# and smoothed values (h2_smoothed, used in the temporal panel) so that the
# zero reference and all gridlines are perfectly continuous across the join.
y_lo_h2 <- floor(  min(c(h2_ts$std_estimate, h2_smoothed$smoothed), na.rm = TRUE) * 10) / 10
y_hi_h2 <- ceiling(max(c(h2_ts$std_estimate, h2_smoothed$smoothed), na.rm = TRUE) * 10) / 10
shared_ylim_h2 <- c(y_lo_h2, y_hi_h2)


# ── Abbreviated pathway labels for the marginal panel x-axis ─────────────────
# scale_x_discrete(labels = ...) matches entries by their BREAK VALUES, which
# are the full label strings stored in levels(h2_marginal$label). Therefore the
# names of this vector must be the full label strings, not the path_id keys.
path_labels_short <- setNames(
  c(
    "Direct",
    paste0("mc1\u2192fauna"),
    paste0("mc2\u2192fauna"),
    "root\u2192fauna",
    "mc1\u2192root\n\u2192fauna",
    "mc2\u2192root\n\u2192fauna",
    "Total\nindirect",
    "Total\neffect"
  ),
  path_meta$label  # names = full label strings used as discrete breaks
)
```

```{r fig-h2-combined}
#| label: fig-h2-combined
#| fig-cap: >
#|   **Indirect effects of land-use type on faunal activity through contrasting
#|   causal pathways (H2).** (Left) Marginal distribution of standardised
#|   β-products across all 144 rolling windows per pathway. Thick bar = IQR;
#|   thin bar = 95% range (after winsorising); open circle = rolling-window
#|   median. Open diamond (nudged 0.28 units right) = global piecewise SEM
#|   estimate; horizontal bar = ±1 SE (first-order product approximation). A
#|   global estimate outside the rolling-window IQR indicates that the
#|   full-series fit is pulled by seasonal extremes away from the typical
#|   relationship. (Right) LOESS-smoothed (span = 0.30, surface = "direct")
#|   time series of standardised β-products. Solid line = direct or composite;
#|   dashed = two-step indirect (lu→X→fauna); dotted = three-step indirect
#|   (lu→X→root→fauna). Shaded bands = meteorological seasons. Sign convention:
#|   A = 0, C = 1 — negative values indicate a stronger effect in the treed
#|   system. Both panels share the same y-axis.
#| fig-width: 16
#| fig-height: 6

# ── Shape scale shared across both panels ─────────────────────────────────────
shape_values <- c(median = 21, global = 23)  # 21 = circle, 23 = diamond

# Phantom data for the shape legend: coordinates are NA so nothing is drawn,
# but ggplot still emits one legend entry per shape. Without this layer, shapes
# mapped inside individual aes() calls do not appear in the legend automatically.
legend_dummy <- tibble(
  x     = c(NA_character_, NA_character_),
  y     = c(NA_real_,      NA_real_),
  shape = c("median",      "global")
)


# ── Left panel: marginal summary (pathway on x, β on y) ──────────────────────
# The global-model diamond is nudged 0.28 units right of the rolling median
# so the two symbols are visually separable within each pathway column.
# The right border is suppressed so the two panels merge at their shared edge.
p_marginal <- ggplot(h2_marginal, aes(colour = path_id)) +

  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.35) +

  # Thin bar: 95% range — captures rare seasonal extremes
  geom_linerange(aes(x = label, ymin = q025, ymax = q975),
                 linewidth = 0.5, alpha = 0.30) +
  # Thick bar: IQR — typical window-to-window spread
  geom_linerange(aes(x = label, ymin = q25, ymax = q75),
                 linewidth = 2.2, alpha = 0.60) +
  # Open circle: rolling-window median
  geom_point(aes(x = label, y = median_b, shape = "median"),
             size = 2.4, fill = "white", stroke = 1.1) +

  # SE bar for global model (nudged right to separate from rolling median)
  geom_linerange(aes(x    = label,
                     ymin = b_global - se_global,
                     ymax = b_global + se_global),
                 position  = position_nudge(x = 0.28),
                 linewidth = 0.5, alpha = 0.35, na.rm = TRUE) +
  # Open diamond: global-model β-product (nudged right)
  geom_point(aes(x = label, y = b_global, shape = "global"),
             position = position_nudge(x = 0.28),
             size = 2.2, fill = "white", stroke = 0.8, na.rm = TRUE) +

  # Phantom layer — generates shape legend entries without drawing visible points
  geom_point(data    = legend_dummy,
             mapping = aes(x = x, y = y, shape = shape),
             colour = "grey30", fill = "white",
             size = 2.2, stroke = 1.0, na.rm = TRUE) +

  scale_shape_manual(
    name   = NULL,
    values = shape_values,
    labels = c(median = "Rolling median (IQR / 95%)",
               global = "Global model (\u00b1 1 SE)"),
    guide  = guide_legend(
      override.aes = list(fill = c("white", "white"), colour = "grey30",
                          size = c(2.4, 2.2))
    )
  ) +
  scale_colour_manual(values = path_colours, guide = "none") +
  scale_x_discrete(limits = levels(h2_marginal$label),
                   labels  = path_labels_short) +
  coord_cartesian(ylim = shared_ylim_h2) +
  labs(x = NULL, y = "Standardised \u03b2") +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(size = 5.5, angle = 40, hjust = 1,
                                      lineheight = 0.85),
    axis.text.y        = element_text(size = 6.5),
    legend.text        = element_text(size = 6.5),
    legend.key.size    = unit(10, "pt"),
    panel.border       = element_blank(),
    axis.line.y.left   = element_line(colour = "grey20", linewidth = 0.3),
    axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.3),
    plot.margin        = margin(t = 5, r = 0, b = 5, l = 5, unit = "pt")
  )


# ── Right panel: LOESS temporal time series ───────────────────────────────────
# y-axis is removed — carried entirely by the left marginal panel.
# Left border and left plot margin are zeroed to close the join flush.
p_time <- ggplot(
  h2_smoothed %>% mutate(date_start = as.Date(date_start)),
  aes(x        = date_start,
      y        = smoothed,
      colour   = path_id,
      linetype = path_id)
) +
  make_season_layers(date_range_h2,
                     y_label_pos = shared_ylim_h2[2] * 0.9) +
  geom_vline(
    data = interventions,
    aes(xintercept = date),
    color = "darkred",
    linetype = "dotted",
    alpha = 0.6
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.35) +
  geom_line(linewidth = 0.65) +
  scale_colour_manual(values = path_colours, labels = path_labels,
                      name   = "Pathway") +
  scale_linetype_manual(values = path_ltypes, labels = path_labels,
                        name   = "Pathway") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  coord_cartesian(ylim = shared_ylim_h2) +
  labs(x = NULL, y = NULL) +
  theme_sem(legend_pos = "bottom") +
  theme(
    axis.text.x        = element_text(angle = 30, hjust = 1, size = 6.5),
    legend.text        = element_text(size = 6.0),
    legend.title       = element_text(size = 6.5, face = "bold"),
    legend.key.width   = unit(16, "pt"),
    legend.key.height  = unit(10, "pt"),
    axis.title.y       = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.border       = element_blank(),
    axis.line.y.right  = element_line(colour = "grey20", linewidth = 0.3),
    axis.line.x.bottom = element_line(colour = "grey20", linewidth = 0.3),
    plot.margin        = margin(t = 5, r = 5, b = 5, l = 0, unit = "pt")
  )


# ── Combine with patchwork ────────────────────────────────────────────────────
# guides = "keep" preserves each panel's own legend (shape legend below marginal,
# pathway legend below temporal). "collect" would merge them and break the
# spatial association with each panel.
# widths = c(1, 3): temporal panel is 3× wider than the marginal summary.
h2_combined <- (p_marginal + p_time) +
  plot_layout(widths = c(1, 3), guides = "keep")

h2_combined

ggsave("../output/Fig2_H2_combined.tiff", h2_combined,
       width = 28, height = 16, units = "cm", dpi = 300, compression = "lzw")
```

---

## 3. H3 — Taxonomic specificity of causal pathway responses

> **Taxonomic aggregation at the community level buffers individual taxon
> specificities, but substantial between-taxon variability exists — particularly
> for the direct pathway and during phenologically active periods.**

### How the values are computed

The same rolling-window piecewise SEM is fitted separately for each taxon. For
each taxon × window, the eight pathway products are computed with identical
algebra to `all_taxa`. Within-taxon distributions are summarised as median and
IQR across all retained windows. Between-taxon variability is quantified at each
date as the SD across taxa, providing a time-varying measure of taxonomic
divergence. `all_taxa` (community-level pooled count) serves as the reference
and is excluded from between-taxon variability calculations.

```{r h3-prep}
# ══════════════════════════════════════════════════════════════════════════════
# PREPARE H3 PER-TAXON DATA
#
# Mirrors the H2 preparation but computed for all taxa rather than only
# all_taxa. The community row is retained so it can serve as a reference
# in figures.
#
# Steps:
#   1. Pivot indirect_effects to long format, one row per (taxon × pathway × window)
#   2. Bind with the direct path from sem_results
#   3. Winsorise per (taxon × pathway) independently — each taxon has its own range
#   4. Compute marginal summaries (median, IQR, 95% range) per (taxon × pathway)
# ══════════════════════════════════════════════════════════════════════════════

direct_all <- sem_results %>%
  filter(predictor == "land_use", response == "fauna") %>%
  transmute(taxon, path_id = "direct",
            date_start = as.Date(date_start), std_estimate)

indirect_all <- indirect_effects %>%
  pivot_longer(
    c(via_mc1, via_mc2, via_root, via_mc1_root, via_mc2_root,
      total_indirect, total_effect),
    names_to  = "path_id",
    values_to = "std_estimate"
  ) %>%
  filter(!is.na(std_estimate)) %>%
  mutate(date_start = as.Date(date_start))

h3_ts <- bind_rows(direct_all, indirect_all) %>%
  left_join(path_meta, by = "path_id") %>%
  mutate(
    path_id = factor(path_id, levels = path_meta$path_id),
    label   = factor(label,   levels = path_meta$label)
  ) %>%
  # Winsorise per (taxon × pathway): each taxon has its own natural range,
  # so a global clip across all taxa would over-shrink rare or extreme taxa.
  group_by(taxon, path_id) %>%
  mutate(
    lo           = quantile(std_estimate, 0.025, na.rm = TRUE),
    hi           = quantile(std_estimate, 0.975, na.rm = TRUE),
    std_estimate = pmax(lo, pmin(hi, std_estimate))
  ) %>%
  ungroup()


# ── Marginal summaries per (taxon × pathway) ──────────────────────────────────
# sd_within captures temporal variability within a taxon (how much that taxon
# fluctuates across the study period). Used in the variance decomposition below.
h3_marginal <- h3_ts %>%
  group_by(taxon, path_id, label, colour, ltype) %>%
  summarise(
    median_b  = median(std_estimate, na.rm = TRUE),
    sd_within = sd(std_estimate,     na.rm = TRUE),
    q025      = quantile(std_estimate, 0.025, na.rm = TRUE),
    q975      = quantile(std_estimate, 0.975, na.rm = TRUE),
    q25       = quantile(std_estimate, 0.25,  na.rm = TRUE),
    q75       = quantile(std_estimate, 0.75,  na.rm = TRUE),
    n_windows = sum(!is.na(std_estimate)),
    .groups   = "drop"
  ) %>%
  # Flag whether each taxon's direction agrees with the community estimate for
  # this pathway. Used for sign-consistency colouring in the heatmap.
  group_by(path_id) %>%
  mutate(
    ref_sign     = sign(median_b[taxon == "all_taxa"]),
    sign_agree   = sign(median_b) == ref_sign,
    is_community = taxon == "all_taxa"
  ) %>%
  ungroup()


# ── Between-taxon variance decomposition ─────────────────────────────────────
# Per pathway, compare:
#   sd_between     — SD of per-taxon medians (how different taxa are from each other)
#   mean_sd_within — mean temporal SD within a taxon (how variable each taxon is over time)
#   var_ratio      — sd_between / mean_sd_within
#
# A ratio > 1 means taxa differ from each other more than any individual taxon
# fluctuates over time, supporting the H3 claim that aggregation masks
# taxon-specific responses.
h3_variance <- h3_marginal %>%
  filter(taxon != "all_taxa") %>%
  group_by(path_id, label, colour) %>%
  summarise(
    sd_between     = sd(median_b,   na.rm = TRUE),
    mean_sd_within = mean(sd_within, na.rm = TRUE),
    var_ratio      = sd_between / mean_sd_within,
    n_taxa         = n(),
    .groups        = "drop"
  )


# ── LOESS smoothers per taxon ─────────────────────────────────────────────────
# Restricted to direct and total_effect pathways, which have the largest
# amplitude and show the most interpretable taxon-level variation.
# Other pathways are too flat for meaningful between-taxon comparison.
h3_smoothed <- h3_ts %>%
  filter(path_id %in% c("direct", "total_effect")) %>%
  group_by(taxon, path_id) %>%
  group_modify(~ smooth_path(.x)) %>%  # reuses smooth_path() from h2-temporal-prep
  ungroup() %>%
  mutate(
    is_community = taxon == "all_taxa",
    date_start   = as.Date(date_start)
  )
```

```{r fig-h3-diversity}
#| label: fig-h3-diversity
#| fig-cap: >
#|   **Response diversity across taxa (H3).** (A) Per-taxon rolling-window
#|   median standardised β for each causal pathway. Each point is one taxon's
#|   median across all valid windows; horizontal bars span the IQR. Community
#|   estimate (all_taxa, filled circle) is highlighted in black. Colour indicates
#|   sign agreement with the community estimate. Taxa ordered by median
#|   direct-effect estimate. (B) Variance ratio (between-taxon SD of medians /
#|   mean within-taxon temporal SD) per pathway. Ratios > 1 (dashed line) mean
#|   taxa differ more from each other than any individual taxon fluctuates over
#|   time, supporting H3.
#| fig-width: 16
#| fig-height: 10

# ── Consistent taxon order ────────────────────────────────────────────────────
# Sorted by median direct-effect β (ascending), excluding all_taxa so it does
# not influence the order. all_taxa is appended last so it always appears at
# the bottom of dot plots and at the top of heatmaps.
taxon_order <- h3_marginal %>%
  filter(path_id == "direct", taxon != "all_taxa") %>%
  arrange(median_b) %>%
  pull(taxon) %>%
  c("all_taxa")

h3_marginal <- h3_marginal %>%
  mutate(taxon = factor(taxon, levels = taxon_order))


# ── Panel A: dot-and-range plot, faceted by pathway ──────────────────────────
p_diversity <- ggplot(
  h3_marginal,
  aes(y      = taxon,
      x      = median_b,
      colour = path_id,
      alpha  = if_else(is_community, 1, 0.75))
) +
  facet_wrap(~ label, nrow = 2, scales = "free_x") +
  geom_vline(xintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.3) +
  geom_linerange(aes(xmin = q25, xmax = q75),
                 linewidth = 1.2, alpha = 0.40) +
  # Community: filled circle (shape 19); individual taxa: open circle (shape 21)
  geom_point(aes(shape = is_community,
                 size  = if_else(is_community, 3.0, 1.8)),
             fill = "white", stroke = 0.9) +
  scale_shape_manual(values = c("FALSE" = 21, "TRUE" = 19), guide = "none") +
  scale_size_identity() +
  scale_alpha_identity() +
  scale_colour_manual(values = path_colours, guide = "none") +
  labs(x = "Rolling-window median standardised \u03b2", y = NULL) +
  theme_sem() +
  theme(
    strip.text    = element_text(size = 6, face = "bold"),
    axis.text.y   = element_text(size = 5.5),
    axis.text.x   = element_text(size = 6),
    panel.spacing = unit(4, "pt")
  )
p_diversity

# ── Panel B: temporal fan plot (direct + total effect only) ──────────────────
# Thin, semi-transparent lines = individual taxa.
# Thick, opaque line = all_taxa community estimate.
p_fan <- ggplot(
  h3_smoothed,
  aes(x         = date_start,
      y         = smoothed,
      group     = interaction(taxon, path_id),
      colour    = path_id,
      linewidth = is_community,
      alpha     = is_community)
) +
  facet_wrap(~ label, ncol = 1) +
  make_season_layers(
    date_range_h2,
    y_label_pos = max(h3_smoothed$smoothed, na.rm = TRUE) * 0.95
  ) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey55", linewidth = 0.3) +
  geom_line() +
  scale_linewidth_manual(values = c("FALSE" = 0.3, "TRUE" = 1.2), guide = "none") +
  scale_alpha_manual(values    = c("FALSE" = 0.35, "TRUE" = 1.0), guide = "none") +
  scale_colour_manual(values = path_colours, guide = "none") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months",
               expand = expansion(mult = 0.01)) +
  labs(x = NULL, y = "Standardised \u03b2 (LOESS-smoothed)") +
  theme_sem() +
  theme(
    axis.text.x   = element_text(angle = 30, hjust = 1, size = 6.5),
    strip.text    = element_text(size = 6.5, face = "bold"),
    panel.spacing = unit(4, "pt")
  )
p_fan
```

```{r tbl-h3-variance}
#| tbl-cap: >
#|   Between-taxon variance decomposition per causal pathway. sd_between =
#|   SD of per-taxon rolling-window medians; mean_sd_within = mean temporal SD
#|   within a taxon across windows; var_ratio = sd_between / mean_sd_within.
#|   Highlighted rows (ratio > 1) indicate taxa differ more from each other
#|   than any individual taxon fluctuates over time. n_taxa = number of taxa
#|   with ≥ 10 valid windows for that pathway.

h3_variance %>%
  left_join(select(path_meta, path_id), by = "path_id") %>%
  arrange(desc(var_ratio)) %>%
  mutate(across(c(sd_between, mean_sd_within, var_ratio), ~ round(., 3))) %>%
  select(
    "Pathway"         = label,
    "SD between taxa" = sd_between,
    "Mean SD within"  = mean_sd_within,
    "Variance ratio"  = var_ratio,
    "N taxa"          = n_taxa
  ) %>%
  kbl(format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("condensed", "hover"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8") %>%
  row_spec(
    which(pull(arrange(h3_variance, desc(var_ratio)), var_ratio) > 1),
    background = "#f0f7f0"
  )
```
