---
title: "Check models performance"
format: html
editor: visual
---

In order to apply SEM analysis, we must make severals choices. The attached script present the analysis and associated results which help us to take the necessary decisions.

## Load package and data

```{r}
invisible(Sys.setlocale("LC_TIME", "C"))
invisible(lapply(c("tidyverse", "lubridate", "ade4", "piecewiseSEM", "performance", "lme4", "slider", "purrr", "MuMIn", "nlme", "moments"),require,character.only = TRUE))

# Fauna data
data_sem_raw <- read.csv("../output/SEM_database.csv", sep = ",", na.strings = c("", "NA"))%>%
  mutate("hum" = hum_soil,
         "temp" = temp_soil,
         "root" = root_cm2_count,
         "fauna" = all_taxa,
         position=as.factor(position))
data_sem_raw$date <- ymd_hms(data_sem_raw$date, tz = "UTC")

# taxa list 
fauna_vars <- readLines("../output/fauna_vars.txt")

# microclimate index 
microclimate_index <- readLines("../output/microclimate_index.txt")

#set.seed(123)
```

Data transformation

```{r}
# to transfrom (sqrt) only positive values of fauna variables: data_sem_raw <- data_sem_raw %>% mutate(across(all_of(fauna_vars), ~ ifelse(. >= 0, sqrt(.), .)),  across(where(is.numeric), ~ as.numeric(scale(.))))

# to transfrom (log) only positive values of fauna variables: 
data_sem_raw <- data_sem_raw %>% mutate(across(all_of(c(fauna_vars, "root")), ~ ifelse(!is.na(.), log(.+0.1), .)),  across(where(is.numeric), ~ as.numeric(scale(.))))
    
# to scale all variables : data_sem_raw <- data_sem_raw %>% mutate(across(where(is.numeric), ~ as.numeric(scale(.))))

```

```{r}
distrib <- data_sem_raw %>%
  pivot_longer(
    cols = all_of(c("root", fauna_vars, microclimate_index)),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~variable, scales = "free") +
  theme_bw()
distrib
```

## Check SEM

### Parameters configuration

```{r}
# window_days: length of each rolling time window (days)
# step_days: temporal shift between consecutive windows (days)
# missing_data_rate: minimum proportion of complete observations
#                     required for a window to be retained
# n_windows_per_position: number of rolling windows randomly
#                          sampled per spatial position
window_days <- 20
step_days <- 1
missing_data_rate <- 0.5
n_windows_per_position <- 10
```

### Rolling time windows

```{r}
# This function generates overlapping time windows of fixed
# duration, shifted by a constant time step.
create_rolling_windows <- function(df, window_days, step_days) {

  start_dates <- seq(
    min(df$date),
    max(df$date) - days(window_days) + hours(6),
    by = paste(step_days, "days")
  )

  windows <- map(
    start_dates,
    ~ df %>% filter(date >= .x & date < .x + days(window_days))
  )

  names(windows) <- format(start_dates, "%Y-%m-%d")
  windows
}
```

### SEM definition

```{r}
# This function defines the SEM structure applied to each
# rolling window. The structure is fixed across all taxa,
# positions, and time windows.
#
# Components:
# - root biomass explained by microclimate
# - fauna explained by microclimate and root
# - random effects nested by orientation / depth
# - AR(1) temporal autocorrelation
fit_sem <- function(data) {

  psem(
    lme(
      root ~ microclimate_1 + microclimate_2,
      data = data,
      random = ~ 1 | orientation / depth,
      correlation = corAR1(form = ~ 1 | orientation / depth)
    ),

    lme(
      fauna ~ microclimate_1 + microclimate_2 + root,
      data = data,
      random = ~ 1 | orientation / depth,
      correlation = corAR1(form = ~ 1 | orientation / depth)
    ),

    microclimate_1 %~~% microclimate_2
  )
}
```

### Function to run SEM by taxa

```{r}
# For a given taxon, this function:
# 1) prepares SEM-ready data,
# 2) creates rolling time windows per position,
# 3) filters windows based on data completeness,
# 4) randomly SAMPLES windows,
# 5) fits SEMs for each selected window,
# 6) extracts model coefficients.
run_sem_for_taxon <- function(taxon_var,
                              data_raw,
                              microclimate_indices,
                              window_days,
                              step_days,
                              missing_data_rate,
                              n_windows_per_position) {
  
  message("Running SEM for taxon: ", taxon_var)
  
  # Prepare SEM data
  sem_data <- data_raw %>%
    mutate(
      fauna = !!sym(taxon_var),
      microclimate_1 = !!sym(microclimate_indices[1]),
      microclimate_2 = !!sym(microclimate_indices[2])
    ) %>%
    select(date, position, orientation, depth,
           fauna, root,
           microclimate_1, microclimate_2) %>%
    mutate(depth = as.factor(depth))
  
  # Get unique positions
  positions <- unique(sem_data$position)
  selected_windows <- list()
  
  # Loop through each position
  for (pos in positions) {
    df_pos <- sem_data %>% filter(position == pos)
    
    # Create all possible rolling windows
    all_windows <- create_rolling_windows(df_pos, window_days, step_days)
    
    # Filter windows with enough non-missing data
    valid_windows <- all_windows[map_lgl(all_windows, function(win) {
      data_non_missing <- win %>% drop_na()
      nrow(data_non_missing) / (window_days * 16) >= missing_data_rate
    })]
    
    # Randomly sample windows if enough valid ones exist
    if (length(valid_windows) >= n_windows_per_position) {
      selected_indices <- sample(seq_along(valid_windows), size = n_windows_per_position)
      final_windows <- valid_windows[selected_indices]
    } else {
      warning(paste("Position", pos, "has only", length(valid_windows), "valid windows."))
      final_windows <- valid_windows
    }
    
    # Store selected windows in a structured list
    for (i in seq_along(final_windows)) {
      window_name <- names(final_windows)[i]
      selected_windows[[length(selected_windows) + 1]] <- list(
        position = pos,
        window_name = window_name,
        data = final_windows[[i]] %>% drop_na()
      )
    }
  }
  
  # Convert list of windows to tibble
  windows_tbl <- map_dfr(selected_windows, ~ tibble(
    position = .x$position,
    window_name = .x$window_name,
    data = list(.x$data)
  ))
  
  # Define SEM structure
  sem_model <- function(data) {
    psem(
      lme(root ~ microclimate_1 + microclimate_2,
          data = data,
          random = ~1 | orientation/depth,
          correlation = corAR1(form = ~1 | orientation/depth)),
      
      lme(fauna ~ microclimate_1 + microclimate_2 + root,
          data = data,
          random = ~1 | orientation/depth,
          correlation = corAR1(form = ~1 | orientation/depth)),
      
      microclimate_1 %~~% microclimate_2
    )
  }
  
  # Fit SEM models and extract results
  results_list <- list()
  for (i in seq_len(nrow(windows_tbl))) {
    data_window <- windows_tbl$data[[i]]
    window_name <- windows_tbl$window_name[i]
    pos <- windows_tbl$position[i]
    
    model_fit <- tryCatch(sem_model(data_window), error = function(e) NULL)
    
    if (!is.null(model_fit)) {
      results_list[[paste(taxon_var, pos, window_name, sep = "_")]] <- list(
        model = model_fit,
        time_window_start = window_name,
        position = pos,
        taxon = taxon_var
      )
    }
  }
  
  # Extract coefficients from SEM models
  extract_results <- function(model_name, model_info) {
    sm <- summary(model_info$model)
    coef_tbl <- as.data.frame(sm$coefficients)
    
    p_values_clean <- coef_tbl$P.Value %>%
      as.character() %>%
      str_replace("<", "") %>%
      as.numeric()
    
    tibble(
      model_name = model_name,
      taxon = model_info$taxon,
      position = model_info$position,
      time_window_start = model_info$time_window_start,
      response = coef_tbl$Response,
      predictor = coef_tbl$Predictor,
      estimate = coef_tbl$Estimate,
      se = coef_tbl$Std.Error,
      df = coef_tbl$DF,
      t_value = coef_tbl$t.value,
      p_value = p_values_clean
    )
  }
  
  summary_tbl <- map2_dfr(names(results_list), results_list, extract_results)
  
  return(list(
    summary = summary_tbl,
    models = results_list
  ))
}
```

### Run SEM

```{r}
results_all <- map(
  fauna_vars, ~ run_sem_for_taxon(
    taxon_var = .x,
    data_raw = data_sem_raw,
    microclimate_indices = microclimate_index,
    window_days = window_days,
    step_days = step_days,
    missing_data_rate = missing_data_rate,
    n_windows_per_position = n_windows_per_position
  )
)
# combine coefficient tables
summary_all <- map_dfr(results_all, "summary")
# collect all fitted SEM models
models_all <- results_all %>%
  map("models") %>%
  unlist(recursive = FALSE)

```

### Model diagnostics

```{r}
# Diagnostics focus on magnitude of deviations rather
# than hypothesis testing, to remain robust with
# large sample sizes and many models.

diagnostics_all <- map_dfr(names(models_all), function(mod_name) {
  
  current_psem <- models_all[[mod_name]]$model
  lme_models <- keep(current_psem, inherits, "lme")
  
  map_dfr(lme_models, function(mod) {
    
    res <- residuals(mod, type = "normalized")
    fit <- fitted(mod)
    n_obs <- length(res)  # nombre d’observations incluses
    
    tibble(
      model_id = mod_name,
      response = all.vars(formula(mod))[1],
      
      # Strength of heteroscedasticity
      hetero_strength = cor(abs(res), fit,
                            method = "spearman",
                            use = "complete.obs"),
      
      # Deviation from normality
      skewness = skewness(res, na.rm = TRUE),
      kurtosis = kurtosis(res, na.rm = TRUE),
      
      # Classical tests (informational)
      p_normality = shapiro.test(res)$p.value,
      p_homo = cor.test(abs(res), fit,
                        method = "spearman")$p.value,
      
      # Model fit
      r2_marginal = tryCatch(r2(mod)$R2_marginal, error = function(e) NA),
      r2_cond = tryCatch(r2(mod)$R2_conditional, error = function(e) NA),
      
      # Number of observations
      n_obs = n_obs
    )
  })
})
```

### Diagnostic summary

```{r}
# Evaluate SEM model residuals in terms of:
# - heteroscedasticity (Spearman correlation of |residuals| vs fitted)
# - normality (skewness and kurtosis of residuals)
# - classical tests (Shapiro-Wilk, Spearman for info only)
# Include number of observations to contextualize diagnostics
# Define explicit thresholds for interpretation:
hetero_threshold_low      <- 0.2  # |ρ| < 0.2 → low heteroscedasticity
hetero_threshold_moderate <- 0.4  # 0.2 ≤ |ρ| < 0.4 → moderate
skewness_threshold        <- 1    # |skewness| < 1 → acceptable
kurtosis_min              <- 2    # kurtosis > 2
kurtosis_max              <- 6    # kurtosis < 6


diagnostic_summary <- diagnostics_all %>%
  mutate(

    # Categorize heteroscedasticity magnitude
    hetero_level = case_when(
      abs(hetero_strength) < hetero_threshold_low      ~ "low",
      abs(hetero_strength) < hetero_threshold_moderate ~ "moderate",
      TRUE                                             ~ "strong"
    ),
    
    # Evaluate normality based on skewness and kurtosis
    normality_ok = (abs(skewness) < skewness_threshold) &
                    (kurtosis > kurtosis_min & kurtosis < kurtosis_max),
    
    # Overall model validity
    is_ok = (abs(hetero_strength) < hetero_threshold_moderate) &
            (abs(skewness) < (skewness_threshold + 0.5))  # slightly more tolerant
  ) %>%
  # Select and order informative columns
  select(
    model_id,
    response,
    n_obs,          
    hetero_strength,
    hetero_level,
    skewness,
    kurtosis,
    normality_ok,
    p_normality,
    p_homo,
    r2_marginal,
    r2_cond,
    is_ok
  )
```
