---
title: "From pixels to patterns"
format:
  html:
    css: css/styles.css  
    embed-resources: true
    code-tools: true
    
execute: 
  warning: false
  message: false
  echo: false
---

## Packages and data

```{r}
Sys.setlocale("LC_TIME", "C") 
if (!dir.exists("../output")) {
  dir.create("../output")
}
invisible(lapply(c("tidyverse", "lubridate", "ade4", "piecewiseSEM", "performance", "patchwork", "slider", "dsem", "kableExtra", "phylopath", "e1071", "nlme"),require,character.only = TRUE))

position_palette <- c("A" = "#385438", "B" = "#609060", "C" = "#a8c090")
season_palette <- c('autumn' = '#CA7842', "summer" = "#DDA853", 'winter' = '#8ACCD5', "spring" = "#A4B465" )

# Données faune
data_sem_raw <- read.csv("../data/SEM_database.csv", sep = ",", na.strings = c("", "NA"))%>%
  rename("root" = root,
         "fauna" = all_taxa)
data_sem_raw$date <- ymd_hms(data_sem_raw$date, tz = "UTC")
```

## H1/ Effect of land use on soil conditions

### Conditions according to land use

```{r}
variable_labels  <- c(hum = "Soil \nhumidity",
        temp = "Soil \ntemperature",
        root = "Root \ndensity",
        fauna = "Invertebrate \nabundance")
position_label <- c(A = "Unmanaged \nsystem",
        C = "Managed \nsystem")

stats <- data_sem_raw %>%
  dplyr::select(date, position, depth, orientation, fauna, root, hum, temp) %>%
  pivot_longer(cols = c('fauna', 'root','hum', 'temp'), names_to = "variable_name") %>%
  mutate(variable_name = factor(variable_name,
                                levels = c("hum", "temp", "root", "fauna")))%>%
  group_by(position, variable_name) %>%
  reframe(
    mean = mean(value, na.rm = TRUE),
    variance = var(value, na.rm = TRUE),
    skewness = skewness(value, na.rm = TRUE),
    kurtosis = kurtosis(value, na.rm = TRUE)
  )

tiff("../output/variable_distribution.tiff", width = 3000, height = 1500, res = 300)
data_sem_raw %>%
  dplyr::select(date, position, depth, orientation, fauna, root, hum, temp) %>%
  pivot_longer(cols = c('hum', 'temp', 'root','fauna'), names_to = "variable_name") %>%
  mutate(variable_name = factor(variable_name,
                                levels = c("hum", "temp", "root", "fauna")))%>%
  ggplot(aes(value, fill = position)) +
  geom_vline(xintercept = 0, linewidth = 1, color = 'grey') +
  geom_histogram(binwidth = 0.1) +
  facet_grid(position ~ variable_name, scales = "free",  labeller = labeller(variable_name = variable_labels, position = position_label)) +
  scale_fill_manual(values = position_palette) +
  guides(fill = "none") +
  labs(y=NULL, x=NULL)+
  theme_minimal() +
  geom_text(
    data = stats,
    aes(x = Inf - 1, y = Inf,
      label = sprintf("moy=%.2f\nvar=%.2f\nskew=%.2f\nkurt=%.2f",
                      mean, variance, skewness, kurtosis)),
    hjust = 1.1, vjust = 1.1, size = 3, color = "black",
    inherit.aes = FALSE) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.grid.minor = element_blank(),
    strip.text.y = element_text(angle = 0, size = 12, margin = margin(t = 10, r = 0, b = 0, l = 10)),
    strip.text.x = element_text(size=12, margin = margin(t = 10, r = 0, b = 10, l = 0)),
    panel.spacing = unit(1.5, "lines"))
dev.off()
```

### Conditions according to seasons

```{r}
variable_labels  <- c(hum = "Soil humidity",
        temp = "Soil temperature",
        root = "Root density",
        fauna = "Invertebrate abundance")
season_levels <- c("winter", "spring", "summer", "autumn")

stats <- data_sem_raw %>%
  dplyr::select(date, season, depth, orientation, fauna, root, hum, temp) %>%
  pivot_longer(cols = c('fauna', 'root','hum', 'temp'), names_to = "variable_name") %>%
  mutate(variable_name = factor(variable_name,
               levels = c("hum", "temp", "root", "fauna")),
         season = factor(season, levels = season_levels)) %>%
  group_by(season, variable_name) %>%
  reframe(
    mean = mean(value, na.rm = TRUE),
    variance = var(value, na.rm = TRUE),
    skewness = skewness(value, na.rm = TRUE),
    kurtosis = kurtosis(value, na.rm = TRUE)
  )

tiff("../output/variable_distribution_across_season.tiff", width = 2800, height = 1600, res = 300)
data_sem_raw %>%
  dplyr::select(date, season, depth, orientation, fauna, root, hum, temp) %>%
  pivot_longer(cols = c('hum', 'temp', 'root','fauna'), names_to = "variable_name") %>%
  mutate(variable_name = factor(variable_name,
                   levels = c("hum", "temp", "root", "fauna")),
        season = factor(season, levels = season_levels))%>%
  ggplot(aes(value, fill = season)) +
  geom_vline(xintercept = 0, linewidth = 1, color = 'grey') +
  geom_histogram(binwidth = 0.1) +
  facet_grid(season ~ variable_name, scales = "free",  labeller = labeller(variable_name = variable_labels)) +
  scale_fill_manual(values = season_palette) +
  guides(fill = "none") +
  labs(y=NULL, x=NULL)+
  theme_minimal() +
  geom_text(
    data = stats,
    aes(x = Inf - 1, y = Inf,
      label = sprintf("moy=%.2f\nvar=%.2f\nskew=%.2f\nkurt=%.2f",
                      mean, variance, skewness, kurtosis)),
    hjust = 1.1, vjust = 1.1, size = 2.5, color = "black",
    inherit.aes = FALSE) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.grid.minor = element_blank(),
    strip.text.y = element_text(angle = 0, size = 12, margin = margin(t = 10, r = 0, b = 0, l = 10)),
    strip.text.x = element_text(size=12, margin = margin(t = 10, r = 0, b = 10, l = 0)),
    panel.spacing.y = unit(0.5, "lines"),  
  panel.spacing.x = unit(1.5, "lines"))
dev.off()
```

## H2/ Effects of land use on the relationships between biotic and abiotic drivers

#### Parameters definition

Definition of the time integration step parameters and maximum allowable missing data rate for calculating the indices.

📌 Once the analytical procedure has been established, the influence of these parameters on the results will need to be explored in order to adapt them as necessary.

```{r}
time_step_days <- 7
window_size <- time_step_days * 4
missing_data_rate <- 0.3
```

#### Calculation of indices

```{r}
# function to compute indices
vars_to_index <- c("temp", "hum", "fauna", "root")
index_compute_tibble <- function(vec) {
  n_total <- length(vec)
  n_missing <- sum(is.na(vec))
  
  if (n_missing / n_total > missing_data_rate) {
    return(tibble(
      median = NA_real_,
      amplitude = NA_real_,
      change_ratio = NA_real_,
      max = NA_real_,
      min = NA_real_
    ))
  }
  
  tibble(
    median = median(vec, na.rm = TRUE),
    amplitude = unname(quantile(vec, 0.95, na.rm = TRUE) - quantile(vec, 0.05, na.rm = TRUE)),
    change_ratio = median(diff(round(vec, 6)), na.rm = TRUE),
    max = unname(quantile(vec, 0.975, na.rm = TRUE)),
    min = unname(quantile(vec, 0.025, na.rm = TRUE))
  )
}


data_sem_indices <- data_sem_raw %>%
  arrange(position, orientation, depth, date) %>%
  group_by(position, orientation, depth) %>%
  mutate(across( all_of(vars_to_index),
      ~ slide(.x, index_compute_tibble, .before = 0, .after = window_size - 1, .complete = TRUE),.names = "{.col}_index")) %>%
  unnest(cols = c(temp_index, hum_index, fauna_index, root_index), names_sep = "_") %>%
  ungroup()


# index selection
# based on a dimensionality reduction procedure
microclim_data <- data_sem_indices %>% ungroup() %>%
  select(starts_with(c("temp", "hum")))%>%
  drop_na() 

acp_microclimate <- dudi.pca(microclim_data, scannf = FALSE, nf = 2)
s.corcircle(acp_microclimate$co)
var_coord <- acp_microclimate$co[, 1:2]
var_names <- rownames(var_coord)

microclimate_index <- combn(var_names, 2, function(p) {
  abs(cor(as.numeric(var_coord[p[1], ]), as.numeric(var_coord[p[2], ])))
}) %>%
  { pair_matrix <- combn(var_names, 2); pair_matrix[, which.min(.)] }

cat("Microclimate index selected :", microclimate_index[1], "and", microclimate_index[2])


```

Graphical representation of indices relative to initial variables

```{r}
fauna_index <- data_sem_indices %>% 
  ggplot(aes(date))+
  geom_line(aes(y= fauna), color="#AAC4F5")+
  geom_line(aes(y=fauna_index_median))+
  facet_grid(depth+orientation~position)+
  theme_minimal()

root_index <- data_sem_indices %>% 
  ggplot(aes(date))+
  geom_line(aes(y= root), color="#C1E59F")+
  geom_line(aes(y=root_index_median))+
  facet_grid(depth+orientation~position)+
  theme_minimal()

temp_index <- data_sem_indices %>% 
  ggplot(aes(date))+
  geom_line(aes(y= temp), color="#FF937E")+
  geom_line(aes(y=temp_index_median))+
  facet_grid(depth~position)+
  theme_minimal()

hum_index <- data_sem_indices %>% filter(orientation=="O")%>%
  ggplot(aes(date))+
  geom_line(aes(y= hum), color="#FCB53B")+
  geom_line(aes(y=hum_index_amplitude))+
  facet_grid(depth~position)+
  theme_minimal()

panel <- (fauna_index | root_index) / (temp_index | hum_index)
```

#### SEM model

```{r}
# database cleaning
data_sem <- data_sem_indices %>%
  mutate(time_step = floor_date(date, unit = "week"))%>%
  select(date, time_step, position, depth, orientation, fauna_index_median, root_index_median, microclimate_index[1], microclimate_index[2])%>%
  rename("root" = root_index_median, 
         "fauna" = fauna_index_median,
         "microclimate_index_1" = microclimate_index[1],
         "microclimate_index_2" = microclimate_index[2])%>%
  drop_na()

#Function for modelling SEM
sem_fit <- function(data) {
  psem(
    lme(root ~ microclimate_index_1 + microclimate_index_2, 
        data = data, 
        random = ~1 | orientation/depth),
    
    lme(fauna ~ microclimate_index_1 + microclimate_index_2 + root, 
        data = data, 
        random = ~1 | orientation/depth),
    
    microclimate_index_1 %~~% microclimate_index_2
  )
}
```

#### Verification of model fit

```{r}
model_root_A <- lme(root ~ microclimate_index_1 + microclimate_index_2, 
        data = data_sem%>% filter(position=="A"), 
        random = ~1 | orientation/depth)
model_fauna_A <- lme(fauna ~ microclimate_index_1 + microclimate_index_2 + root,
        data = data_sem%>% filter(position=="A"), 
        random = ~1 | orientation/depth)

model_root_C <- lme(root ~ microclimate_index_1 + microclimate_index_2, 
        data = data_sem%>% filter(position=="C"), 
        random = ~1 | orientation/depth)
model_fauna_C <- lme(fauna ~ microclimate_index_1 + microclimate_index_2 + root,
        data = data_sem%>% filter(position=="C"), 
        random = ~1 | orientation/depth)

check_model(model_root_A)
check_model(model_fauna_A)
check_model(model_root_C)
check_model(model_fauna_C)
```

#### SEM application

```{r}
positions <- c("A", "C")
time_steps <- as.character(unique(data_sem$time_step))
models_date <- list()

for (pos in positions) {
  for (t in time_steps) {
    
    data_subset <- data_sem %>%
      mutate(depth = as.factor(depth)) %>%
      filter(position == pos, as.character(time_step) == t)
    
    if (nrow(data_subset) < 10) {
      cat("\nposition", pos, "-", t, ": skipped (not enough data)\n")
      next
    }
    
    cat("\nposition", pos, "-", t, ":", nrow(data_subset), "observations used\n")
    
    mod <- tryCatch(
      sem_fit(data_subset),
      error = function(e) {
        cat("  → model failed: singular or non-identifiable\n")
        return(NULL)
      }
    )
    
    if (!is.null(mod)) {
      # stocker le modèle, le time_step et la position
      models_date[[paste0("model_", pos, "_", t)]] <- list(
        model = mod,
        time_step = t,   # string pour l'instant
        position = pos
      )
    }
  }
}

extract_results <- function(model_name, model_list) {
  sm <- summary(model_list$model)
  coef_table <- as.data.frame(sm$coefficients)
  
  p_clean <- coef_table$P.Value |>
    as.character() |>
    stringr::str_replace("<", "") |>
    as.numeric()
  
  tibble::tibble(
    model = model_name,
    time_step = model_list$time_step,
    position = model_list$position,
    response = coef_table$Response,
    predictor = coef_table$Predictor,
    estimate = coef_table$Estimate,
    se = coef_table$Std.Error,
    df = coef_table$DF,
    t_value = coef_table$t.value,
    p_value = p_clean
  )
}

results_df <- purrr::map2_dfr(
  names(models_date),
  models_date,
  extract_results
)

# reconvertir time_step en POSIXct
results_df <- results_df %>%
  mutate(time_step = as.POSIXct(time_step, format = "%Y-%m-%d", tz = "UTC"))

```

Graphical representation of the SEM results

```{r}
microclimate_index_name  <- c("microclimate_index_1" = microclimate_index[1],
                 "microclimate_index_2" = microclimate_index[2])

g1 <- results_df %>%
  filter(!str_detect(predictor, "~~") & response != "") %>%
  mutate(se = as.numeric(se)) %>%
  ggplot(aes(time_step, estimate, group = position, color = position, fill = position)) + 
  geom_hline(yintercept = 0, color= "#FF3838", linetype = "dashed") +
  geom_ribbon(aes(ymin = estimate - se, ymax = estimate + se), alpha = 0.4, color = NA) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = position_palette) +
  scale_fill_manual(values = position_palette) +  
  facet_grid(predictor ~ response,scales = "free_y", labeller = labeller(
      response = as_labeller(function(x) paste0("Response: ", x)),
      predictor = as_labeller(microclimate_index_name))) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  labs(x = "Date", y = "Estimate") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12))


g2 <-results_df %>%
  filter(!str_detect(predictor, "~~") & response != "") %>%
  ggplot(aes(time_step, estimate, group = position, color = position)) + 
  geom_hline(yintercept = 0, color= "#FF3838", linetype = "dashed") +
  geom_smooth(method = "loess", size = 1.5, se = TRUE, alpha = 0.1) +
  scale_color_manual(values = position_palette) +
  facet_grid(predictor ~ response,scales = "free_y", labeller = labeller(
    response = as_labeller(function(x) paste0("Response: ", x)),
    predictor = as_labeller(microclimate_index_name))) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 months") +
  labs(x = "Date", y = "Smoothed estimators (loess method)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12))

print(g1)
print(g2)
```

### DYNAMICAL SEM

```{r}
## model 
sem = "
  microclimate_index_1_20_O -> root_20_O, 0, x2
  microclimate_index_2_20_O -> root_20_O, 0, x3
  microclimate_index_1_20_O -> fauna_20_O, 0, x4
  microclimate_index_2_20_O -> fauna_20_O, 0, x5
  root_20_O -> fauna_20_O, 0, x6

  microclimate_index_1_20_E -> root_20_E, 0, x2
  microclimate_index_2_20_E -> root_20_E, 0, x3
  microclimate_index_1_20_E -> fauna_20_E, 0, x4
  microclimate_index_2_20_E -> fauna_20_E, 0, x5
  root_20_E -> fauna_20_E, 0, x6

  microclimate_index_1_50_O -> root_50_O, 0, x2
  microclimate_index_2_50_O -> root_50_O, 0, x3
  microclimate_index_1_50_O -> fauna_50_O, 0, x4
  microclimate_index_2_50_O -> fauna_50_O, 0, x5
  root_50_O -> fauna_50_O, 0, x6

  microclimate_index_1_50_E -> root_50_E, 0, x2
  microclimate_index_2_50_E -> root_50_E, 0, x3
  microclimate_index_1_50_E -> fauna_50_E, 0, x4
  microclimate_index_2_50_E -> fauna_50_E, 0, x5
  root_50_E -> fauna_50_E, 0, x6


  # AR1
  microclimate_index_1_20_O -> root_20_O, 1, ar2
  microclimate_index_2_20_O -> root_20_O, 1, ar3
  microclimate_index_1_20_O -> fauna_20_O, 1, ar4
  microclimate_index_2_20_O -> fauna_20_O, 1, ar5
  root_20_O -> fauna_20_O, 1, ar6

  microclimate_index_1_20_E -> root_20_E, 1, ar2
  microclimate_index_2_20_E -> root_20_E, 1, ar3
  microclimate_index_1_20_E -> fauna_20_E, 1, ar4
  microclimate_index_2_20_E -> fauna_20_E, 1, ar5
  root_20_E -> fauna_20_E, 1, ar6

  microclimate_index_1_50_O -> root_50_O, 1, ar2
  microclimate_index_2_50_O -> root_50_O, 1, ar3
  microclimate_index_1_50_O -> fauna_50_O, 1, ar4
  microclimate_index_2_50_O -> fauna_50_O, 1, ar5
  root_50_O -> fauna_50_O, 1, ar6

  microclimate_index_1_50_E -> root_50_E, 1, ar2
  microclimate_index_2_50_E -> root_50_E, 1, ar3
  microclimate_index_1_50_E -> fauna_50_E, 1, ar4
  microclimate_index_2_50_E -> fauna_50_E, 1, ar5
  root_50_E -> fauna_50_E, 1, ar6
"
```

```{r}
## values extraction function
get_part = function(x){
  vars = c("microclimate_index_1", "microclimate_index_2", "fauna", "root")
  index = sapply( vars, FUN=\(y) grep(y,rownames(x$coef))[1] )
  x$coef = x$coef[index,index]
  dimnames(x$coef) = list( vars, vars )
  return(x)
}
```

```{r}
## function to run models
run_dsem_model <- function(data, site_pos, sem, nrows_sample = NULL) {
  
  data_wide <- data %>%
    filter(position == site_pos) %>%
    select(-position) %>%
    pivot_wider(
      names_from = c(depth, orientation),
      values_from = c(microclimate_index_1, microclimate_index_2, fauna, root),
      values_fn = mean
    )
  
  Z <- if (!is.null(nrows_sample)) ts(data_wide[1:nrows_sample, -1]) else ts(data_wide[, -1])
  
  fit <- dsem(
    sem = sem,
    tsdata = Z,
    control = dsem_control(use_REML = FALSE, quiet = TRUE)
  )
  
  return(list(fit = fit, data_wide = data_wide))
}

```

```{r}
## function to generate graphs
plot_dsem_results <- function(fit, data_wide) {
  
  # Fonction utilitaire pour extraire les coefficients utiles
  get_part <- function(x) {
    vars <- c("microclimate_index_1", "microclimate_index_2", "fauna", "root")
    index <- sapply(vars, FUN = \(y) grep(y, rownames(x$coef))[1])
    x$coef <- x$coef[index, index]
    dimnames(x$coef) <- list(vars, vars)
    return(x)
  }
  
  # Summary
  fit_sum <- summary(fit) %>%
    select(path, lag, name, start, parameter, first, second, direction,
           Estimate, Std_Error, z_value, p_value)
  
  # Graphique des estimates
  p1 <- plot(get_part(as_fitted_DAG(fit)), type = "width", show.legend = FALSE)
  p1$layers[[1]]$mapping$edge_width <- 0.5
  
  # Graphique des p-values (éviter de générer des relations fictives)
  pvals <- as_fitted_DAG(fit, what = "p_value")
  pvals_sub <- get_part(pvals)
  coefs_sub <- get_part(as_fitted_DAG(fit))$coef
  pvals_sub$coef[coefs_sub != 0 & pvals_sub$coef == 0] <- 1e-16
  p2 <- plot(pvals_sub, type = "width", show.legend = FALSE, colors = c('black', 'black'))
  p2$layers[[1]]$mapping$edge_width <- 0.1
  
  # Séries temporelles en long format
  longform <- data_wide %>%
    pivot_longer(
      cols = -date,
      names_to = "variable",
      values_to = "value"
    ) %>%
    mutate(
      Site = str_remove(str_extract(variable, "_[0-9]+_[OE]$"), "^_"),
      Species = str_remove(variable, "_[0-9]+_[OE]$")
    )
  
  p3 <- ggplot(longform, aes(x = date, y = value, color = Species)) +
    facet_wrap(~Site, scales = "free_y") +
    geom_line() +
    theme_bw()
  
  return(list(summary = fit_sum, p1 = p1, p2 = p2, p3 = p3))
}

```

```{r}
res_A <- run_dsem_model(data_sem, "A", sem, nrows_sample = 846) 
res_C <- run_dsem_model(data_sem, "C", sem, nrows_sample = 846)               

plots_A <- plot_dsem_results(res_A$fit, res_A$data_wide)
plots_C <- plot_dsem_results(res_C$fit, res_C$data_wide)

# A results
kable(plots_A$summary, digits = 4, caption = "DSEM summary - unmanaged system")
kable(plots_C$summary, digits = 4, caption = "DSEM summary - managed system")

p1 <- plots_A$p1+ ggtitle("estimate values for unmanaged system")
p2 <- plots_C$p1+ ggtitle("estimate values for managed system")
p3 <- plots_A$p2+ ggtitle("p values for unmanaged system")
p4 <- plots_C$p2+ ggtitle("p values for managed system")

tiff("../output/dsem_results.tif", width = 3000, height = 3000, res = 300)
(p1+p2+p3+p4)+plot_annotation(title= "dsem all taxa n= 846")
dev.off()
```
